### 2021-07-07 aircraft_wildlife_doc.Rmd
Initial commit

---
title: "Aircraft Wildlife collisions"
author: "Alexander Flick & Vy Nguyen"
date: "21/07/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

Do a PCA for a dataset with 2 or 3 groups (The dataset should have at least 4 continuous variables). The aim is to compare the results of the PCA analyses for the groups. You may consider for example:

* the swiss bank notes (split by Status) - the dataset bank is in package gclus  
* the iris data (split by Species) - the dataset iris is available by default  
* the credit card data (split by card) - the dataset CreditCard is in AER

Document your analysis in an R Markdown file. (It should be setup in a way that I can rerun your analysis). To perform your analysis:

* Do use ggplot2 for graphs. **Display your results for the 2 or 3 groups** (e.g. in side by side panels). Do also modify the default colors used in plots.  
* Start of your analysis: Try out some of the plot types in ggplot2 (for example histogram, density and boxplot).  
* Your principal component analysis: Create scree plots of the eigenvalues and calculated the percentage of variance explained by the first 2 or 3 PCs. Provide also scatterplots (or scatterplot matrices) of the first 3 PCs. Discuss the dimensionality of your datasets (splitted into the groups) and also the differences if the PCA is based on the original (centered) data or the standardized data.  
* Submit your report in both .Rmd and .pdf format.

\newpage

### Iris data plots

```{r echo=FALSE, message=FALSE}
library(ggplot2)
library(patchwork)
library(gclus)
```

To start the analysis we can, for example, plot the Petal Width (in cm), Length (in cm) and the Sepal Length (in cm) of the flowers split by the Species. We will visualize this with a density plot, a histogram and a boxplot.

```{r echo=FALSE, fig.asp=1}
theme_set(theme_classic())
color_pal<-c("seagreen", "steelblue", "darkred")
theme.title<-theme(
    plot.title = element_text(
      hjust = 0.5, # center
      size = 12,
      color = "steelblue",
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0.5, # center
      size = 10,
      color = "gray",
      face = "italic"
    )
  )

dp<-ggplot(iris) +
   aes(x=Petal.Width, color=Species, fill=Species) +
   geom_density(alpha=0.25) + 
   theme(legend.position = "none") +
   scale_color_manual(values = color_pal) + 
   scale_fill_manual(values = color_pal)

bp<-ggplot(iris) +
  aes(x=Petal.Width, y=Species, fill=Species, color=Species) +
  stat_boxplot(geom="errorbar",width=0.2) +
  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
  geom_jitter(alpha = 0.25, width = 0.2) +
  theme(legend.position = "none") +
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal)

h<-ggplot(iris) +
  aes(x=Petal.Width, color=Species, fill=Species) +
  geom_histogram(bins=30, alpha=0.25) + 
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal)

dpl<-ggplot(iris) +
   aes(x=Petal.Length, color=Species, fill=Species) +
   geom_density(alpha=0.25) + 
   theme(legend.position = "none") +
   scale_color_manual(values = color_pal) + 
   scale_fill_manual(values = color_pal)

bpl<-ggplot(iris) +
  aes(y=Petal.Length, x=Species, fill=Species, color=Species) +
  stat_boxplot(geom="errorbar",width=0.2) +
  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
  theme(legend.position = "none") +
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal)

sp<-ggplot(iris) +
  aes(x = Petal.Length, y = Petal.Width, shape = Species, color=Species, fill=Species) +
  geom_point() +
  facet_wrap(~Species) +
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal) + 
  labs(
    title = "Petal Width and Length of 150 flowers of Iris",
    subtitle = "In relation to their species",
    x = "Petal Length in cm",
    y = "Petal Width in cm"
  ) + theme.title

dps<-ggplot(iris) +
   aes(x=Sepal.Length, color=Species, fill=Species) +
   geom_density(alpha=0.25) + 
   theme(legend.position = "none") +
   scale_color_manual(values = color_pal) + 
   scale_fill_manual(values = color_pal)

hs<-ggplot(iris) +
  aes(x=Sepal.Length, color=Species, fill=Species) +
  geom_histogram(bins=30, alpha=0.25) + 
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal)

bps<-ggplot(iris) +
  aes(x=Sepal.Length, y=Species, fill=Species, color=Species) +
  stat_boxplot(geom="errorbar",width=0.2) +
  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
  geom_jitter(alpha = 0.25, width = 0.2) +
  theme(legend.position = "none") +
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal)

sps<-ggplot(iris) +
  aes(x = Sepal.Width, y = Sepal.Length, shape = Species, color=Species, fill=Species) +
  geom_point() +
  facet_wrap(~Species) +
  scale_color_manual(values = color_pal) + 
  scale_fill_manual(values = color_pal) + 
  labs(
    title = "Sepal Width and Length of 150 flowers of Iris",
    subtitle = "In relation to their species",
    x = "Sepal Width in cm",
    y = "Sepal Length in cm"
  ) + theme.title

(dp + h) / bp / (dpl + bpl) / sp
sps/ (dps + hs) / bps
```

### PCA for Iris data

To do a Principal Component Analysis we will use *prcomp()* and plot the eigenvalues in a scree plot. We see that there's an elbow at 2, so we choose 2 dimensions. *iris.pca$x* has the same dimensions as *X*.

```{r }
X<-iris[,1:4]             ## exclude the Species column
S <- cov(X)               ## compute variance-covariance matrix 
iris.pca <- prcomp(X)     ## perform PCA
lambda <- eigen(S)$values ## calculate eigenvalues
```

```{r echo=FALSE}
## scree plot
plot(lambda/sum(lambda),                        ## or instead of lambda: iris.pca$sdev^2
    main = "Scree plot of eigenvalues in %",
    col=c("black", "red", "black", "black"))
lines(lambda/sum(lambda))                       ## or: iris.pca$sdev^2
```

### 2021-07-07 state_wildlife.R

@@ -1,0 +1,100 @@
+library(ggplot2)
+library(openintro)
+data("birds")
+
+######################################
+## Change levels
+############
+factor_phase_flt<-c("Parked", "Taxi", "Take-off run", "Climb", "En Route", "Descent", "Approach", "Landing Roll")
+birds$phase_of_flt<-factor(birds$phase_of_flt, levels=factor_phase_flt)
+
+factor_struck<-c("0", "1", "2-10", "11-100", "Over 100")
+birds$birds_struck<-factor(birds$birds_struck, levels=factor_struck)
+
+
+######################################
+## Missing values
+############
+table(birds$birds_struck)
+sum(is.na(birds$birds_struck)) # amount of na values for birds struck: 39
+sum(birds$birds_struck=="0", na.rm=T)  # 33
+
+#get rows where birds struck is missing
+df <- birds[is.na(birds$birds_struck), c("remarks", "birds_seen", "effect", "birds_struck")]
+
+no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
+sum(no_strike)  # 19
+
+df[no_strike,]
+# set birds_struck to 0 for "no strike"
+birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
+
+sum(is.na(birds$birds_struck))  # now 20
+birds[is.na(birds$birds_struck),] 
+sum(birds$birds_struck=="0", na.rm=T)  # now 52
+
+
+######################################
+## State - Wildlife relationship
+############
+
+# 20 most frequent species
+ggplot(birds[birds$species %in% names(sort(table(birds$species), decreasing = TRUE)[1:20]), ], aes(..count.., species)) + geom_bar(aes(fill = state)) + 
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+table(birds$species, birds$state)
+
+# Number of Gulls collisions by state
+gulls_collisions <- birds[birds$species == 'GULLS',]
+ggplot(gulls_collisions, aes(..count.., state)) + geom_bar(aes(fill = state)) + 
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+# Species in California collisions
+california_collisions<-birds[birds$state == 'CA',]
+ggplot(california_collisions, aes(..count.., species)) + 
+  geom_bar(aes(fill = species))
+
+# Number of collisions per state
[…]

### 2021-07-07 time_line.R

@@ -1,0 +1,105 @@
+library(ggplot2)
+library(openintro)
+data("birds")
+library(patchwork)
+
+######################################
+## Change levels
+############
+factor_phase_flt<-c("Parked", "Taxi", "Take-off run", "Climb", "En Route", "Descent", "Approach", "Landing Roll")
+birds$phase_of_flt<-factor(birds$phase_of_flt, levels=factor_phase_flt)
+
+factor_struck<-c("0", "1", "2-10", "11-100", "Over 100")
+birds$birds_struck<-factor(birds$birds_struck, levels=factor_struck)
+
+
+######################################
+## Missing values
+############
+table(birds$birds_struck)
+sum(is.na(birds$birds_struck)) # amount of na values for birds struck: 39
+sum(birds$birds_struck=="0", na.rm=T)  # 33
+
+#get rows where birds struck is missing
+df <- birds[is.na(birds$birds_struck), c("remarks", "birds_seen", "effect", "birds_struck")]
+
+no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
+sum(no_strike)  # 19
+
+df[no_strike,]
+# set birds_struck to 0 for "no strike"
+birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
+
+sum(is.na(birds$birds_struck))  # now 20
+birds[is.na(birds$birds_struck),] 
+sum(birds$birds_struck=="0", na.rm=T)  # now 52
+
+
+######################################
+## Time line
+############
+
+str(birds$date)
+birds$date<-gsub(pattern="\\s?\\d{1,2}:\\d{2}:\\d{2}", replacement="", x=birds$date)
+birds$month<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%m")
+
+# Data from 1995 (as test)
+collisions.1995<-birds[grepl("1995", birds[["date"]]),]
+date_collisions <- ggplot(collisions.1995) +
+  aes(x = as.Date(as.character(date), format = "%m/%d/%Y"), y = birds_struck) +
+  geom_point()
+date_collisions + scale_x_date(date_breaks = "30 days", date_labels = "%d %b") +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+# monthly grouping over all years:
+month_plot<-ggplot(birds[!is.na(birds$birds_struck),]) +
[…]

### 2021-07-08 aircraft_wildlife_doc.Rmd
First changes in Rmd. Added analysis of factors influencing birds_struck.

@@
 
-## Task 1
+# Data preparation
 
-Do a PCA for a dataset with 2 or 3 groups (The dataset should have at least 4 continuous variables). The aim is to compare the results of the PCA analyses for the groups. You may consider for example:
+```{r echo=FALSE, fig.asp=1}
 
-* the swiss bank notes (split by Status) - the dataset bank is in package gclus  
-* the iris data (split by Species) - the dataset iris is available by default  
-* the credit card data (split by card) - the dataset CreditCard is in AER
-
-Document your analysis in an R Markdown file. (It should be setup in a way that I can rerun your analysis). To perform your analysis:
-
-* Do use ggplot2 for graphs. **Display your results for the 2 or 3 groups** (e.g. in side by side panels). Do also modify the default colors used in plots.  
-* Start of your analysis: Try out some of the plot types in ggplot2 (for example histogram, density and boxplot).  
-* Your principal component analysis: Create scree plots of the eigenvalues and calculated the percentage of variance explained by the first 2 or 3 PCs.
-Provide also scatterplots (or scatterplot matrices) of the first 3 PCs.
-Discuss the dimensionality of your datasets (splitted into the groups) and also the differences if the PCA is based on the original (centered) data or the standardized data.
-* Submit your report in both .Rmd and .pdf format.
-
-\newpage
-
-### Iris data plots
-
-```{r echo=FALSE, message=FALSE}
+library(openintro)
+data("birds")
 library(ggplot2)
 library(patchwork)
-library(gclus)
-```
+library(ggmosaic)
 
-To start the analysis we can, for example, plot the Petal Width (in cm), Length (in cm) and the Sepal Length (in cm) of the flowers split by the Species.
-We will visualize this with a density plot, a histogram and a boxplot.
-
-```{r echo=FALSE, fig.asp=1}
 theme_set(theme_classic())
-color_pal<-c("seagreen", "steelblue", "darkred")
+color_pal<-c("seagreen", "steelblue", "darkred", "orange")
 theme.title<-theme(
@@
   )
+
+factor_phase_flt<-c("Parked", "Taxi", "Take-off run", "Climb", "En Route", "Descent", "Approach", "Landing Roll")
+birds$phase_of_flt<-factor(birds$phase_of_flt, levels=factor_phase_flt)
+
+factor_struck<-c("0", "1", "2-10", "11-100", "Over 100")
+birds$birds_struck<-factor(birds$birds_struck, levels=factor_struck)
+
+birds$num_engs <- as.factor(birds$num_engs)
+
+#\newpage
+```
+
+## Handling missing values
+
+```{r echo=FALSE, message=FALSE}
+#get rows where birds struck is missing
+df <- birds[is.na(birds$birds_struck), c("remarks", "birds_seen", "effect", "birds_struck")]
+no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
+# set birds_struck to 0 for "no strike"
+birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
+
+# Amount of na for each column
+missing <- sapply(birds, function(x) sum(is.na(x)))
+
+barplot(missing)
+```
+
+# Analysis of strikes
+
+```{r echo=FALSE, fig.asp=1}
+
+above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
+height_by_phase <- ggplot(above_ground) +
+  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
+  stat_boxplot(geom="errorbar", width=0.4) +
+  geom_boxplot()
+# Heights during Descent are on average higher than 
+# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?
+count_phase <- ggplot(above_ground) +
+  aes(x = phase_of_flt, fill=phase_of_flt) +
+  geom_bar()
+
+height_by_phase / count_phase
+## descent and en-route are about equal
+
+```
+
+## Closer look on En Rout and Descent
+```{r echo=TRUE, fig.asp=1}
+#plot height density for Descent and EnRoute
+ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
+  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
+  geom_density(adjust=1.5,alpha=.8)
+
+#Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
+t.test(height~phase_of_flt, data = birds[as.integer(birds$phase_of_flt)>4 & as.integer(birds$phase_of_flt)<7 & !is.na(birds$phase_of_flt),])
+
+```
+
+check if clouds have an influence on effect 
+hyptohesis: clouds lead to less vision therefore the pilot might not see the birds
+
+```{r echo=FALSE, fig.asp=1}
+birds.filtered<-birds[!is.na(birds$effect) & !is.na(birds$sky),]
+birds.filtered$effectYes<-as.factor(birds.filtered$effect != "None")
+
+prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
+mosaicplot(prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1))
+
+prop.table(table(birds.filtered$sky,birds.filtered$birds_struck),1)
+
+```
+
+results show, that clouds do not have an impact on effect variable, 
+the proportion of effect=true is actually a little higher for some clouds and overcast
+
+## Analysis of effect variable
+
+```{r echo=FALSE, fig.asp=1}
+ggplot(birds.filtered[!is.na(birds.filtered$birds_struck),]) +
+  aes(x = birds_struck, fill = effectYes) +
+  geom_bar(position="fill")
+
+#comparison right? because we actually want to compare here the positive impact
+ggplot(birds.filtered) +
+  aes(x = birds_struck, fill = effect) +
+  geom_bar(position="fill")
+
+#mosaic in ggplot2
+#https://xang1234.github.io/mosaicdiagrams/
+
+#mosaic plot better than barplot visualizes also absolute differences
+ggplot(birds.filtered[!is.na(birds.filtered$ac_mass),]) +
+  geom_mosaic(aes(x = product(ac_mass), fill=effectYes))
+
+# Most collisions during the day #genauer betrachten
+ggplot(birds) + 
+  aes(x=time_of_day, fill=birds_struck) + 
+  geom_bar()
+
+```
 
@@
 sps/ (dps + hs) / bps
-```
 

### 2021-07-13 aircraft_wildlife_doc.Rmd
Added birds seen analysis and restructured into analysis of birds strikes, birds seen and effect variable.

@@
 
+#birds seen: remove one wrong entry
+which(birds$birds_seen==":-10")
+birds$birds_seen[9412] <- NA
+
+factor_seen<-c("2-10", "11-100")
+birds$birds_seen <- factor(birds$birds_seen, levels=factor_seen) 
+  # drop unused level ":-10" by assigning levels again
+levels(birds$birds_seen)
+
 #\newpage
@@
 #get rows where birds struck is missing
-df <- birds[is.na(birds$birds_struck), c("remarks", "birds_seen", "effect", "birds_struck")]
+df <- birds[is.na(birds$birds_struck), 
+            c("remarks", "birds_seen", "effect", "birds_struck")]
 no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
@@
 
+#TODO: get ac mass and num_engs from aircraft model
+
 # Amount of na for each column
 missing <- sapply(birds, function(x) sum(is.na(x)))
+missing_df <- data.frame("column" = names(missing), 
+                         "count" = missing, row.names = NULL)
+names(missing_df)[2] <- "count"
 
-barplot(missing)
+ggplot(missing_df) + 
+  aes(x=reorder(column, -count), y=count) +
+  geom_bar(stat="identity") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
 ```
 
-# Analysis of strikes
+# Analysis of strikes (birds_struck)
 
@@
 
-check if clouds have an influence on effect 
-hyptohesis: clouds lead to less vision therefore the pilot might not see the birds
+```{r echo=FALSE, fig.asp=1}
+#relationship of amount of birds struck with clouds?
+birds$effectYes<- as.factor(birds$effect != "None")
+birds.filtered<-birds[!is.na(birds$effect) & !is.na(birds$sky),]
+
+prop.table(table(birds.filtered$sky,birds.filtered$birds_struck),1)
+ggplot(birds.filtered)+
+  aes(x=birds_struck, fill=sky)+
+  geom_bar(position = "fill")
+```
+For 0 birs struck we have highest percentage of no clouds, so clouds might have an affect on avoiding striking birds. Its about 20% higher compared to 1 bird struck.
+for birdsstruck by sky we can see that with increasing amount of birds the sky tends to be more more overcast, as no clouds and some clouds decreases
+
+#Clouds with birds seen
+
+there are only 2 levels in the data available for birds seen
 
 ```{r echo=FALSE, fig.asp=1}
-birds.filtered<-birds[!is.na(birds$effect) & !is.na(birds$sky),]
-birds.filtered$effectYes<-as.factor(birds.filtered$effect != "None")
+#relationship of amount of birds seen with clouds?
+table(birds$birds_seen)
 
-prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
-mosaicplot(prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1))
+prop.table(table(birds.filtered$sky,birds.filtered$birds_seen),1)
+ggplot(birds.filtered[!is.na(birds.filtered$birds_seen),])+
+  aes(x=birds_seen, fill=sky)+
+  geom_bar(position = "fill")
+```
 
-prop.table(table(birds.filtered$sky,birds.filtered$birds_struck),1)
+the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
+
+```{r echo=FALSE, fig.asp=1}
+#relationship of amount of birds seen with daytime?
+table(birds$birds_seen)
+
+prop.table(table(birds.filtered$time_of_day,birds.filtered$birds_seen),1)
+ggplot(birds.filtered[!is.na(birds.filtered$birds_seen) & !is.na(birds.filtered$time_of_day),])+
+  aes(x=birds_seen, fill=time_of_day)+
+  geom_bar(position = "fill")
+```
+## Birds struck out of birds seen
+
+```{r echo=FALSE, fig.asp=1}
+#Create new column
+#calculate only when seen is not na
+df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
+```
+
+check speed
+
+```{r echo=FALSE, fig.asp=1}
+# does speed has an influence on seen? the faster the less likely to see ?
+
+spineplot(birds$speed, as.factor(is.na(birds$birds_seen)))
 
 ```
-
-results show, that clouds do not have an impact on effect variable, 
-the proportion of effect=true is actually a little higher for some clouds and overcast
+The higher the speed, the less likely to see a bird
 
@@
 ```
+
+check if clouds have an influence on effect 
+hyptohesis: clouds lead to less vision therefore the pilot might not see the birds
+
+```{r echo=FALSE, fig.asp=1}
+prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
+ggplot(birds[!is.na(birds$effect) & !is.na(birds$sky),])+
+  aes(x=sky, fill=effectYes)+
+  geom_bar(position = "fill")
+```
+
+results show, that clouds do not have an impact on effect variable, 
+the proportion of effect=true is a little higher for some clouds and overcast
+
+check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured
+
+```{r echo=FALSE, fig.asp=1}
+#anteil der effect=true vergleichen when seen and not seen
+#...
+effectYes_seen <- ggplot(birds[!is.na(birds$effectYes),]) +
+  aes(x=is.na(birds_seen), fill=effectYes) +
+  geom_bar(position="fill")
+
+#differences in kind of effect?
+effect_seen <- ggplot(birds[!is.na(birds$effect) & birds$effect != "None" ,]) +
+  aes(x=is.na(birds_seen), fill=effect) +
+  geom_bar(position="fill")
+
+effectYes_seen + effect_seen
+
+```
+
+proportion of effect is a little higher when birds were seen
 

### 2021-07-14 state_wildlife.R
Update State and Time Analysis

@@ -73,12 +73,18 @@ birds$struck_num<-unclass(birds$birds_struck)
 
 birds.with.statenames<-merge(x = birds, y = df.state, by.x = "state", by.y = "state.abb")
 filtered.birds<-birds.with.statenames[!is.na(birds.with.statenames$birds_struck),]
+
+require(plyr)
+birds.filtered$struck_num <- mapvalues(birds.filtered$struck_num, 
+                                     from=c(1,2,3,4,5), 
+                                     to=c(0,1,5,50,100))
+
 num.collisions.per.state<-aggregate(filtered.birds$struck_num, by=list(region=filtered.birds$state.name), FUN=sum)
 dfs.merged<-merge(us.state, num.collisions.per.state, by = "region")
 
 ggplot(data=dfs.merged, aes(x=long, y=lat, fill=x, group=group)) + 
   geom_polygon(color = "white") + 
-  scale_fill_continuous(low="grey", high="red")
+  scale_fill_continuous(low="grey", high="red") +
   guides(fill=FALSE) + 
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
         axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
@@ -95,6 +101,41 @@ ggplot(data=us.state, aes(x=long, y=lat, fill=region, group=group)) +
   coord_fixed(1.3)
 
 
-# Kategorisierung Bird/Non-Bird ?  (Grep wo species bird beinhaltet & dann nach height; alles kleiner als z.B. 10 ist nicht Vogel)
-# Look at different species (what's closer to equator)
+## Kategorisierung Bird/Non-Bird (Grep wo species bird beinhaltet & dann nach height; alles kleiner als z.B. 10 ist nicht Vogel)
[…]

### 2021-07-14 time_line.R

@@ -101,5 +101,66 @@ ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) 
   geom_bar()
 
 # TODO:
-# 1. Anzahl der Kollisionen auf Map plotten Sommer/Winter für eine bestimmte Spezies
+## 1. Anzahl der Kollisionen auf Map plotten Sommer/Winter für eine bestimmte Spezies
+library(maps)
+library(mapdata)
+
+usa <- map_data('usa')
+us.state <- map_data("state")
+df.state <- as.data.frame(cbind(state.abb, state.name)); df.state
+df.state$state.name <- tolower(df.state$state.name)
+
+birds$struck_num<-unclass(birds$birds_struck)
+
+birds.with.statenames<-merge(x = birds, y = df.state, by.x = "state", by.y = "state.abb")
+filtered.has.species<-birds.with.statenames[!is.na(birds.with.statenames$species),]
+is.hawk<-grepl("HAWK", filtered.has.species$species)
+filtered.hawk <- filtered.has.species[is.hawk,]
+
+filtered.hawk.summer<-filtered.hawk[as.numeric(filtered.hawk$month) >= 4 & as.numeric(filtered.hawk$month) < 10,]
+filtered.hawk.winter<-filtered.hawk[as.numeric(filtered.hawk$month) <= 3 | as.numeric(filtered.hawk$month) >= 10,]
+
+num.collisions.per.state.summer<-aggregate(filtered.hawk.summer$struck_num, by=list(region=filtered.hawk.summer$state.name), FUN=sum)
+num.collisions.per.state.winter<-aggregate(filtered.hawk.winter$struck_num, by=list(region=filtered.hawk.winter$state.name), FUN=sum)
+
+dfs.merged.summer<-merge(us.state, num.collisions.per.state.summer, by = "region")
+dfs.merged.winter<-merge(us.state, num.collisions.per.state.winter, by = "region")
+
+plot_summer<-ggplot(data=dfs.merged.summer, aes(x=long, y=lat, fill=x, group=group)) + 
+  geom_polygon(color = "white") + 
+  scale_fill_continuous(low="grey", high="red") +
+  guides(fill=FALSE) + 
+  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
+        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
[…]

### 2021-07-14 state_wildlife.R
Fix map for State and Time analysis

@@ -81,8 +81,14 @@ birds.filtered$struck_num <- mapvalues(birds.filtered$struck_num,
 
 num.collisions.per.state<-aggregate(filtered.birds$struck_num, by=list(region=filtered.birds$state.name), FUN=sum)
 dfs.merged<-merge(us.state, num.collisions.per.state, by = "region")
+length(unique(us.state$region))  # 49
+length(num.collisions.per.state$region) # 50
 
-ggplot(data=dfs.merged, aes(x=long, y=lat, fill=x, group=group)) + 
+`%nin%` = Negate(`%in%`)
+unique(us.state$region)[unique(us.state$region) %nin% num.collisions.per.state$region]   # District Of Columbia
+unique(num.collisions.per.state$region)[unique(num.collisions.per.state$region) %nin% us.state$region]  # Alaska, Hawaii
+
+ggplot(data=dfs.merged[with(dfs.merged, order(-group, order)), ], aes(x=long, y=lat, fill=x, group=group)) + 
   geom_polygon(color = "white") + 
   scale_fill_continuous(low="grey", high="red") +
   guides(fill=FALSE) + 
@@ -101,7 +107,8 @@ ggplot(data=us.state, aes(x=long, y=lat, fill=region, group=group)) +
   coord_fixed(1.3)
 
 
-## Kategorisierung Bird/Non-Bird (Grep wo species bird beinhaltet & dann nach height; alles kleiner als z.B. 10 ist nicht Vogel)
+## Kategorisierung Bird/Non-Bird (Grep wo species bird beinhaltet & dann nach height; alles kleiner als z.B. 7 ist nicht Vogel)
+
[…]

### 2021-07-14 time_line.R

@@ -99,6 +99,8 @@ ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) 
 ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) +
   aes(y = species, group=birds_struck, fill=birds_struck) +
   geom_bar()
+# Sort
+str(birds$species)
 
 # TODO:
 ## 1. Anzahl der Kollisionen auf Map plotten Sommer/Winter für eine bestimmte Spezies
@@ -123,10 +125,12 @@ filtered.hawk.winter<-filtered.hawk[as.numeric(filtered.hawk$month) <= 3 | as.nu
 num.collisions.per.state.summer<-aggregate(filtered.hawk.summer$struck_num, by=list(region=filtered.hawk.summer$state.name), FUN=sum)
 num.collisions.per.state.winter<-aggregate(filtered.hawk.winter$struck_num, by=list(region=filtered.hawk.winter$state.name), FUN=sum)
 
-dfs.merged.summer<-merge(us.state, num.collisions.per.state.summer, by = "region")
-dfs.merged.winter<-merge(us.state, num.collisions.per.state.winter, by = "region")
+dfs.merged.summer<-merge(us.state, num.collisions.per.state.summer, by = "region", all.x = TRUE)
+dfs.merged.winter<-merge(us.state, num.collisions.per.state.winter, by = "region", all.x = TRUE)
+length(unique(dfs.merged.summer$region))
+# Proportional?
 
-plot_summer<-ggplot(data=dfs.merged.summer, aes(x=long, y=lat, fill=x, group=group)) + 
+plot_summer<-ggplot(data=dfs.merged.summer[with(dfs.merged.summer, order(-group, order)), ], aes(x=long, y=lat, fill=x, group=group)) + 
   geom_polygon(color = "white") + 
   scale_fill_continuous(low="grey", high="red") +
   guides(fill=FALSE) + 
@@ -135,9 +139,9 @@ plot_summer<-ggplot(data=dfs.merged.summer, aes(x=long, y=lat, fill=x, group=gro
   ggtitle('Hawk collisions during Summer') + 
   coord_fixed(1.3)
 
-plot_winter<-ggplot(data=dfs.merged.winter, aes(x=long, y=lat, fill=x, group=group)) + 
[…]

### 2021-07-14 aircraft_wildlife_doc.Rmd
Added Aircraft analysis and formatted rmd.

@@
 
+sum(is.na(birds$atype)) #0
+birds[,c("atype", "num_engs", "ac_mass")]
+sapply(birds[,c("atype", "num_engs", "ac_mass")], function(x) sum(is.na(x)))
+
 # Amount of na for each column
@@
 
+which(birds$height==max(birds$height, na.rm = T))
+birds[15053,] # >30000 foot (1 meter = 3,28084 foot)
+?birds
 ```
@@
 ## Closer look on En Rout and Descent
-```{r echo=TRUE, fig.asp=1}
+```{r echo=TRUE, fig.asp=0.5}
 #plot height density for Descent and EnRoute
@@
 
+```
+
+t.test
+```{r echo=TRUE, fig.asp=1}
 #Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 #relationship of amount of birds struck with clouds?
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 #relationship of amount of birds seen with clouds?
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 #relationship of amount of birds seen with daytime?
@@
 ```
-## Birds struck out of birds seen
+check speed
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
+# does speed has an influence on seen? the faster the less likely to see ?
+spineplot(birds$speed, as.factor(is.na(birds$birds_seen)))
+```
+The higher the speed, the less likely to see a bird
+
+### Birds struck out of birds seen
+
+```{r echo=FALSE, fig.asp=0.5}
 #Create new column
@@
 df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
+table(birds$birds_seen, birds$birds_struck)
+
+df$birds_struck <- factor(df$birds_struck)
+
+#mosaic plot better than barplot visualizes also absolute differences
+ggplot(df) +
+  geom_mosaic(aes(x = product( birds_struck,birds_seen ), fill=birds_struck))
 ```
 
-check speed
+```{r echo=FALSE, fig.asp=0.5}
+#Create new column
+#calculate only when seen is not na
+df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
+table(birds$birds_seen, birds$birds_struck)
 
-```{r echo=FALSE, fig.asp=1}
-# does speed has an influence on seen? the faster the less likely to see ?
+df$birds_struck <- factor(df$birds_struck)
 
-spineplot(birds$speed, as.factor(is.na(birds$birds_seen)))
+levels(df$birds_seen)
+levels(df$birds_struck)
+df$difference_seen_strike <- as.integer(df$birds_struck) < as.integer(df$birds_seen)+2
+table(df$difference_seen_strike)
 
+barplot(prop.table(table(df$difference_seen_strike,df$effectYes),2))
+
+ggplot(df) +
+  aes(x=difference_seen_strike, fill=effectYes) +
+  geom_bar()
 ```
-The higher the speed, the less likely to see a bird
 
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 ggplot(birds.filtered[!is.na(birds.filtered$birds_struck),]) +
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
 #anteil der effect=true vergleichen when seen and not seen
@@
 ```
-
 proportion of effect is a little higher when birds were seen
-
-dp<-ggplot(iris) +
-   aes(x=Petal.Width, color=Species, fill=Species) +
-   geom_density(alpha=0.25) + 
-   theme(legend.position = "none") +
-   scale_color_manual(values = color_pal) + 
-   scale_fill_manual(values = color_pal)
-
-bp<-ggplot(iris) +
-  aes(x=Petal.Width, y=Species, fill=Species, color=Species) +
-  stat_boxplot(geom="errorbar",width=0.2) +
-  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
-  geom_jitter(alpha = 0.25, width = 0.2) +
-  theme(legend.position = "none") +
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal)
-
-h<-ggplot(iris) +
-  aes(x=Petal.Width, color=Species, fill=Species) +
-  geom_histogram(bins=30, alpha=0.25) + 
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal)
-
-dpl<-ggplot(iris) +
-   aes(x=Petal.Length, color=Species, fill=Species) +
-   geom_density(alpha=0.25) + 
-   theme(legend.position = "none") +
-   scale_color_manual(values = color_pal) + 
-   scale_fill_manual(values = color_pal)
-
-bpl<-ggplot(iris) +
-  aes(y=Petal.Length, x=Species, fill=Species, color=Species) +
-  stat_boxplot(geom="errorbar",width=0.2) +
-  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
-  theme(legend.position = "none") +
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal)
-
-sp<-ggplot(iris) +
-  aes(x = Petal.Length, y = Petal.Width, shape = Species, color=Species, fill=Species) +
-  geom_point() +
-  facet_wrap(~Species) +
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal) + 
-  labs(
-    title = "Petal Width and Length of 150 flowers of Iris",
-    subtitle = "In relation to their species",
-    x = "Petal Length in cm",
-    y = "Petal Width in cm"
-  ) + theme.title
-
-dps<-ggplot(iris) +
-   aes(x=Sepal.Length, color=Species, fill=Species) +
-   geom_density(alpha=0.25) + 
-   theme(legend.position = "none") +
-   scale_color_manual(values = color_pal) + 
-   scale_fill_manual(values = color_pal)
-
-hs<-ggplot(iris) +
-  aes(x=Sepal.Length, color=Species, fill=Species) +
-  geom_histogram(bins=30, alpha=0.25) + 
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal)
-
-bps<-ggplot(iris) +
-  aes(x=Sepal.Length, y=Species, fill=Species, color=Species) +
-  stat_boxplot(geom="errorbar",width=0.2) +
-  geom_boxplot(varwidth = TRUE, alpha=0.2) + 
-  geom_jitter(alpha = 0.25, width = 0.2) +
-  theme(legend.position = "none") +
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal)
-
-sps<-ggplot(iris) +
-  aes(x = Sepal.Width, y = Sepal.Length, shape = Species, color=Species, fill=Species) +
-  geom_point() +
-  facet_wrap(~Species) +
-  scale_color_manual(values = color_pal) + 
-  scale_fill_manual(values = color_pal) + 
-  labs(
-    title = "Sepal Width and Length of 150 flowers of Iris",
-    subtitle = "In relation to their species",
-    x = "Sepal Width in cm",
-    y = "Sepal Length in cm"
-  ) + theme.title
-
-(dp + h) / bp / (dpl + bpl) / sp
-sps/ (dps + hs) / bps
-
-### PCA for Iris data
-
-To do a Principal Component Analysis we will use *prcomp()* and plot the eigenvalues in a scree plot.
-We see that there's an elbow at 2, so we choose 2 dimensions.
-*iris.pca$x* has the same dimensions as *X*.
-
-```{r }
-X<-iris[,1:4]             ## exclude the Species column
-S <- cov(X)               ## compute variance-covariance matrix 
-iris.pca <- prcomp(X)     ## perform PCA
-lambda <- eigen(S)$values ## calculate eigenvalues
-```
-
-```{r echo=FALSE}
-## scree plot
-plot(lambda/sum(lambda),                        ## or instead of lambda: iris.pca$sdev^2
-    main = "Scree plot of eigenvalues in %",
-    col=c("black", "red", "black", "black"))
-lines(lambda/sum(lambda))                       ## or: iris.pca$sdev^2
-```

### 2021-07-14 aircraft_models.R

@@ -1,0 +1,135 @@
+machines <- unique(birds[,c("atype", "num_engs", "ac_mass")])
+sapply(machines, function(x) sum(is.na(x)))
+
+unique(unique(machines)[,1]) #one machine with different uniques
+
+which(is.na(unique(birds[,c("atype", "num_engs", "ac_mass")]))) #19 machines with na
+
+machines[is.na(machines$num_engs) | is.na(machines$ac_mass),]
+
+na_machines <- machines[is.na(machines$num_engs) | is.na(machines$ac_mass),1]
+na_machines$atype
+
+birds[birds$atype %in% na_machines$atype,]
+
+
+#check unknown as type
+birds[birds$atype=="UNKNOWN",]
+birds[birds$atype=="HELICOPTER",]
+birds[birds$atype=="MILITARY",]
+birds[birds$atype=="HOMEBUILT",]
+
+birds[birds$operator=="PRIVATELY OWNED",]
+birds[birds$operator=="GOVERNMENT",]
+
+operator <- unique(birds[,"operator"])
+grep(".*CHARTER.*", operator$operator)
+operator[grep(".*CHARTER.*", operator$operator),]$operator
+birds[birds$operator %in% operator[grep(".*CHARTER.*", operator$operator),]$operator, ]
+birds[grep(".*EXPRESS.*", birds$operator),]
+plot(machines$ac_mass, machines$num_engs )
+
+#create aircraft statistics
+install.packages("dplyr")
+install.packages("openintro")
+require(dplyr)
+require(openintro)
+data("birds")
+
+df <- birds[!is.na(birds$height) & !is.na(birds$speed),]
+machines <- df %>%
+  group_by(atype) %>%
+  summarise(max_speed = max(speed),
+            max_height = max(height),
+            num_engs = median(as.integer(num_engs)),
+            ac_mass = median(as.integer(ac_mass)))
+
+machines <- as.data.frame(machines)
+
+machines$num_engs <- factor(machines$num_engs, levels=1:4)
+machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
+
+machines$ac_mass2 <- ifelse(as.integer(machines$ac_mass)<3,1,ifelse(as.integer(machines$ac_mass)==3,2,3))
+machines$ac_mass2 <- factor(machines$ac_mass2, levels=1:3)  
[…]

### 2021-07-16 aircraft_wildlife_doc.Rmd
Restructuring and combining plots.

@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, message= FALSE, fig.asp=1}
 
@@
 
+factor_sky<-c("No Cloud", "Some Cloud", "Overcast")
+birds$sky <- factor(birds$sky, levels=factor_sky)
+
+factor_effect <- c("None","Other","Aborted Take-off","Engine Shut Down","Precautionary Landing")
+birds$effect <- factor(birds$effect, levels=factor_effect)
+
+birds$effectYes<- as.factor(birds$effect != "None")
+
 birds$num_engs <- as.factor(birds$num_engs)
@@
 #birds seen: remove one wrong entry
-which(birds$birds_seen==":-10")
+  # which(birds$birds_seen==":-10")
 birds$birds_seen[9412] <- NA
 
-factor_seen<-c("2-10", "11-100")
+factor_seen<-c("None","2-10", "11-100")
 birds$birds_seen <- factor(birds$birds_seen, levels=factor_seen) 
   # drop unused level ":-10" by assigning levels again
-levels(birds$birds_seen)
-
+birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
+#set Na values in birds seen to "None"
+birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
 #\newpage
@@
 
-```{r echo=FALSE, message=FALSE}
+```{r echo=FALSE, message=FALSE, fig.asp=0.5}
 #get rows where birds struck is missing
@@
 birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
-
-#TODO: get ac mass and num_engs from aircraft model
-
-sum(is.na(birds$atype)) #0
-birds[,c("atype", "num_engs", "ac_mass")]
-sapply(birds[,c("atype", "num_engs", "ac_mass")], function(x) sum(is.na(x)))
 
@@
 
-```{r echo=FALSE, fig.asp=1}
+```{r echo=FALSE, fig.asp=0.5}
+birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck),]) +
+  aes(x = phase_of_flt, fill=birds_struck) +
+  geom_bar()
 
+#relationship of amount of birds struck with clouds?
+birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
+  aes(x=birds_struck, fill=sky)+
+  geom_bar(position = "fill")
+
+# Most collisions during the day #genauer betrachten
+birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
+  aes(x=time_of_day, fill=birds_struck) + 
+  geom_bar()
+
+### Birds struck out of birds seen
+birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
+  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
+  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+(birds_strike_phase + birds_strike_sky) / (birds_strike_time + birds_struck_seen)
+## descent and en-route are about equal 
+```
+TODO: add Chi-squared independence test for sky and birds struck
+http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r
+
+the higher the amount of birds struck the more clouds
+
+## Closer look on En Rout and Descent
+```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
 above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
@@
 # hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?
-count_phase <- ggplot(above_ground) +
-  aes(x = phase_of_flt, fill=phase_of_flt) +
-  geom_bar()
 
-height_by_phase / count_phase
-## descent and en-route are about equal
-
-which(birds$height==max(birds$height, na.rm = T))
-birds[15053,] # >30000 foot (1 meter = 3,28084 foot)
-?birds
-```
-
-## Closer look on En Rout and Descent
-```{r echo=TRUE, fig.asp=0.5}
 #plot height density for Descent and EnRoute
-ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
+density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
   aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
   geom_density(adjust=1.5,alpha=.8)
+
+height_by_phase + density
 
@@
 t.test
-```{r echo=TRUE, fig.asp=1}
+```{r echo=FALSE, fig.asp=1}
 #Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
@@
 ```
-
-```{r echo=FALSE, fig.asp=0.5}
-#relationship of amount of birds struck with clouds?
-birds$effectYes<- as.factor(birds$effect != "None")
-birds.filtered<-birds[!is.na(birds$effect) & !is.na(birds$sky),]
-
-prop.table(table(birds.filtered$sky,birds.filtered$birds_struck),1)
-ggplot(birds.filtered)+
-  aes(x=birds_struck, fill=sky)+
-  geom_bar(position = "fill")
-```
 For 0 birs struck we have highest percentage of no clouds, so clouds might have an affect on avoiding striking birds. Its about 20% higher compared to 1 bird struck.
@@
 
-#Clouds with birds seen
+## density birdsstuck by height
+```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
+#could include only birds from vys categorization
+#plot height density for Descent and EnRoute
+density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
+  aes(x = height, group=birds_struck,fill=birds_struck)+
+  geom_density(adjust=1.5,alpha=.6) +
+  xlim(0, 3500)
+density
+
+```
+
+## Analysis of birds seen
 
@@
 #relationship of amount of birds seen with clouds?
-table(birds$birds_seen)
+birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
+  aes(x=time_of_day, fill=birds_seen)+
+  geom_bar(position = "fill") +
+  theme(legend.position="none")
 
-prop.table(table(birds.filtered$sky,birds.filtered$birds_seen),1)
-ggplot(birds.filtered[!is.na(birds.filtered$birds_seen),])+
-  aes(x=birds_seen, fill=sky)+
-  geom_bar(position = "fill")
+birds_seenNo_sky <- ggplot(birds[!is.na(birds$sky),])+
+  aes(x=sky, fill=birds_seen)+
+  geom_bar(position="fill")
+
+#relationship of seen or not seen with daytime?
+#table(is.na(birds$birds_seen), birds$time_of_day)
+
+birds_seen_plt + birds_seenNo_sky
 ```
+**Time of Day:** Based on the assumption that Missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is `r sum(birds$birds_seen!="None")/nrow(birds)`.
+The contingencytable looks like the following: `addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean)`.
 
-the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
+The proportion of overall birds seen at night is much lower compared to the other times (Dawn, Day, Dusk).
+And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).
 
-```{r echo=FALSE, fig.asp=0.5}
-#relationship of amount of birds seen with daytime?
-table(birds$birds_seen)
+**Sky:** the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
+contradiction: would have expected to see less birds when amount of clouds increases. but especially the amount of birds in groups from 11-100 increases when the amount of clouds increases.
 
-prop.table(table(birds.filtered$time_of_day,birds.filtered$birds_seen),1)
-ggplot(birds.filtered[!is.na(birds.filtered$birds_seen) & !is.na(birds.filtered$time_of_day),])+
-  aes(x=birds_seen, fill=time_of_day)+
-  geom_bar(position = "fill")
+```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
+par(mfrow=c(1,2))
+spineplot(birds$height, birds$birds_seen)
+spineplot(birds$speed, birds$birds_seen)
 ```
-check speed
+The higher the speed & height, the less likely to see a bird before collision
+for height > 2000 foot probability to see a birds before collision decreases 
+for speed > 140 knots probability to see a bird before collision decreases
 
-```{r echo=FALSE, fig.asp=0.5}
-# does speed has an influence on seen? the faster the less likely to see ?
-spineplot(birds$speed, as.factor(is.na(birds$birds_seen)))
-```
-The higher the speed, the less likely to see a bird
-
-### Birds struck out of birds seen
+## effect when birdsstruck < birds seen
 
@@
 #calculate only when seen is not na
-df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
-table(birds$birds_seen, birds$birds_struck)
+seen_vs_strike <- birds[birds$birds_seen!="None" & !is.na(birds$birds_struck),]
+seen_vs_strike$birds_struck <- factor(seen_vs_strike$birds_struck)
+seen_vs_strike$birds_seen <- factor(seen_vs_strike$birds_seen)
+seen_vs_strike$less_strike_than_seen <- as.factor(as.integer(seen_vs_strike$birds_struck) < as.integer(seen_vs_strike$birds_seen)+2)
 
-df$birds_struck <- factor(df$birds_struck)
+#TODO: Why is barplot different from ggplot???
+#barplot(prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2), xlab = #"less_strike_than_seen")
+#prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2)
 
-#mosaic plot better than barplot visualizes also absolute differences
-ggplot(df) +
-  geom_mosaic(aes(x = product( birds_struck,birds_seen ), fill=birds_struck))
+# phase of flight by birdsstruck < birds seen
+seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt),]) +
+  aes(x=phase_of_flt, fill=less_strike_than_seen) +
+  geom_bar(position = "fill")
+
+#ac_mass by leess-strike-than-seen (MOSAIC)
+seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
+  geom_mosaic(aes(x=product(ac_mass), fill=less_strike_than_seen)) +
+   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+#num_engs by leess-strike-than-seen (MOSAIC)
+seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
+  geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
+   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+seen_vs_strike_phase / (seen_vs_strike_ac_mass + seen_vs_strike_num_engs)
 ```
-
-```{r echo=FALSE, fig.asp=0.5}
-#Create new column
-#calculate only when seen is not na
-df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
-table(birds$birds_seen, birds$birds_struck)
-
-df$birds_struck <- factor(df$birds_struck)
-
-levels(df$birds_seen)
-levels(df$birds_struck)
-df$difference_seen_strike <- as.integer(df$birds_struck) < as.integer(df$birds_seen)+2
-table(df$difference_seen_strike)
-
-barplot(prop.table(table(df$difference_seen_strike,df$effectYes),2))
-
-ggplot(df) +
-  aes(x=difference_seen_strike, fill=effectYes) +
-  geom_bar()
-```
+the num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
+lighter machines and machines with less sngines tend to hit less wildlife out of the seen wildlife. This might be to better maneuvarability of the smaller aircrafts.
 
@@
 ```{r echo=FALSE, fig.asp=0.5}
-ggplot(birds.filtered[!is.na(birds.filtered$birds_struck),]) +
-  aes(x = birds_struck, fill = effectYes) +
-  geom_bar(position="fill")
-
 #comparison right? because we actually want to compare here the positive impact
-ggplot(birds.filtered) +
+birds_struck_effect <- ggplot(birds[!is.na(birds$birds_struck) & !is.na(birds$effect),]) +
   aes(x = birds_struck, fill = effect) +
@@
 #mosaic plot better than barplot visualizes also absolute differences
-ggplot(birds.filtered[!is.na(birds.filtered$ac_mass),]) +
-  geom_mosaic(aes(x = product(ac_mass), fill=effectYes))
+birds_effect_acmass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$effect),]) +
+  geom_mosaic(aes(x = product(ac_mass), fill=effect)) +
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
 
-# Most collisions during the day #genauer betrachten
-ggplot(birds) + 
-  aes(x=time_of_day, fill=birds_struck) + 
-  geom_bar()
+#birdds_seenYes by effect
+birds_effect_seenYes <- ggplot(birds[!is.na(birds$effect),]) +
+  geom_mosaic(aes(x = product(birds_seenYes), fill=effect))+
+  #theme(legend.title = element_blank())
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+birds_struck_effect / (birds_effect_acmass + birds_effect_seenYes)
 
 ```
-
-check if clouds have an influence on effect 
-hyptohesis: clouds lead to less vision therefore the pilot might not see the birds
-
-```{r echo=FALSE, fig.asp=0.5}
-prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
-ggplot(birds[!is.na(birds$effect) & !is.na(birds$sky),])+
-  aes(x=sky, fill=effectYes)+
-  geom_bar(position = "fill")
-```
-
-results show, that clouds do not have an impact on effect variable, 
-the proportion of effect=true is a little higher for some clouds and overcast
-
 check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured
 
-```{r echo=FALSE, fig.asp=0.5}
-#anteil der effect=true vergleichen when seen and not seen
-#...
-effectYes_seen <- ggplot(birds[!is.na(birds$effectYes),]) +
-  aes(x=is.na(birds_seen), fill=effectYes) +
-  geom_bar(position="fill")
-
-#differences in kind of effect?
-effect_seen <- ggplot(birds[!is.na(birds$effect) & birds$effect != "None" ,]) +
-  aes(x=is.na(birds_seen), fill=effect) +
-  geom_bar(position="fill")
-
-effectYes_seen + effect_seen
-
-```
-proportion of effect is a little higher when birds were seen
+proportion of effect is a little higher when birds were seen => contradiction, we would expect to have less amount of effect when pilot is able to see birds before, cause he could

### 2021-07-16 aircraft_wildlife_doc.Rmd
Added color palette & proportional values

@@
 output:
+  pdf_document: default
   html_document: default
-  pdf_document: default
 ---
@@
 library(ggmosaic)
+library("viridis")
+library(RColorBrewer) # https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/  -> Purples, BuPu; Map: BrBG
 
-theme_set(theme_classic())
+theme_set(theme_minimal())
 color_pal<-c("seagreen", "steelblue", "darkred", "orange")
+color_pal_sky<-c("lightgoldenrod","honeydew","skyblue3")
+color_pal_TF<-c("darksalmon","cadetblue")
 theme.title<-theme(
@@
   aes(x=reorder(column, -count), y=count) +
-  geom_bar(stat="identity") +
+  geom_bar(stat="identity", fill="snow3") +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
@@
   aes(x = phase_of_flt, fill=birds_struck) +
+  scale_fill_brewer(palette="Blues") +
   geom_bar()
@@
   aes(x=birds_struck, fill=sky)+
+  scale_fill_manual(values = color_pal_sky)+
   geom_bar(position = "fill")
@@
   aes(x=time_of_day, fill=birds_struck) + 
+  scale_fill_brewer(palette="Blues") +
   geom_bar()
@@
   geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
+  scale_fill_brewer(palette="Blues") +
   theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())
@@
   stat_boxplot(geom="errorbar", width=0.4) +
+  scale_fill_brewer(palette="Pastel2") +
   geom_boxplot()
@@
   aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
+  scale_fill_brewer(palette='Pastel2') +
   geom_density(adjust=1.5,alpha=.8)
@@
   geom_density(adjust=1.5,alpha=.6) +
+  scale_fill_brewer(palette="Blues") +
   xlim(0, 3500)
@@
   geom_bar(position = "fill") +
+  scale_fill_brewer(palette="BuGn") +
   theme(legend.position="none")
@@
   aes(x=sky, fill=birds_seen)+
+  scale_fill_brewer(palette="BuGn") +
   geom_bar(position="fill")
@@
 par(mfrow=c(1,2))
-spineplot(birds$height, birds$birds_seen)
-spineplot(birds$speed, birds$birds_seen)
+spineplot(birds$height, birds$birds_seen, col = rev(brewer.pal(n=3,name="BuGn")))
+spineplot(birds$speed, birds$birds_seen, col = rev(brewer.pal(n=3,name="BuGn")))
 ```
@@
   aes(x=phase_of_flt, fill=less_strike_than_seen) +
+  scale_fill_manual(values = color_pal_TF)+
   geom_bar(position = "fill")
@@
   geom_mosaic(aes(x=product(ac_mass), fill=less_strike_than_seen)) +
+  scale_fill_manual(values = color_pal_TF)+
   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
@@
   geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
+  scale_fill_manual(values = color_pal_TF)+
   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
@@
   aes(x = birds_struck, fill = effect) +
+  scale_fill_brewer(palette="OrRd") +
   geom_bar(position="fill")
@@
   geom_mosaic(aes(x = product(ac_mass), fill=effect)) +
+  scale_fill_brewer(palette="OrRd") +
   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
@@
   geom_mosaic(aes(x = product(birds_seenYes), fill=effect))+
+  scale_fill_brewer(palette="OrRd") +
   #theme(legend.title = element_blank())

### 2021-07-16 state_wildlife.R

@@ -39,7 +39,7 @@ sum(birds$birds_struck=="0", na.rm=T)  # now 52
 ############
 
 # 20 most frequent species
-ggplot(birds[birds$species %in% names(sort(table(birds$species), decreasing = TRUE)[1:20]), ], aes(..count.., species)) + geom_bar(aes(fill = state)) + 
+ggplot(as.data.frame(sort(table(birds$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + geom_bar(stat="identity") + 
   theme(axis.text.x = element_text(angle = 60, hjust = 1))
 
 table(birds$species, birds$state)
@@ -55,9 +55,9 @@ ggplot(california_collisions, aes(..count.., species)) +
   geom_bar(aes(fill = species))
 
 # Number of collisions per state
-ggplot(data = birds[birds$birds_struck != "0" | !is.na(birds$birds_struck),], aes(..count.., state, fill = ..count..)) + 
-  geom_bar() +
-  scale_fill_continuous(low="grey", high="red")
+ggplot(data = birds[birds$birds_struck != "0" | !is.na(birds$birds_struck),], aes(..count.., state, fill = as.factor(ac_mass))) + 
+  geom_bar()
+  #scale_fill_continuous(low="grey", high="red")
 
 
 # Number of collisions per state in a map
@@ -118,6 +118,8 @@ ggplot(birds.with.statenames[!birds.with.statenames$is.bird & !is.na(birds.with.
 
 filtered.isbird<-birds.with.statenames[!is.na(birds.with.statenames$is.bird),]
 is.bird.state<-aggregate(filtered.isbird$is.bird, by=list(region=filtered.isbird$state.name), FUN=sum)
+num.coll.per.state<-as.data.frame(table(filtered.isbird$state.name))
+is.bird.state$x <- is.bird.state$x / num.coll.per.state$Freq
 dfs.merged.2<-merge(us.state, is.bird.state, by = "region")
 
 # The redder the more birds
@@ -130,6 +132,8 @@ guides(fill=FALSE) +
   ggtitle('U.S. Map with States') + 
   coord_fixed(1.3)
 
+## Birds/Non-Birds in zwei Karten nebeneinander
+# https://ggplot2.tidyverse.org/reference/geom_map.html
 
 ## Look at different species (what's closer to equator)
 # Eine Spezies raussuchen: mappen
[…]

### 2021-07-16 time_line.R

@@ -57,6 +57,11 @@ month_plot<-ggplot(birds[!is.na(birds$birds_struck),]) +
   geom_bar()
 # -> more collisions during autumn/fall
 
+ggplot(birds[!is.na(birds$sky),]) +
+  aes(x = month, group=sky, fill=sky) +
+  geom_bar(position="fill")
+# -> less overcast during summer months
+
 # Or weekly:
 birds$day<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%a")
 week_plot<-ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) +
@@ -78,8 +83,7 @@ woy_plot<-ggplot(birds) +
   aes(x = calendar_week, group=birds_struck, fill=birds_struck) +
   geom_bar()
 
-month_plot / woy_plot
-week_plot / tod_plot
+(month_plot + woy_plot) / (week_plot + tod_plot)
 
 month_plot_filt<-ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) +
   aes(x = month, group=birds_struck, fill=birds_struck) +
@@ -99,7 +103,7 @@ ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) 
 ggplot(birds[!is.na(birds$birds_struck) & as.integer(birds$birds_struck) > 3,]) +
   aes(y = species, group=birds_struck, fill=birds_struck) +
   geom_bar()
-# Sort
+# Take top 3, small graph
 str(birds$species)
 
 # TODO:
@@ -113,6 +117,10 @@ df.state <- as.data.frame(cbind(state.abb, state.name)); df.state
 df.state$state.name <- tolower(df.state$state.name)
 
 birds$struck_num<-unclass(birds$birds_struck)
+require(plyr)
+birds$struck_num <- mapvalues(birds$struck_num, 
+                                       from=c(1,2,3,4,5), 
+                                       to=c(0,1,5,50,100))
 
 birds.with.statenames<-merge(x = birds, y = df.state, by.x = "state", by.y = "state.abb")
 filtered.has.species<-birds.with.statenames[!is.na(birds.with.statenames$species),]
@@ -122,8 +130,13 @@ filtered.hawk <- filtered.has.species[is.hawk,]
 filtered.hawk.summer<-filtered.hawk[as.numeric(filtered.hawk$month) >= 4 & as.numeric(filtered.hawk$month) < 10,]
[…]

### 2021-07-18 state-time.Rmd
Add Markdown File for state/time questions & gitignore

---
title: "DV-State-Time"
author: "Vy Nguyen"
date: "7/16/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r echo=FALSE, message= FALSE, fig.asp=1}

library(openintro)
data("birds")
library(ggplot2)
library(patchwork)
library(ggmosaic)
library("viridis")
library(RColorBrewer) # https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/  -> Purples, BuPu; Map: BrBG
library(knitr)
library(maps)
library(mapdata)
require(plyr)
library(dplyr)
library(forcats)
library(ggwordcloud)

every_nth = function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

usa <- map_data('usa')
us.state <- map_data("state")
df.state <- as.data.frame(cbind(state.abb, state.name))
df.state$state.name <- tolower(df.state$state.name)

options(knitr.table.format = "latex")

theme_set(theme_minimal())
color_pal<-c("seagreen", "steelblue", "darkred", "orange")
color_pal_sky<-c("lightgoldenrod","honeydew","skyblue3")
color_pal_TF<-c("darksalmon","cadetblue")
color_pal_TOD<-c("paleturquoise","lightgoldenrod","plum2","midnightblue")
theme.title<-theme(
    plot.title = element_text(
      hjust = 0, # 0.5 for center
      size = 12,
      color = "steelblue",
      face = "bold"
    ),
    plot.subtitle = element_text(
      hjust = 0, # 0.5 for center
      size = 10,
      color = "gray",
      face = "italic"
    )
  )

birds_orig<-birds

factor_phase_flt<-c("Parked", "Taxi", "Take-off run", "Climb", "En Route", "Descent", "Approach", "Landing Roll")
birds$phase_of_flt<-factor(birds$phase_of_flt, levels=factor_phase_flt)

factor_struck<-c("0", "1", "2-10", "11-100", "Over 100")
birds$birds_struck<-factor(birds$birds_struck, levels=factor_struck)

factor_sky<-c("No Cloud", "Some Cloud", "Overcast")
birds$sky <- factor(birds$sky, levels=factor_sky)

factor_effect <- c("None","Other","Aborted Take-off","Engine Shut Down","Precautionary Landing")
birds$effect <- factor(birds$effect, levels=factor_effect)

birds$effectYes<- as.factor(birds$effect != "None")

birds$num_engs <- as.factor(birds$num_engs)

#birds seen: remove one wrong entry
  # which(birds$birds_seen==":-10")
birds$birds_seen[9412] <- NA

factor_seen<-c("None","2-10", "11-100")
birds$birds_seen <- factor(birds$birds_seen, levels=factor_seen) 
  # drop unused level ":-10" by assigning levels again
birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
#set Na values in birds seen to "None"
birds[is.na(birds$birds_seen),"birds_seen"] <- "None"

#get rows where birds struck is missing
df <- birds[is.na(birds$birds_struck), 
            c("remarks", "birds_seen", "effect", "birds_struck")]
no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
# set birds_struck to 0 for "no strike"
birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"

# Amount of na for each column
missing <- sapply(birds, function(x) sum(is.na(x)))
missing_df <- data.frame("column" = names(missing), 
                         "count" = missing, row.names = NULL)
names(missing_df)[2] <- "count"
```

# The Data

To begin, we first have a look at the dataset we're working with. This is the *birds* dataset form the *openintro* package. It is a collection of all collisions between aircraft in wildlife that were reported to the US Federal Aviation Administration between 1990 and 1997, with details on the circumstances of the collision. It consists of 19302 observations and 17 variables which are displayed below.

```{r echo=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
str(birds_orig)
```

We have 11 factor variables, two categorical ones (with one of them being date information), two discrete numeric variables and two continuous variables.
There are also some missing values which we will look at later.

## Some data exploration

To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:

```{r echo=FALSE, message=FALSE, warning=FALSE}
height.plot<-ggplot(birds) + 
  aes(x = height, y = ..density..) + 
  geom_histogram() +
  geom_density(alpha = 0.25)

speed.plot<-ggplot(birds) + 
  aes(x = speed, y = ..density..) + 
  geom_histogram() +
  geom_density(alpha = 0.25)

struck.plot<-ggplot(birds[!is.na(birds$birds_struck),]) +
  aes(y = birds_struck, x=speed, color=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  scale_color_brewer(palette="Blues") +
  geom_boxplot(alpha=0.2)

time.of.day.plot<-ggplot(birds[!is.na(birds$time_of_day),]) +
  aes(y = time_of_day, x=speed, color=time_of_day, fill=time_of_day) +
  scale_fill_manual(values = color_pal_TOD) +
  scale_color_manual(values = color_pal_TOD) +
  theme(legend.position = "none") +
  geom_boxplot(alpha=0.2)

effect.plot<-ggplot(birds_orig[!is.na(birds$effect),]) +
  aes(x = effect, y=speed, color=effect, fill=effect) +
  scale_fill_brewer(palette="OrRd") +
  scale_color_brewer(palette="OrRd") +
  theme(legend.position = "none") +
  geom_boxplot(alpha=0.2)

dat.plot<-ggplot(birds[!is.na(birds$birds_struck) & birds$birds_struck != "0",]) +
  aes(x = speed, y = height, color=birds_struck, fill=birds_struck) +
  geom_point() +
  facet_wrap(~birds_struck) +
  scale_fill_brewer(palette="Blues") +
  scale_color_brewer(palette="Blues") +
  theme_classic() +
  labs(
    title = "Speed and Height of collisions",
    subtitle = "In relation to the number of wildlife strucked each time",
    x = "Speed in Knots",
    y = "Height in Feet"
  ) +
  theme.title

(height.plot + speed.plot) / dat.plot
```

# Relationship state/wildlife

For the next research question we wanted to explore if there are different kinds of wildlife that aircrafts collide with depending on the state they're in: So essentially the relationship between the State and the Species.

## Species by State

To begin we looked at the **20 most frequent species** in the whole US. The figure shows the number of animals of a specific type which has been displayed in a sorted barplot. We can see that the most frequent species that appear in the data are birds which could not be identified further. Apart from those we see that most bird strikes involve birds with big populations, particularly geese and gulls in the United States. The third most frequent animal, for example, are gulls.

```{r echo=FALSE, fig.asp=0.7}
top20_plot<-ggplot(as.data.frame(sort(table(birds$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title="Top 20 Most Frequent Species",
       subtitle="Wildlife that aircrafts have collided with the most",
       x = "Number of Collisions",
       y = "Species") +
  theme.title

# Species in West Virginia collisions
california_collisions<-birds[birds$state == 'WV',]
cal_plot<-ggplot(as.data.frame(sort(table(california_collisions$species), decreasing = TRUE)[1:20]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkolivegreen") +
  labs(title="Top 20 in WV",
       subtitle="Most frequent Wildlife in West Virginia",
       x = "Number of Collisions",
       y = "") +
  theme.title

top20_plot + cal_plot
```

When we compare that with the most frequent species in West Virginia (WV), the most frequent species that aircrafts collided with there is surprisingly not a bird but white-tailed deers. And foxes are also among the list. For the rest there are minor differences, for example European Starlings, Rock Pigeons, Sparrows etc. are still one of the most frequent wildlife animals. However, one can see that for example ducks are not that frequent on in WV compared to the nationwide average. Owls also appear more there than in the US in general.

To further examine the state and species variables we could, for example look at all the collisions where gulls where involved and check what states those happened in the most. This is again displayed in barplots:

``` {r echo=FALSE}
# Number of Gulls collisions by state
gulls_collisions <- birds[birds$species == 'GULLS' & !is.na(birds$state),]
gulls_plot<-ggplot(as.data.frame(sort(table(gulls_collisions$state), decreasing = TRUE)), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill = "darkseagreen") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position = "none") + 
  labs(title="Gulls collisions by State",
       subtitle="Number of aircraft/gulls collisions by State",
       x = "Number of Collisions",
       y = "U.S State") +
  theme.title +
  scale_y_discrete(breaks=every_nth(5))

top5_gulls<-ggplot(as.data.frame(sort(table(gulls_collisions$state), decreasing = TRUE)[1:5]), aes(y=reorder(Var1, Freq), x=Freq)) + 
  geom_bar(stat="identity", fill="darkseagreen") +
  labs(subtitle="Top 10 states with most number of gulls collisions",
       x = "Number of Collisions",
       y = "") + 
  theme.title

(gulls_plot + top5_gulls)
```

When we look at the top states we see that those are all mainly US states that lie on the coastline, for example California, New York and Florida. This therefore makes sense that collisions with gulls happen there a lot since those birds are often found near where the sea is.

## Collisions per State

We also further examined the number of collisions per state which is displayed below.  
For this, we wanted to display our findings in a map for better visualization. In the following we have displayed the number of collisions, so the variable *birds_struck*, depending on the state. Since this is a factor variable we have unclassed it and mapped the levels to the numeric values below. We then grouped the data by state and summed those up to display in the U.S. map.

```{r echo=FALSE, results='asis' }
map_table<-data.frame(factor = c("1", "2", "3", "4", "5"),
                      level = c("0", "1", "2-10", "11-100", "Over 100"),
                      'mapped value' = c("0", "1", "5", "50", "100"))
kable(map_table, format="markdown", caption="Mapping of the factor variable 'birds_struck'")
```

``` {r echo=FALSE}
birds.with.statenames<-merge(x = birds, y = df.state, by.x = "state", by.y = "state.abb")
filtered.birds<-birds.with.statenames[!is.na(birds.with.statenames$birds_struck),]

filtered.birds$struck_num<-unclass(filtered.birds$birds_struck)
filtered.birds$struck_num <- mapvalues(filtered.birds$struck_num, 
                                     from=c(1,2,3,4,5), 
                                     to=c(0,1,5,50,100))

num.collisions.per.state<-aggregate(filtered.birds$struck_num, by=list(region=filtered.birds$state.name), FUN=sum)
dfs.merged<-merge(us.state, num.collisions.per.state, by = "region")

num.coll.map<-ggplot(data=dfs.merged[with(dfs.merged, order(-group, order)), ], aes(x=long, y=lat, fill=x, group=group)) + 
  geom_polygon(color = "white") + 
  scale_fill_distiller(palette="Blues", trans="reverse") +
  guides(fill=guide_colorbar(title="No. of strucks")) + 
  #theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
  #      axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  labs(x = "Longitude", y = "Latitude") +
  ggtitle('Number of strucks per State') + 
  theme.title +
  coord_fixed(1.3)

num.coll.map
```
The states with the most collisions seem to be California, Florida, New York and Texas. This is reasonable since those states are the most known ones and there is a lot of air traffic going on: People travel a lot to and from those states so it makes sense that there are the most collisions.

We then also included the variable *ac_mass* to look at:

``` {r echo=FALSE}
# Number of collisions per state
state_plot_df<-birds[birds$birds_struck != "0" & !is.na(birds$birds_struck) & !is.na(birds$state) & !is.na(birds$ac_mass),]
state_plot_df$ac_mass_renamed<-revalue(as.factor(state_plot_df$ac_mass), 
     c("1"="2250 kg or less", "2"="2251-5700 kg", "3"="5701-27000 kg", "4"="27001-272000 kg", "5"="above 272000 kg"))
state_plot<-ggplot(data = state_plot_df) +
  aes(y = fct_rev(fct_infreq(state)), fill = ac_mass_renamed) + 
  guides(fill=guide_legend(title="Mass of the aircraft")) + 
  scale_fill_brewer(palette="Purples") +
  geom_bar() + 
  labs(title="Number of collisions by State",
       subtitle="Divided by the mass of the aircraft",
       x = "Number of Collisions",
       y = "U.S State") +
  theme.title + 
  theme(legend.position="right") +
  scale_y_discrete(breaks=every_nth(5))

state_plot
```

Here, what's interesting to observe is that collisions of aircrafts with a high mass (above 272000 kg) seem to mostly happen in the State of Washington, New York and California. This could be due to the fact that those states are frequently flown to and from, so bigger aircrafts like big passenger planes are used there more than in other states. Another reason could be that there's a big aerospace industry in Washington and the Boeing Everett factory. It's said to be the best state for building boeings.

## Bird collisions vs. non-bird collisions

We further investigated by dividing the data into strucks with birds and strucks with non-birds/other wildlife. To do this we created another variable called *is.bird*: When "bird" appeared in the *remarks* or the *species* variable (along with other words), or when the height where the collision happened was more than 7 Feet above ground level, we categorized those collisions as bird collisions. The remaining ones were then classified as non-bird ones. In the following we see wordclouds of the bird and non-bird species, the size of the words in regard to their respective frequency in the data. (Top 20 for each, the number of the words have been limited due to size.)

``` {r echo=FALSE, include=FALSE}
is.bird<-grepl("BIRD|DUCK|PIGEON|HAWK|GULL|DOVE|PARROT|SWAN|GEESE|GOOSE|TURKEY|NIGHTJAR|GREBE|CARDINAL|MARTIN|BLUE JAY|CRANE|WHIP-POOR-WILL|TOWHEE|WARBLER|CORMORANT|HARRIER|BUNTING|PHEASANT|SANDPIPER|PELICAN|OSPREY|ANHINGA|WREN|MAGPIE|MANNIKIN|FINCH|ORIOLE|SNIPE|WIGEON|STARLING|AVOCET|GROUSE|BOBWHITE|CROW|TERN|VULTURE|ROBIN|OWL|FALCON|KILLDEER|GRACKLE|MYNA|PLOVER|MALLARD|SWALLOW|EGRET|KESTREL|LARK|SPARROW|HERON|EAGLE|CHICKADEE|MERGANSER", birds.with.statenames$species) | grepl("BIRD", birds.with.statenames$remarks) | birds.with.statenames$height >= 7
birds.with.statenames$is.bird<-is.bird

# Species that are bird/non bird
birds.with.statenames.filt<-birds.with.statenames[!is.na(birds.with.statenames$is.bird),]

is.bird.species<-as.data.frame(table(birds.with.statenames.filt[birds.with.statenames.filt$is.bird,]$species))
bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) +
  labs(subtitle="Frequent bird species") +
  theme(plot.subtitle = element_text(vjust = - 20))
  scale_size_area(max_size = 5)

non.bird.species<-as.data.frame(table(birds.with.statenames.filt[!birds.with.statenames.filt$is.bird,]$species))
non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
  geom_text_wordcloud_area(area_corr_power = 1) +
  labs(subtitle="Frequent non-bird species") +
  theme(plot.subtitle = element_text(vjust = - 20))
  scale_size_area(max_size = 5)
```
``` {r echo=FALSE}
bird_wordcloud + non_bird_wordcloud
```

We again can observe what we already saw earlier in the barplot for the most frequent species in general: The most frequent birds are unknown, but also gulls and a loth of other bird types can be found hee again. For the non-bird wildlife we see the white-tailed deer again which is, by far, the most frequent one in this category (visualized by the size of the word), followed by coyotes, mule deers and foxes.

Next, we want to find out if there's a difference in bird and non-bird collisions depending on what state the aircraft is in. To do this, we displayed the number of collisions and used the *is.bird* variable we created to divide the data. This is visualized in the maps below:
The left column shows the bird collisions, thr right one the non-bird ones. The upper maps show the ratio, so the proportion of bird/non-bird collisions and total collisions by state. The two bottom maps show the absolute values, so the number of collisions by state, for bird and non-bird encounters.

``` {r echo=FALSE, fig.asp=0.8}
state.coll.bird.nonbird<-as.data.frame.matrix(table(birds.with.statenames.filt$state.name, birds.with.statenames.filt$is.bird))
colnames(state.coll.bird.nonbird)<-c("nonbird", "bird")
state.coll.bird.nonbird <- cbind(state = rownames(state.coll.bird.nonbird), state.coll.bird.nonbird)
rownames(state.coll.bird.nonbird) <- 1:nrow(state.coll.bird.nonbird)

## Birds/Non-Birds in two maps next to each other
map.bird<-ggplot(state.coll.bird.nonbird, aes(map_id = state)) +
  geom_map(aes(fill = bird), map = us.state) +
  scale_fill_distiller(palette = "PuBu", trans="reverse") +
  guides(fill=guide_colorbar(title="No. of collisions")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  labs(x = "Longitude", y = "Latitude") +
  #ggtitle('Bird strucks per State') + 
  theme.title +
  theme(legend.position="bottom") +
   theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  coord_fixed(1.3)

map.non.bird<-ggplot(state.coll.bird.nonbird, aes(map_id = state)) +
  geom_map(aes(fill = nonbird), map = us.state) +
  scale_fill_distiller(palette = "PuBu", trans="reverse") +
  guides(fill=guide_colorbar(title="No. of collisions")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  labs(x = "Longitude", y = "") +
  #ggtitle('Non-bird strucks per State') + 
  theme.title +
  theme(legend.position="bottom") +
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  coord_fixed(1.3)

is.bird.state<-aggregate(birds.with.statenames.filt$is.bird, by=list(region=birds.with.statenames.filt$state.name), FUN=sum)
num.coll.per.state<-as.data.frame(table(birds.with.statenames.filt$state.name))
is.bird.state$x <- is.bird.state$x / num.coll.per.state$Freq
colnames(is.bird.state)<-c("state", "proportion_birds")

is.not.bird.state<-aggregate(!birds.with.statenames.filt$is.bird, by=list(region=birds.with.statenames.filt$state.name), FUN=sum)
is.not.bird.state$x <- is.not.bird.state$x / num.coll.per.state$Freq
colnames(is.not.bird.state)<-c("state", "proportion_non_birds")

map.bird.prop<-ggplot(is.bird.state, aes(map_id = state)) +
  geom_map(aes(fill = proportion_birds), map = us.state) +
  scale_fill_distiller(palette = "BuGn", trans="reverse") +
  guides(fill=guide_colorbar(title="Ratio")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  labs(x = "", y = "Latitude") +
  ggtitle('Bird collisions per State') + 
  theme.title +
  theme(legend.position="top") +
  coord_fixed(1.3)

map.non.bird.prop<-ggplot(is.not.bird.state, aes(map_id = state)) +
  geom_map(aes(fill = proportion_non_birds), map = us.state) +
  scale_fill_distiller(palette = "BuGn", trans="reverse") +
  guides(fill=guide_colorbar(title="Ratio")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  labs(x = "", y = "") +
  ggtitle('Non-bird collisions per State') + 
  theme.title +
  theme(legend.position="top") +
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
  coord_fixed(1.3)

(map.bird.prop + map.non.bird.prop) / (map.bird + map.non.bird)

########### Ratio of nonbird/bird collisions doesn't match
state.coll.nonbird<-select(state.coll.bird.nonbird, state, nonbird)
state.coll.nonbird$is.bird<-FALSE
colnames(state.coll.nonbird)[2]<-"Freq"
state.coll.bird<-select(state.coll.bird.nonbird, state, bird)
state.coll.bird$is.bird<-TRUE
colnames(state.coll.bird)[2]<-"Freq"
state.coll.bird.nonbird<-rbind(state.coll.nonbird, state.coll.bird)

nbp<-ggplot(state.coll.bird.nonbird, aes(map_id = state)) +
  geom_map(aes(fill = Freq), map = us.state) +
  scale_fill_distiller(palette = "BrBG") +
  guides(fill=guide_colorbar(title="No. of strucks")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  facet_wrap( ~ is.bird)
###########
```

We see that, all in all, the states that have the most bird strucks also have the most non-bird strucks. Though states in the midwest and that do not lie on the coastline have a higher ratio of non-bird collisions than those that do.  
And, what was already quite obvious in the beginning: There are a lot more aircraft-wildlife collisions with birds than other animals (Over 90%). West Virginia has the lowest ratio of all states with approximately 71.28%. This fits with our findings from earlier where we looked at the most frequent species of that state: The white-tailed deer, which is obviously a non-bird animal, was the most common one during aircraft collisions. Though West Virginia doesn't have the most non-bird collisions in absolute values - this spot is taken by Pennsylvania - it has the most in relation to the total number of collisions of the given state.

## Vulture collisions per State

Lastly, for the state/wildlife relationship, we will look at a specific animal, specifically vultures. We look for all collisions with vultures and see which states those happened in. We will look at the proportional values, so the percentage of collisions in a state where vultures were involved in:

```{r echo=FALSE}
is.vulture<-grepl("VULTURE", birds.with.statenames.filt$species)
birds.with.statenames.filt$is.vulture <- is.vulture
is.vulture<-aggregate(birds.with.statenames.filt$is.vulture, by=list(region=birds.with.statenames.filt$state.name), FUN=sum)
is.vulture$x <- is.vulture$x / num.coll.per.state$Freq
colnames(is.vulture)<-c("state", "proportion_vultures")

ggplot(is.vulture, aes(map_id = state)) +
  geom_map(aes(fill = proportion_vultures), map = us.state) +
  scale_fill_distiller(palette = "Reds", trans="reverse") +
  guides(fill=guide_colorbar(title="Ratio of vulture strucks")) + 
  expand_limits(x = us.state$long, y = us.state$lat) +
  labs(x = "Longitude", y = "Latitude") +
  ggtitle('Proportion of vulture strucks per State') + 
  theme.title +
  coord_fixed(1.3)
```

Vultures are more frequent in the South and warm regions which fits with our map above. Their habitat is not in the midwest and hence, there aren't any vulture collisions happening there.

So all in all, we can see that there is a relationship between wildlife and the state. Depending on where the aircraft collided wih wildlife the most frequent species differ from ech other. We can also extract information about the habitat of certain wildlife with the given data (e.g. vulture example).

# Timeline

For the last research question we involved the *date* variable of the dataset. We wanted to investigate the following: Have there been any changes over time regarding the wildlife collisions or are there differences depending on the time of year?  
To do this we first had to perform some feature engineering on *date*: We removed the time part using Regex and formatted it into an actual date variable. We then extracted the month, the year and the weekday into separate variables.

```{r echo=FALSE}
birds$date<-gsub(pattern="\\s?\\d{1,2}:\\d{2}:\\d{2}", replacement="", x=birds$date)
birds$year<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%Y")
birds$month<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%m")
birds$day<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%a")
birds$calendar_week<-format(as.Date(as.character(birds$date), format = "%m/%d/%Y"), "%W")
```
```{r echo=FALSE, results='asis' }
map_date<-data.frame('original date' = c("09/30/1990 0:00:00", "11/29/1993 0:00:00"),
                      date=c("09/30/1990", "11/29/1993"),
                      year = c("1990", "1993"),
                      month = c("09", "11"),
                      day=c("So", "Mo"),
                      calendar_week=c("39", "48"))
kable(map_date, format="markdown", caption="Feature Engineering of the date variable")
```

The above table shows the original *date* variable in the far left column, the other columns are the new variables that *date* has been mapped to.
We now use those to investigate the data further.  
The number of collisions is used as reference against the different date/time variables and also in regards to the *sky*, the *time_of_day* and the *birds_struck* variable.

```{r echo=FALSE}
birds.filt<-birds[!is.na(birds$birds_struck),]

# monthly grouping over all years:
month_plot<-ggplot(birds.filt) +
  aes(x = month, group=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  labs(x = "Month", y = "Number of collisions") +
  theme(axis.title.y=element_blank()) +
  guides(fill=FALSE) +
  scale_y_continuous(breaks=seq(0,3000,2000)) +
  geom_bar()
# -> more collisions during autumn/fall

# Week of the Year/Calendar Week
woy_plot<-ggplot(birds.filt) +
  aes(x = calendar_week, group=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  labs(x = "Calendar Week", y = "Number of collisions") +
  guides(fill=guide_legend(title="Birds/Wildlife strucked")) +
  geom_bar() +
  scale_y_continuous(breaks=seq(0,600,600)) +
  scale_x_discrete(breaks=every_nth(5))

# Or weekly:
week_plot<-ggplot(birds.filt) + #[as.integer(birds.filt$birds_struck) > 3,]) +
  aes(x = day, group=birds_struck, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  guides(fill=FALSE) +
  labs(x = "Weekday", y = "Prop. number of collisions") +
  geom_bar(position="fill")
# -> bird collisions not dependent on weekdays

month_plot_filt<-ggplot(birds.filt[as.integer(birds.filt$birds_struck) > 3,]) +
  aes(x = month, group=birds_struck, fill=birds_struck) +
  scale_fill_manual(values=c("#3182bd", "#08519c")) +
  guides(fill=FALSE) +
  theme(axis.title.y=element_blank()) +
  labs(x = "Month", y = "Number of collisions") +
  scale_y_continuous(breaks=seq(0, 30, 30)) +
  geom_bar()

birds.filt$first_half_year<-factor(as.factor(as.numeric(birds.filt$month) <= 6), levels=c(TRUE, FALSE))
half.year.plot<-ggplot(birds.filt[as.integer(birds.filt$birds_struck) > 3,]) +
  aes(x = as.factor(first_half_year), group=birds_struck, fill=birds_struck) +
  labs(x = "In the first half of the year", y = "Number of collisions") +
  guides(fill=guide_legend(title="Birds/Wildlife strucked")) +
  theme(legend.position = "none") +
  scale_fill_manual(values=c("#3182bd", "#08519c")) +
  geom_bar()
# way more collisions in second half of the year for birds -> mehr Zugvögel?
# But they would have to come back in spring which we can't see in the diagram?

# Months by sky
birds.filt$month<-format(as.Date(as.character(birds.filt$date), format = "%m/%d/%Y"), "%b")
birds.filt$month<-factor(factor(birds.filt$month), c("Jan","Feb", "Mär", "Apr", "Mai", "Jun", "Jul", "Aug", "Sep", "Okt", "Nov", "Dez"))
xlabels <- c("","02","","04","","06","","08","","10","","12")
#xlabels[seq(1, length(xlabels), 2)] <- ""

sky_month_plot<-ggplot(birds.filt[!is.na(birds.filt$sky),]) +
  aes(x = month, group=sky, fill=sky) +
  scale_fill_manual(values = color_pal_sky) +
  guides(fill=guide_legend(title="Sky")) +
  labs(x = "Month", y = "Prop. number of collisions") +
  geom_bar(position="fill") +
  scale_x_discrete(labels=xlabels)
# -> less overcast during summer months

# Time of Day
tod_plot<-ggplot(birds.filt[!is.na(birds.filt$time_of_day),]) +
  aes(x = month, fill=time_of_day) +
  scale_fill_manual(values = color_pal_TOD) +
  labs(x = "Month", y = "Prop. number of collisions") +
  guides(fill=guide_legend(title="Time of Day")) +
  geom_bar(position="fill") +
  scale_x_discrete(labels=xlabels)
# -> possible explanation: It gets dark later in the summer (in the US)
# -> less strucks during the night in the summer
```
```{r echo=FALSE, fig.asp=0.5}
(sky_month_plot + tod_plot)
```
We can observe that the *sky* variable changes depending on the month: There is less overcast during summer months which is reasonable since generally during summer it rains less.  
We also see that there are less collisions during the night in the summer. One possible explanation here could be that, at least in the U.S., during summer days are longer and it gets darker later. This would also match with the observation we made jut before with the *sky* variable.
So just based on this dataset alone we can also see changes in the weather depending on the season.

```{r echo=FALSE, fig.asp=1.0}
(month_plot / woy_plot) / (week_plot + half.year.plot)

#grid.arrange(month_plot, woy_plot, month_plot_filt, top = "Hawk collisions dover the Year", nrow=3) + theme.title
```

When looking at the number of collisions over the year we can see that there is an increase in collisions during autumn/fall which is especially evident when looking at the collisions where the amount of birds struck was high (where *birds_struck* is "11-100" or "Over 100"). This is displayed in the plot in the bottom right corner: There are way more collisions in the second half of the year than there are in the first half.  
This could be connected to the fact that a lot of birds are migratory birds: So once it gets colder they migrate in big flocks which then leads to more collisions with a higher amount of birds during the fall months. 
However, there is also already an increase during the summer (July, August) and also a slight increase in May, which could again be because of spring migration. For the former, people tend to travel a lot more during summer so this could also explain the increase.  
So with the *birds* dataset we are also able to visualize the migratory behavior of birds over the year.  
We also examined the proportional number of collisions based on the different weekdays but these seem, as expected, to have no influence. So the number of bird/wildlife collisions is not dependent on the days of the week.

## Grouping per Year

For the grouping over the years we will again use the *struck_num* variable from earlier which was based on the *birds_struck* column. We sum up the number of strucked birds/wildlife for each year and display this in a line/point plot. We also do the same again but also divided by each month. Both plots are displayed below.  
Fazit???

```{r echo=FALSE, fig.asp=0.8}
birds$struck_num<-unclass(birds$birds_struck)
birds$struck_num <- mapvalues(birds$struck_num, 
                                     from=c(1,2,3,4,5), 
                                     to=c(0,1,5,50,100))

birds.filtered.struck.num<-birds[!is.na(birds$struck_num),]

num.collisions.per.month<-aggregate(birds.filtered.struck.num$struck_num, by=list(month=birds.filtered.struck.num$month), FUN=sum)

strucks.per.year.month<-birds.filtered.struck.num %>%
  group_by(year, month) %>% 
  summarize_at("struck_num",sum,na.rm=TRUE)

coll.year.month<-ggplot(strucks.per.year.month, aes(x = month, y = struck_num, colour=year, group = year)) + 
  labs(x = "Month", y = "Number of strucks", 
       subtitle="(For the years 1990 - 1997)", 
       title = "Number of strucks over the months") +
  theme.title +
  geom_line() + geom_point()

strucks.per.year<-as.data.frame(table(birds.filtered.struck.num$year))
coll.year<-ggplot(strucks.per.year) + 
  aes(x = Var1, y = Freq) + 
  labs(x = "Year", y = "Number of strucks", 
       subtitle="(Aircraft-wildlife collisions from 1990 - 1997)", 
       title = "Number of strucks over the years") +
  theme.title +
  geom_line(group = 1) + geom_point()

coll.year.bar<-ggplot(birds.filtered.struck.num, aes(x = year, group = year)) + geom_bar()

coll.year / coll.year.month #/ coll.year.bar
```

We again see that for almost all years there is an increase in strucked birds in the fall and less collisions in the first half of the year. The reasons for this are mentioned above. With each year the number of collisions increases. However, the result for the years "1998" and "1999" are quite low compared to the previous years. After some research there was no valid reason for this observation as normally wildlife strikes tend to increase over time which we have seen with the previous years. Therefore we assume that, at lest in this dataset, for the years "1998" and "1999" there is data missing, so not as much strikes are reported here.  
Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.

## Hawk collisions over time

Finally, we pick a wildlife species as example: We want to investigate whether we can see changes for hawk collisions depending on the season (summer/winter) and also if there are any significant observations when we also look at the location.  
To do this we look for all entries in the dataset where hawks were involved, so where "hawk" appears in *species*. We then divide the data into two dataframes: One where the month is >= 4 and < 10 (so the summer months from April to September) and the other one with the remaining entries.
For each set we look at the *state* variable too see if the number of strikes changes depending on location. (We again use the *struck_num* variable from earlier.) The values shown are the proportional hawk strikes, so e.g. the hawk strikes in summer in relation to the total hawk strikes per state.

```{r echo=FALSE, fig.asp=1.0}
birds.with.statenames<-merge(x = birds.filtered.struck.num, y = df.state, by.x = "state", by.y = "state.abb")

is.hawk<-grepl("HAWK", birds.with.statenames$species)
filtered.hawk <- birds.with.statenames[is.hawk,]

filtered.hawk.summer<-filtered.hawk[as.numeric(filtered.hawk$month) >= 4 & as.numeric(filtered.hawk$month) < 10,]
filtered.hawk.winter<-filtered.hawk[as.numeric(filtered.hawk$month) <= 3 | as.numeric(filtered.hawk$month) >= 10,]

num.hawk.coll.per.state<-as.data.frame(table(filtered.hawk$state.name))
num.collisions.per.state.summer<-aggregate(filtered.hawk.summer$struck_num, by=list(region=filtered.hawk.summer$state.name), FUN=sum)
num.collisions.per.state.summer<-merge(num.collisions.per.state.summer, num.hawk.coll.per.state, by.x = "region", by.y="Var1")
num.collisions.per.state.summer$x <- num.collisions.per.state.summer$x / num.collisions.per.state.summer$Freq
num.collisions.per.state.winter<-aggregate(filtered.hawk.winter$struck_num, by=list(region=filtered.hawk.winter$state.name), FUN=sum)
num.collisions.per.state.winter<-merge(num.collisions.per.state.winter, num.hawk.coll.per.state, by.x = "region", by.y="Var1")
num.collisions.per.state.winter$x <- num.collisions.per.state.winter$x / num.collisions.per.state.winter$Freq

dfs.merged.summer<-merge(us.state, num.collisions.per.state.summer, by = "region", all.x = TRUE)
dfs.merged.summer$x[is.na(dfs.merged.summer$x)] <- 0
dfs.merged.winter<-merge(us.state, num.collisions.per.state.winter, by = "region", all.x = TRUE)
dfs.merged.winter$x[is.na(dfs.merged.winter$x)] <- 0

plot_summer<-ggplot(data=dfs.merged.summer[with(dfs.merged.summer, order(-group, order)), ], aes(x=long, y=lat, fill=x, group=group)) + 
  geom_polygon(color = "white") + 
  #scale_fill_continuous(low="grey", high="red") +
  scale_fill_distiller(palette = "Reds", trans="reverse") +
  guides(fill=FALSE) + 
  labs(x = "Longitude", y = "Latitude", 
       subtitle="In relation to total hawk trikes", 
       title = "Hawk strikes in Summer") +
  #ggtitle('Hawk collisions during Summer') + 
  theme.title +
  coord_fixed(1.3)

plot_winter<-ggplot(data=dfs.merged.winter[with(dfs.merged.winter, order(-group, order)), ], aes(x=long, y=lat, fill=x, group=group)) + 
  geom_polygon(color = "white") + 
  #scale_fill_continuous(low="grey", high="pink") +
  scale_fill_distiller(palette = "Reds", trans="reverse") +
  guides(fill=guide_colorbar(title="")) +
  labs(x = "Longitude", y = "Latitude", 
       subtitle="In relation to total hawk strikes", 
       title = "Hawk strikes in Winter") +
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
  #ggtitle('Hawk collisions during Winter') + 
  theme.title +
  coord_fixed(1.3)

#library(gridExtra)
#grid.arrange(plot_summer, plot_winter, top = paste("Hawk collisions depending on the season", "In relation to total hawk collisions", sep="\n"), nrow=1) + theme.title
plot_summer + plot_winter
```

We observe that in the northern U.S. states the proportion of hawk strikes during the summer is the highest. Compared to the winter months the ratio of hawk strikes there is the lowest. And the ratio in the Southeast region increases slightly. And overall there seem to be less hawk strikes during the winter.
This again can be explained with what was mentioned earlier: Hawks are migratory birds as well. During the winter months they travel to the southern regions, hence why there are almost no hawk strikes in the North during winter.

By visualizing our findings in a map and dividing by the date variable we were able to further show the migratory behavior of the birds in the U.S. and how this influences the number of collisions and strikes with aircrafts.

### 2021-07-18 state_wildlife.R

@@ -154,3 +154,16 @@ ggplot(data=dfs.merged.3[with(dfs.merged.3, order(-group, order)), ], aes(x=long
   coord_fixed(1.3)
 
 # Proportionale Kollisionen?
+
+non.bird.species<-as.data.frame(table(birds.with.statenames[birds.with.statenames$is.bird & !is.na(birds.with.statenames$is.bird),]$species))
+ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),], aes(label = Var1, size=Freq)) +
+  geom_text_wordcloud_area() +
+  scale_size_area(max_size = 10)
+
+
+data("love_words_small")
+set.seed(42)
+ggplot(love_words_small, aes(label = word, size = speakers)) +
+  geom_text_wordcloud() +
+  scale_size_area(max_size = 40) +
+  theme_minimal()

### 2021-07-18 state-time.Rmd
Updated State/Time Markdown

@@
 output:
-  pdf_document: default
-  html_document: default
+  pdf_document:
+    toc: true
+    toc_depth: 3
+  html_document:
+    toc: true
+    toc_depth: 3
 ---
@@
 color_pal_TF<-c("darksalmon","cadetblue")
-color_pal_TOD<-c("paleturquoise","lightgoldenrod","plum2","midnightblue")
+color_pal_TOD<-c("#fed98e","#fff7bc","#d4b9da","#43a2ca")
 theme.title<-theme(
@@
       size = 12,
-      color = "steelblue",
+      color = "#016c59",
       face = "bold"
@@
       size = 10,
-      color = "gray",
+      color = "gray50",
       face = "italic"
@@
 ```
+\newpage
 
@@
 It is a collection of all collisions between aircraft in wildlife that were reported to the US Federal Aviation Administration between 1990 and 1997, with details on the circumstances of the collision.
-It consists of 19302 observations and 17 variables which are displayed below.
+It consists of 19302 observations and 17 variables which are given in the table below.
 
 ```{r echo=FALSE}
-knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
-str(birds_orig)
+#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
+#str(birds_orig)
+data_table<-data.frame(Variable = c("opid", "operator", "atype", "remarks", "phase_of_flt", "ac_mass", "num_engs", "date", "time_of_day", "state", "height", "speed", "effect", "sky", "species", "birds_seen", "birds_struck"),
+                      Description = c("Three letter identification code for the operator (carrier) of the aircraft.", 
+                                      "Name of the aircraft operator.", "Make and model of aircraft.", "Verbal remarks regarding the collision.", "hase of the flight during which the collision occurred.", "Mass of the aircraft classified in kg.", "Number of engines on the aircraft.", "Date of the collision.", "Light conditions", "Two letter abbreviation of the US state in which the collision occurred.", "Feet above ground level.", "Knots (indicated air speed).", "Effect on flight.", "Type of cloud cover, if any.", "Common name for bird or other wildlife.", "Number of birds/wildlife seen by pilot", "Number of birds/wildlife struck."),
+                      Type = c("Factor w/ 285 levels", "Factor w/ 285 levels", "Factor w/ 284 levels", "Categorical", "Factor w/ 8 levels", "Discrete (1-5)", "Discrete (1-4)", "Categorical", "Factor w/ 4 levels", "Factor w/ 58 levels", "Continuous (0-32500)", "Continuous (0-400)", "Factor w/ 5 levels", "Factor w/ 3 levels", "Factor w/ 241 levels", "Factor w/ 3 levels", "Factor w/ 5 levels"))
+kable(data_table, format="markdown", caption="Variables of the birds dataset")
 ```
@@
 
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+
 ## Some data exploration
 
-To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:
+#To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
 height.plot<-ggplot(birds) + 
@@
 
-(height.plot + speed.plot) / dat.plot
+#(height.plot + speed.plot) / dat.plot
 ```
+\newpage
 
@@
   theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
-  labs(title="Top 20 Most Frequent Species",
-       subtitle="Wildlife that aircrafts have collided with the most",
+  labs(title="Top 20 in the U.S.",
+       subtitle="Species that aircrafts have collided with the most",
        x = "Number of Collisions",
        y = "Species") +
+  theme(plot.title.position = "plot") +
   theme.title
@@
   geom_bar(stat="identity", fill="darkolivegreen") +
-  labs(title="Top 20 in WV",
-       subtitle="Most frequent Wildlife in West Virginia",
+  labs(title="Top 20 in West Virginia",
+       subtitle="Most frequent Wildlife species in WV",
        x = "Number of Collisions",
        y = "") +
+  theme(plot.title.position = "plot") +
   theme.title
@@
 ```
+
 The states with the most collisions seem to be California, Florida, New York and Texas.
@@
 In the following we see wordclouds of the bird and non-bird species, the size of the words in regard to their respective frequency in the data.
-(Top 20 for each, the number of the words have been limited due to size.)
+(Top 30 for each, the number of the words have been limited due to size and readability.)
 
@@
 
+set.seed(43)
 is.bird.species<-as.data.frame(table(birds.with.statenames.filt[birds.with.statenames.filt$is.bird,]$species))
-bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
-  geom_text_wordcloud_area(area_corr_power = 1) +
-  labs(subtitle="Frequent bird species") +
-  theme(plot.subtitle = element_text(vjust = - 20))
+bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
+  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="bird.png")) +
+  labs(title="Frequent bird species") +
+  theme(plot.title = element_text(
+      vjust = -30,
+      hjust = 0.5, # 0.5 for center
+      size = 10,
+      color = "darkcyan",
+      face = "bold"
+    )) +
   scale_size_area(max_size = 5)
 
+set.seed(43)
 non.bird.species<-as.data.frame(table(birds.with.statenames.filt[!birds.with.statenames.filt$is.bird,]$species))
-non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:20,], aes(label = Var1, size=Freq)) +
-  geom_text_wordcloud_area(area_corr_power = 1) +
-  labs(subtitle="Frequent non-bird species") +
-  theme(plot.subtitle = element_text(vjust = - 20))
+non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
+  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="deer.png")) +
+  labs(title="Frequent non-bird species") +
+  theme(plot.title = element_text(
+      vjust = -30,
+      hjust = 0.5, # 0.5 for center
+      size = 10,
+      color = "darkcyan",
+      face = "bold"
+    )) +
   scale_size_area(max_size = 5)
@@
 vulture example).
+\newpage
 
 # Timeline
+
+## The date variable
 
@@
 We now use those to investigate the data further.
+
+## Monthly and weekly grouping
+
 The number of collisions is used as reference against the different date/time variables and also in regards to the *sky*, the *time_of_day* and the *birds_struck* variable.
@@
   labs(x = "Month", y = "Number of collisions") +
-  theme(axis.title.y=element_blank()) +
+  #theme(axis.title.y=element_blank()) +
   guides(fill=FALSE) +
@@
   aes(x = as.factor(first_half_year), group=birds_struck, fill=birds_struck) +
-  labs(x = "In the first half of the year", y = "Number of collisions") +
+  labs(x = "Is first half of the year", y = "Number of collisions") +
   guides(fill=guide_legend(title="Birds/Wildlife strucked")) +
@@
 ```
+
 We can observe that the *sky* variable changes depending on the month: There is less overcast during summer months which is reasonable since generally during summer it rains less.  
@@
 Both plots are displayed below.
-Fazit???
 
@@
 We again see that for almost all years there is an increase in strucked birds in the fall and less collisions in the first half of the year.
-The reasons for this are mentioned above.
+The reasons for this are the same as above.
 With each year the number of collisions increases.
 However, the result for the years "1998" and "1999" are quite low compared to the previous years.
-After some research there was no valid reason for this observation as normally wildlife strikes tend to increase over time which we have seen with the previous years.
-Therefore we assume that, at lest in this dataset, for the years "1998" and "1999" there is data missing, so not as much strikes are reported here.
+After some research we found there was no valid reason for this observation as normally wildlife strikes tend to increase over time as which we have seen with the previous years.
+Therefore we assume that, at least in this dataset, for the years "1998" and "1999" there are data/reports missing.
+Especially since the number of collisions for 1999 in total is very close to 0 which is very unlikely to be real.
 Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.
 
-## Hawk collisions over time
+## Hawk collisions over time and per State
 

### 2021-07-18 aircraft_models.R
Added Plot conclusions till t-test.

@@ -30,14 +30,13 @@ birds[grep(".*EXPRESS.*", birds$operator),]
 plot(machines$ac_mass, machines$num_engs )
 
 #create aircraft statistics
-install.packages("dplyr")
-install.packages("openintro")
+#install.packages("dplyr")
+#install.packages("openintro")
 require(dplyr)
 require(openintro)
 data("birds")
 
-df <- birds[!is.na(birds$height) & !is.na(birds$speed),]
-machines <- df %>%
+machines <- birds[!is.na(birds$height) & !is.na(birds$speed),] %>%
   group_by(atype) %>%
   summarise(max_speed = max(speed),
             max_height = max(height),
@@ -45,6 +44,8 @@ machines <- df %>%
             ac_mass = median(as.integer(ac_mass)))
 
 machines <- as.data.frame(machines)
+aircraft_strike_count <- birds %>% group_by(atype) %>% tally()
+machines<-merge(x = machines, y = aircraft_strike_count, by.x = "atype", by.y = "atype")
 
 machines$num_engs <- factor(machines$num_engs, levels=1:4)
 machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
@@ -56,6 +57,14 @@ machines$ac_mass2 <- factor(machines$ac_mass2, levels=1:3)
 table(machines$ac_mass2)
 table(machines$ac_mass)
 
+top10_aircrafts <- machines %>%
+arrange(desc(n))%>%
+head(10)%>%
+ggplot(aes(x=reorder(atype, n), y=n))+
+  geom_bar(stat="identity", fill="snow3") +
+  coord_flip()
+
+
 ggplot(machines[!is.na(machines$ac_mass),]) +
   aes(x=max_speed, y=max_height, color = ac_mass) +
   geom_point()
@@ -64,15 +73,33 @@ ggplot(machines[!is.na(machines$num_engs),]) +
   aes(x=max_speed, y=max_height, color = num_engs) +
   geom_point()
 
-ggplot(machines[!is.na(machines$ac_mass),]) +
+ggplot(machines[!is.na(machines$ac_mass) & machines$ac_mass!="5",]) +
   aes(x=max_height, y=max_speed, color = ac_mass) +
-  geom_point()
+  geom_point() +
+  geom_smooth(se=FALSE)
+
+ggplot(machines[!is.na(machines$ac_mass) & machines$ac_mass!="5",]) +
[…]

### 2021-07-18 state-time.Rmd
Minor changes and added bibliography

@@
     toc_depth: 3
+bibliography: references.bib  
 ---
@@
 
-# The Data
+# The Birds Data
 
 To begin, we first have a look at the dataset we're working with.
-This is the *birds* dataset form the *openintro* package.
-It is a collection of all collisions between aircraft in wildlife that were reported to the US Federal Aviation Administration between 1990 and 1997, with details on the circumstances of the collision.
-It consists of 19302 observations and 17 variables which are given in the table below.
+This is the *birds* dataset form the *openintro* package [@birds_doc]: It is a collection of all **collisions between aircraft and wildlife** that were reported to the US Federal Aviation Administration **between 1990 and 1997**, with details on the circumstances of the collision.
+It consists of **19302 observations and 17 variables** which are given in the table below.
 
@@
 #str(birds_orig)
-data_table<-data.frame(Variable = c("opid", "operator", "atype", "remarks", "phase_of_flt", "ac_mass", "num_engs", "date", "time_of_day", "state", "height", "speed", "effect", "sky", "species", "birds_seen", "birds_struck"),
+data_table<-data.frame(
+  Variable = c("opid", "operator", "atype", "remarks", "phase_of_flt", "ac_mass", "num_engs", "date", "time_of_day", 
+               "state", "height", "speed", "effect", "sky", "species", "birds_seen", "birds_struck"),
   Description = c("Three letter identification code for the operator (carrier) of the aircraft.", 
-                                      "Name of the aircraft operator.", "Make and model of aircraft.", "Verbal remarks regarding the collision.", "hase of the flight during which the collision occurred.", "Mass of the aircraft classified in kg.", "Number of engines on the aircraft.", "Date of the collision.", "Light conditions", "Two letter abbreviation of the US state in which the collision occurred.", "Feet above ground level.", "Knots (indicated air speed).", "Effect on flight.", "Type of cloud cover, if any.", "Common name for bird or other wildlife.", "Number of birds/wildlife seen by pilot", "Number of birds/wildlife struck."),
-                      Type = c("Factor w/ 285 levels", "Factor w/ 285 levels", "Factor w/ 284 levels", "Categorical", "Factor w/ 8 levels", "Discrete (1-5)", "Discrete (1-4)", "Categorical", "Factor w/ 4 levels", "Factor w/ 58 levels", "Continuous (0-32500)", "Continuous (0-400)", "Factor w/ 5 levels", "Factor w/ 3 levels", "Factor w/ 241 levels", "Factor w/ 3 levels", "Factor w/ 5 levels"))
+                  "Name of the aircraft operator.", 
+                  "Make and model of aircraft.", 
+                  "Verbal remarks regarding the collision.", 
+                  "Phase of the flight during which the collision occurred.",
+                  "Mass of the aircraft in kg classified.", 
+                  "Number of engines on the aircraft.", 
+                  "Date of the collision. (MM/DD/YYYY 0:00:00)", 
+                  "Light conditions.", 
+                  "Two letter abbreviation of the US state in which the collision occurred.", 
+                  "Feet above ground level.", 
+                  "Knots (indicated air speed).", 
+                  "Effect on flight.", 
+                  "Type of cloud cover, if any.", 
+                  "Common name for bird or other wildlife.", 
+                  "Number of birds/wildlife seen by pilot.", 
+                  "Number of birds/wildlife struck."),
+  Type = c("Factor w/ 285 levels", "Factor w/ 285 levels", "Factor w/ 284 levels", "Categorical", "Factor w/ 8 levels", "Discrete (1-5)", 
+           "Discrete (1-4)", "Categorical", "Factor w/ 4 levels", "Factor w/ 58 levels", "Continuous (0-32500)", "Continuous (0-400)", 
+           "Factor w/ 5 levels", "Factor w/ 3 levels", "Factor w/ 241 levels", "Factor w/ 3 levels", "Factor w/ 5 levels"))
 kable(data_table, format="markdown", caption="Variables of the birds dataset")
@@
 
-# Relationship state/wildlife
+# Is there a Relationship between State and Wildlife?
 
@@
 
-We also further examined the number of collisions per state which is displayed below.  
+We also further examined the **number of collisions per state** which is displayed below.  
 For this, we wanted to display our findings in a map for better visualization.
@@
 
-We further investigated by dividing the data into strucks with birds and strucks with non-birds/other wildlife.
+We further investigated by dividing the data into **strucks with birds and strucks with non-birds/other wildlife**.
 To do this we created another variable called *is.bird*: When "bird" appeared in the *remarks* or the *species* variable (along with other words), or when the height where the collision happened was more than 7 Feet above ground level, we categorized those collisions as bird collisions.
@@
 Lastly, for the state/wildlife relationship, we will look at a specific animal, specifically vultures.
-We look for all collisions with vultures and see which states those happened in.
+We look for all **collisions with vultures** and see which states those happened in.
 We will look at the proportional values, so the percentage of collisions in a state where vultures were involved in:
@@
 So all in all, we can see that there is a relationship between wildlife and the state.
-Depending on where the aircraft collided wih wildlife the most frequent species differ from ech other.
+Depending on where the aircraft collided with wildlife the most frequent species differ from each other.
 We can also extract information about the habitat of certain wildlife with the given data (e.g.
@@
 
-# Timeline
+# Timeline - Does time matter?
 
@@
 For the last research question we involved the *date* variable of the dataset.
-We wanted to investigate the following: Have there been any changes over time regarding the wildlife collisions or are there differences depending on the time of year?
+We wanted to investigate the following: Have there been any **changes over time regarding the wildlife collisions** or are there **differences depending on the time of year**?
 To do this we first had to perform some feature engineering on *date*: We removed the time part using Regex and formatted it into an actual date variable.
@@
 This would also match with the observation we made jut before with the *sky* variable.
+There are correspondingly also more collisions during the day in the summer which could also be because there are more flights happening compared to thee winter time.
+Therefore the probability of more collisions is higher then, too.
 So just based on this dataset alone we can also see changes in the weather depending on the season.
@@
 
-For the grouping over the years we will again use the *struck_num* variable from earlier which was based on the *birds_struck* column.
+For the **grouping over the years** we will again use the *struck_num* variable from earlier which was based on the *birds_struck* column.
 We sum up the number of strucked birds/wildlife for each year and display this in a line/point plot.
@@
   labs(x = "Month", y = "Number of strucks", 
-       subtitle="(For the years 1990 - 1997)", 
+       subtitle="(For the years 1990 - 1999)", 
        title = "Number of strucks over the months") +
@@
   labs(x = "Year", y = "Number of strucks", 
-       subtitle="(Aircraft-wildlife collisions from 1990 - 1997)", 
+       subtitle="(Aircraft-wildlife collisions from 1990 - 1999)", 
        title = "Number of strucks over the years") +
@@
 However, the result for the years "1998" and "1999" are quite low compared to the previous years.
-After some research we found there was no valid reason for this observation as normally wildlife strikes tend to increase over time as which we have seen with the previous years.
+After some research we found there was no valid reason for this observation as normally wildlife strikes tend to increase over time like as we have seen with the previous years.
 Therefore we assume that, at least in this dataset, for the years "1998" and "1999" there are data/reports missing.
 Especially since the number of collisions for 1999 in total is very close to 0 which is very unlikely to be real.
+When looking at the documentation of the Birds dataset [@birds_doc] it also says data from 1999 to 1997.
+So this is another indicator supporting the claim.
 Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.
 
-## Hawk collisions over time and per State
+## Hawk collisions over Time and by State
 
-Finally, we pick a wildlife species as example: We want to investigate whether we can see changes for hawk collisions depending on the season (summer/winter) and also if there are any significant observations when we also look at the location.
+Finally, we pick a wildlife species as example: We want to investigate whether we can see changes for **hawk collisions depending on the season (summer/winter)** and also if there are any significant observations when we also look at the **location**.
 To do this we look for all entries in the dataset where hawks were involved, so where "hawk" appears in *species*.
@@
 and how this influences the number of collisions and strikes with aircrafts.
+
+# References

### 2021-07-18 aircraft_wildlife_doc.Rmd

@@
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
 ---
@@
 
+Cleaning birds_seen variable: The 9412. observation had one wrong entry in the birds seen variable which has been replaced with NA, because we did not want to introduce a new level for one entry.
+
+Factor Levels: The factor levels of the following variables have been ordered for visualization purposes
+
+* Order from low to high: 
+  + birds_struck
+  + birds_seen
+  + sky (by cloudiness)
+  + ac_mass
+  + num_engs
+
+* Individual ordering: 
+  + effect: reordering to put "None" effect first and seperate it from the "effect", to achieve a separation between "None" effect and effects
+  + phase_of_flt: ordering according from parking, over start of flight to end of flight
+
 ```{r echo=FALSE, message= FALSE, fig.asp=1}
@@
 library(ggmosaic)
+library("ggwordcloud")
 library("viridis")
 library(RColorBrewer) # https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/  -> Purples, BuPu; Map: BrBG
+library(dplyr)
 
@@
 
+birds$ac_mass <- as.factor(birds$ac_mass)
 birds$num_engs <- as.factor(birds$num_engs)
@@
 birds$birds_seen[9412] <- NA
-
 factor_seen<-c("None","2-10", "11-100")
@@
   # drop unused level ":-10" by assigning levels again
-birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
-#set Na values in birds seen to "None"
-birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
+
 #\newpage
@@
 ## Handling missing values
+
+birds_struck: We are using the "remarks" variable to check for spare parts from which we can determine that no wild life has been strike.
+By doing this we were able to decrease the missing values by 19 (from 39 to 20).
+
+birds_seen: We assume that missing values in the "birds_seen" variable indicate that the pilot did not see wildlife before the strike occurred.
+Thus, we set all NA values in the "birds_seen" variable as "None".
+This leads to `r sum(is.na(birds$birds_seen))` values which are set as "None".
 
@@
 birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
+
+#birds_seen: set NA as None
+birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
+#set Na values in birds seen to "None"
+birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
 
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
 ```
 
-# Analysis of strikes (birds_struck)
+# 1. Aircraft models
+
+The dataset contains `r length(unique(birds$atype))` different aircraft models.
+We extracted the overall maximum speed and height for each aircraft to get an overview of the aircraft models and the relationship between speed and height.
+The following left plot shows the Top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the Top 50 aircrafts colored by their ac_mass.
+
+```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
+
+machines <- birds[!is.na(birds$height) & !is.na(birds$speed),] %>%
+  group_by(atype) %>%
+  summarise(max_speed = max(speed),
+            max_height = max(height),
+            num_engs = median(as.integer(num_engs)),
+            ac_mass = median(as.integer(ac_mass)))
+
+machines <- as.data.frame(machines)
+aircraft_strike_count <- birds %>% group_by(atype) %>% tally()
+machines<-merge(x = machines, y = aircraft_strike_count, by.x = "atype", by.y = "atype")
+
+machines$num_engs <- factor(machines$num_engs, levels=1:4)
+machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
+
+machines$ac_mass2 <- ifelse(as.integer(machines$ac_mass)<3,1,ifelse(as.integer(machines$ac_mass)==3,2,3))
+machines$ac_mass2 <- factor(machines$ac_mass2, levels=1:3)
+
+top10_aircrafts_bar <- machines %>%
+arrange(desc(n)) %>%
+head(10) %>%
+ggplot(aes(x=reorder(atype, n), y=n, fill = ac_mass))+
+  geom_bar(stat="identity") +
+  theme(legend.position="none") +
+  labs(title= "Top 10 frequent strikes by aircraft", x="aircraft model") +
+  coord_flip()
+
+set.seed(43)
+top50 <- machines[order(-machines$n)[1:50],] %>% 
+  mutate(angle = 30 * sample(-2:2, n(), replace = TRUE, prob = c(1, 1, 4, 1, 1)))
+top50_aircrafts_wordcloud <- ggplot(top50) +
+  aes(label = atype, size=n, color=ac_mass, angle=angle) +
+  geom_text_wordcloud_area(
+    mask = png::readPNG(source="C:/Users/Alex/Documents/flugzeug6-3.png"),
+    rm_outside = TRUE) +
+  scale_size_area(max_size = 3) +
+  scale_color_brewer(palette="BuPu") +#scale_color_continuous("Blues") +
+  #ggtitle("top 50 aircraft models causing strikes")
+  theme(plot.title = element_text(vjust=-20)) +
+  labs(title="Top 50 aircrafts causing strikes")
+
+aircraft_speed_height <- ggplot(top50[!is.na(top50$ac_mass),]) +
+
+#aircraft_speed_height <- ggplot(machines[!is.na(machines$ac_mass),]) +
+  aes(x=max_speed, y=max_height, color = ac_mass) +
+  geom_point() +
+  geom_smooth(method = lm, aes(x=max_speed, y = max_height, group = "black", fill = "black", ), se = FALSE) +
+  theme(legend.position="top") +
+  scale_color_brewer(palette="BuPu") +
+  guides(fill = FALSE)
+
+top50_aircrafts_wordcloud + aircraft_speed_height
+```
+
+The left plot shows that especially the B-737 class (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727.
+"B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas.
+The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes.
+Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this.
+But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.
+
+From the right plot we can see, that aircraft models with higher max speed are also able to fly higher.
+This can be also seen from the linear model, that has been estimated by geom_smooth.
+The correlation of max_speed and height for the Top50 models is: `r cor(top50$max_speed, top50$max_height)`, which is positive and quite high.
+In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height and that the lighter aircrafts under 5700 kg (ac_mass=2) or even 27000 kg (ac_mass=3) are less prominent.
+The group with ac_mass=4 (27001-272000 kg) is the most prominent group under the Top50 aircrafts.
+Two aircraft models with ac_mass = 5 are also present in the Top50 which are the only aircrafts with ac_mass=5.
+
+The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all aircraft models (n=`r nrow(machines)`.
+Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.
+
+```{r echo=FALSE, fig.asp=0.5}
+machines_c <- machines[complete.cases(machines),]
+p <- ggplot(machines[!is.na(machines$num_engs) & !is.na(machines$ac_mass),]) +
+  aes(x=max_speed, y=max_height, color = ac_mass) +
+  geom_point() +
+  scale_color_brewer(palette="BuPu") +
+  theme_bw()
+
+p + facet_grid(. ~ num_engs)
+```
+
+From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines.
+We can also see that lighter machines tend to have less engines first group includes nearly only machines with ac_mass=1, second group is mixed with ac_mass 2-4 and the last two groups include mostly aircrafts with ac_mass=4.
+Looking at the contingency table also confirms that:
+```{r echo=FALSE, fig.asp=0.5}
+table(machines$num_engs, machines$ac_mass)
+```
+Interesting is that for aircrafts with num_engs=1, there is no model with a speed > 200 and height > 10000.
+So we could conclude from that this is the maximum height and speed possible with one engine.
+
+# 2. Analysis of influences for birds_struck, birds_seen and effect
+In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft.
+We will first start with the analysis of strikes.
+
+## 2.1 Analysis of strikes (birds_struck)
+
+The following plot on the left shows the amount of birds struck out birds_seen.
+On the top right we can see the amount of collisions by each phase of flight and on the bottom right we can see the amount of collisions by time of day.
+For coloring we use the birds_struck variable.
 
@@
   scale_fill_brewer(palette="Blues") +
-  geom_bar()
+  geom_bar() +
+  coord_flip() +
+  theme(legend.position="none")
 
+# Most collisions during the day #genauer betrachten
+birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
+  aes(x=time_of_day, fill=birds_struck) + 
+  scale_fill_brewer(palette="Blues") +
+  geom_bar() +
+  coord_flip() +
+  theme(legend.position="none")
+
+### Birds struck out of birds seen
+birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
+  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
+  scale_fill_brewer(palette="Blues") +
+  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank(), legend.position="top")
+
+birds_struck_seen + birds_strike_phase / birds_strike_time
+## descent and en-route are about equal 
+```
+
+birds_struck out of birds_seen: The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
+So most often we do not see a birds before a strike happens.
+But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones.
+Most often 1 bird is struck.
+Most interesting here is that if the pilot sees birds for the group birds seen = 2-10 , he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+The event of hitting less birds than seen will later be analyzed more in detail.
+
+The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens the Landing, Climb and Take-Off phase.
+
+From the bottom right plot we can see that strikes happen most often during the Day with over 10500 occurences followed by Night with about 4500 occurences.
+Interesting to see is that strikes with larger groups of birds (11-100)
+
+We will now have a look on the influence of the sky variable for the different amounts of birds struck.
+
+```{r echo=FALSE, fig.asp=0.5}
 #relationship of amount of birds struck with clouds?
@@
 
-# Most collisions during the day #genauer betrachten
-birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
-  aes(x=time_of_day, fill=birds_struck) + 
+birds_strike_ac_mass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$birds_struck) & (birds$birds_struck=="0" | as.integer(birds$birds_struck)>3),])+
+aes(x=ac_mass, fill=birds_struck)+
   scale_fill_brewer(palette="Blues") +
-  geom_bar()
+  #scale_fill_manual(values = color_pal_sky)+
+  geom_bar(position = "fill")
 
-### Birds struck out of birds seen
-birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
-  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
-  scale_fill_brewer(palette="Blues") +
-  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())
+#birds_strike_ac_mass <- ggplot(birds[!is.na(birds$birds_struck) & !is.na(birds$ac_mass),]) +
+#  aes(x=ac_mass, fill=birds_struck)+
+#  scale_fill_brewer(palette="Blues") +
+  #scale_fill_manual(values = color_pal_sky)+
+#  geom_bar(position = "fill")
 
-(birds_strike_phase + birds_strike_sky) / (birds_strike_time + birds_struck_seen)
-## descent and en-route are about equal 
+birds_strike_sky +birds_strike_ac_mass
 ```
+
+From the plot we can see that with increasing number of birds_struck the proportion of overcast increases and the proportion of no clouds decreases.
+The proportion of some clouds is more or less the same across all groups.
+From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy.
+
 TODO: add Chi-squared independence test for sky and birds struck
@@
 
-the higher the amount of birds struck the more clouds
+We will now have a look on the height variable together with phase of flight.
 
-## Closer look on En Rout and Descent
+#### Closer look on En Rout and Descent
+
+The plot on the left shows a boxplot of height for flight phases, which are above ground and on the right side we will have a closer on the height for the flight phases "En Route" and "Descent".
+
 ```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
@@
 
+The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
+The other two boxplots for "En Route" and "Descent" are more far away from ground.
+Interesting here is that the 25% quantile of the Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
+
+We are performing a t.test to compare the two groups and check for equality.
+
 t.test
@@
 t.test(height~phase_of_flt, data = birds[as.integer(birds$phase_of_flt)>4 & as.integer(birds$phase_of_flt)<7 & !is.na(birds$phase_of_flt),])
+```
 
-```
-For 0 birs struck we have highest percentage of no clouds, so clouds might have an affect on avoiding striking birds. Its about 20% higher compared to 1 bird struck.
-for birdsstruck by sky we can see that with increasing amount of birds the sky tends to be more more overcast, as no clouds and some clouds decreases
+The T-Test also show that the group means are different from each other.
+But surprisingly the mean in Group "Descent" is higher than mean of "En Route".
+One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach.
+So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route".
+The main difference between these flight phases is that the aircraft is not flying parralel to the ground when being in phase "Descent".
 
-## density birdsstuck by height
+## density birdsstuck by height for flight phases above ground (Climb, En Route, Descent, Approach)
 ```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
@@
   scale_fill_brewer(palette="Blues") +
+  theme(legend.position = "top") + 
   xlim(0, 3500)
 density
-
 ```
 
-## Analysis of birds seen
+Peek for "Over 100" birds struck at a height of about 1800 foot is due to the small amount of observations from that category. But nevertheless 8 out of 17 strikes happened here.
+
+### 2.2.1 Analysis of birds seen
 
@@
 ```
+
 **Time of Day:** Based on the assumption that Missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is `r sum(birds$birds_seen!="None")/nrow(birds)`.
-The contingencytable looks like the following: `addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean)`.
+The contingencytable looks like the following:
+
+```{r echo=FALSE, fig.asp=0.5}
+addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean) 
+```
 
@@
 ```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
-par(mfrow=c(1,2))
-spineplot(birds$height, birds$birds_seen, col = rev(brewer.pal(n=3,name="BuGn")))
-spineplot(birds$speed, birds$birds_seen, col = rev(brewer.pal(n=3,name="BuGn")))
+par(mfrow=c(1,2),las = 2, cex.axis = 0.75, mar = c(7, 4.1, 4.1, 2.1))
+spineplot(birds$height, birds$birds_seen, ylab = "birds_seen", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
+spineplot(birds$speed, birds$birds_seen, ylab = "birds_seen", xlab = "speed",col = rev(brewer.pal(n=3,name="BuGn")))
 ```
+
 The higher the speed & height, the less likely to see a bird before collision
@@
 
-## effect when birdsstruck < birds seen
+### 2.2.2 effect when birdsstruck < birds seen
 
@@
   scale_fill_manual(values = color_pal_TF)+
-  geom_bar(position = "fill")
+  geom_bar(position = "fill") +
+  coord_flip() +
+  theme(legend.position="top")
 
@@
   scale_fill_manual(values = color_pal_TF)+
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())+
+  ylab("")
 
@@
   scale_fill_manual(values = color_pal_TF)+
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  theme(legend.position="none", axis.text.y=element_blank(),axis.ticks.y = element_blank())+
+  ylab("")
 
-seen_vs_strike_phase / (seen_vs_strike_ac_mass + seen_vs_strike_num_engs)
+seen_vs_strike_phase + seen_vs_strike_ac_mass / seen_vs_strike_num_engs
 ```
+
 the num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
@@
 
-## Analysis of effect variable
+## 2.3 Analysis of effect variable
 
@@
 ```
+
 check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured

### 2021-07-18 draft_final.Rmd
Draft final Rmd + PDF & updated bibliography

--- 2021-07-18 state-time.Rmd
@@
 ---
-title: "DV-State-Time"
-author: "Vy Nguyen"
-date: "7/16/2021"
+title: "Aircraft Wildlife Collisions"
+author: "Alexander Flick & Vy Nguyen"
+date: "21/07/2021"
 output:
-  pdf_document:
+  html_document:
     toc: true
     toc_depth: 3
-  html_document:
+  pdf_document:
     toc: true
@@
 
-```{r echo=FALSE, message= FALSE, fig.asp=1}
+```{r echo=FALSE, message= FALSE}
 
@@
 library(ggmosaic)
-library("viridis")
-library(RColorBrewer) # https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/  -> Purples, BuPu; Map: BrBG
+library(RColorBrewer)
 library(knitr)
@@
     plot.title = element_text(
-      hjust = 0, # 0.5 for center
+      hjust = 0, 
       size = 12,
@@
     plot.subtitle = element_text(
-      hjust = 0, # 0.5 for center
+      hjust = 0, 
       size = 10,
@@
   )
-
-birds_orig<-birds
 
@@
   # drop unused level ":-10" by assigning levels again
-birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
-#set Na values in birds seen to "None"
-birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
+```
 
-#get rows where birds struck is missing
-df <- birds[is.na(birds$birds_struck), 
-            c("remarks", "birds_seen", "effect", "birds_struck")]
-no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
-# set birds_struck to 0 for "no strike"
-birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
-
-# Amount of na for each column
-missing <- sapply(birds, function(x) sum(is.na(x)))
-missing_df <- data.frame("column" = names(missing), 
-                         "count" = missing, row.names = NULL)
-names(missing_df)[2] <- "count"
-```
 \newpage
 
-# The Birds Data
+# 1. The Data
+
+## 1.1 The Birds Dataset
 
@@
 ```{r echo=FALSE}
-#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=10), tidy=TRUE)
-#str(birds_orig)
 data_table<-data.frame(
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
+## 1.2 Data preparation
 
-## Some data exploration
+Cleaning birds_seen variable: The 9412. observation had one wrong entry in the birds seen variable which has been replaced with NA, because we did not want to introduce a new level for one entry.
 
-#To start the analysis we will explore the data and look at some of the variables. We can, for example, plot *height* and *speed* as well as the factor variable *birds_struck*:
+Factor Levels: The factor levels of the following variables have been ordered for visualization purposes
 
-height.plot<-ggplot(birds) + 
-  aes(x = height, y = ..density..) + 
-  geom_histogram() +
-  geom_density(alpha = 0.25)
+* Order from low to high: 
+  + birds_struck
+  + birds_seen
+  + sky (by cloudiness)
+  + ac_mass
+  + num_engs
 
-speed.plot<-ggplot(birds) + 
-  aes(x = speed, y = ..density..) + 
-  geom_histogram() +
-  geom_density(alpha = 0.25)
+* Individual ordering: 
+  + effect: reordering to put "None" effect first and seperate it from the "effect", to achieve a separation between "None" effect and effects
+  + phase_of_flt: ordering according from parking, over start of flight to end of flight
 
-struck.plot<-ggplot(birds[!is.na(birds$birds_struck),]) +
-  aes(y = birds_struck, x=speed, color=birds_struck, fill=birds_struck) +
-  scale_fill_brewer(palette="Blues") +
-  scale_color_brewer(palette="Blues") +
-  geom_boxplot(alpha=0.2)
+## 1.3 Handling missing values
 
-time.of.day.plot<-ggplot(birds[!is.na(birds$time_of_day),]) +
-  aes(y = time_of_day, x=speed, color=time_of_day, fill=time_of_day) +
-  scale_fill_manual(values = color_pal_TOD) +
-  scale_color_manual(values = color_pal_TOD) +
-  theme(legend.position = "none") +
-  geom_boxplot(alpha=0.2)
+birds_struck: We are using the "remarks" variable to check for spare parts from which we can determine that no wild life has been strike.
+By doing this we were able to decrease the missing values by 19 (from 39 to 20).
 
-effect.plot<-ggplot(birds_orig[!is.na(birds$effect),]) +
-  aes(x = effect, y=speed, color=effect, fill=effect) +
-  scale_fill_brewer(palette="OrRd") +
-  scale_color_brewer(palette="OrRd") +
-  theme(legend.position = "none") +
-  geom_boxplot(alpha=0.2)
+birds_seen: We assume that missing values in the "birds_seen" variable indicate that the pilot did not see wildlife before the strike occurred.
+Thus, we set all NA values in the "birds_seen" variable as "None".
+This leads to `r sum(is.na(birds$birds_seen))` values which are set as "None".
 
-dat.plot<-ggplot(birds[!is.na(birds$birds_struck) & birds$birds_struck != "0",]) +
-  aes(x = speed, y = height, color=birds_struck, fill=birds_struck) +
-  geom_point() +
-  facet_wrap(~birds_struck) +
-  scale_fill_brewer(palette="Blues") +
-  scale_color_brewer(palette="Blues") +
-  theme_classic() +
-  labs(
-    title = "Speed and Height of collisions",
-    subtitle = "In relation to the number of wildlife strucked each time",
-    x = "Speed in Knots",
-    y = "Height in Feet"
-  ) +
-  theme.title
+```{r echo=FALSE, message=FALSE, fig.asp=0.5}
+#get rows where birds struck is missing
+df <- birds[is.na(birds$birds_struck), 
+            c("remarks", "birds_seen", "effect", "birds_struck")]
+no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
+# set birds_struck to 0 for "no strike"
+birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"
 
-#(height.plot + speed.plot) / dat.plot
+#birds_seen: set NA as None
+birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
+#set Na values in birds seen to "None"
+birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
+
+# Amount of na for each column
+missing <- sapply(birds, function(x) sum(is.na(x)))
+missing_df <- data.frame("column" = names(missing), 
+                         "count" = missing, row.names = NULL)
+names(missing_df)[2] <- "count"
+
+ggplot(missing_df) + 
+  aes(x=reorder(column, -count), y=count) +
+  geom_bar(stat="identity", fill="snow3") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ```
+
 \newpage
 
-# Is there a Relationship between State and Wildlife?
+# 2. Aircraft models
+
+The dataset contains `r length(unique(birds$atype))` different aircraft models.
+We extracted the overall maximum speed and height for each aircraft to get an overview of the aircraft models and the relationship between speed and height.
+The following left plot shows the Top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the Top 50 aircrafts colored by their ac_mass.
+
+```{r echo=FALSE, message=FALSE, fig.asp=0.5}
+
+machines <- birds[!is.na(birds$height) & !is.na(birds$speed),] %>%
+  group_by(atype) %>%
+  summarise(max_speed = max(speed),
+            max_height = max(height),
+            num_engs = median(as.integer(num_engs)),
+            ac_mass = median(as.integer(ac_mass)))
+
+machines <- as.data.frame(machines)
+aircraft_strike_count <- birds %>% group_by(atype) %>% tally()
+machines<-merge(x = machines, y = aircraft_strike_count, by.x = "atype", by.y = "atype")
+
+machines$num_engs <- factor(machines$num_engs, levels=1:4)
+machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
+
+machines$ac_mass2 <- ifelse(as.integer(machines$ac_mass)<3,1,ifelse(as.integer(machines$ac_mass)==3,2,3))
+machines$ac_mass2 <- factor(machines$ac_mass2, levels=1:3)
+
+top10_aircrafts_bar <- machines %>%
+arrange(desc(n)) %>%
+head(10) %>%
+ggplot(aes(x=reorder(atype, n), y=n, fill = ac_mass))+
+  geom_bar(stat="identity") +
+  theme(legend.position="none") +
+  labs(title= "Top 10 frequent strikes by aircraft", x="aircraft model") +
+  coord_flip()
+
+set.seed(43)
+top50 <- machines[order(-machines$n)[1:50],] %>% 
+  mutate(angle = 30 * sample(-2:2, n(), replace = TRUE, prob = c(1, 1, 4, 1, 1)))
+top50_aircrafts_wordcloud <- ggplot(top50) +
+  aes(label = atype, size=n, color=ac_mass, angle=angle) +
+  geom_text_wordcloud_area(
+    mask = png::readPNG(source="flugzeug6-3.png"),
+    rm_outside = TRUE) +
+  scale_size_area(max_size = 3) +
+  scale_color_brewer(palette="BuPu") +#scale_color_continuous("Blues") +
+  #ggtitle("top 50 aircraft models causing strikes")
+  theme(plot.title = element_text(vjust=-20)) +
+  labs(title="Top 50 aircrafts causing strikes")
+
+aircraft_speed_height <- ggplot(top50[!is.na(top50$ac_mass),]) +
+
+#aircraft_speed_height <- ggplot(machines[!is.na(machines$ac_mass),]) +
+  aes(x=max_speed, y=max_height, color = ac_mass) +
+  geom_point() +
+  geom_smooth(method = lm, aes(x=max_speed, y = max_height, group = "black", fill = "black", ), se = FALSE) +
+  theme(legend.position="top") +
+  scale_color_brewer(palette="BuPu") +
+  guides(fill = FALSE)
+
+top50_aircrafts_wordcloud + aircraft_speed_height
+```
+
+The left plot shows that especially the B-737 class (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727.
+"B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas.
+The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes.
+Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this.
+But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.
+
+From the right plot we can see, that aircraft models with higher max speed are also able to fly higher.
+This can be also seen from the linear model, that has been estimated by geom_smooth.
+The correlation of max_speed and height for the Top50 models is: `r cor(top50$max_speed, top50$max_height)`, which is positive and quite high.
+In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height and that the lighter aircrafts under 5700 kg (ac_mass=2) or even 27000 kg (ac_mass=3) are less prominent.
+The group with ac_mass=4 (27001-272000 kg) is the most prominent group under the Top50 aircrafts.
+Two aircraft models with ac_mass = 5 are also present in the Top50 which are the only aircrafts with ac_mass=5.
+
+The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all aircraft models (n=`r nrow(machines)`.
+Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.
+
+```{r echo=FALSE, fig.asp=0.5}
+machines_c <- machines[complete.cases(machines),]
+p <- ggplot(machines[!is.na(machines$num_engs) & !is.na(machines$ac_mass),]) +
+  aes(x=max_speed, y=max_height, color = ac_mass) +
+  geom_point() +
+  scale_color_brewer(palette="BuPu") +
+  theme_bw()
+
+p + facet_grid(. ~ num_engs)
+```
+
+From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines.
+We can also see that lighter machines tend to have less engines first group includes nearly only machines with ac_mass=1, second group is mixed with ac_mass 2-4 and the last two groups include mostly aircrafts with ac_mass=4.
+Looking at the contingency table also confirms that:
+```{r echo=FALSE, fig.asp=0.5}
+table(machines$num_engs, machines$ac_mass)
+```
+Interesting is that for aircrafts with num_engs=1, there is no model with a speed > 200 and height > 10000.
+So we could conclude from that this is the maximum height and speed possible with one engine.
+
+\newpage
+
+# 3. Analysis of influences for birds_struck, birds_seen and effect
+In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft.
+We will first start with the analysis of strikes.
+
+## 3.1 Analysis of strikes (birds_struck)
+
+The following plot on the left shows the amount of birds struck out birds_seen.
+On the top right we can see the amount of collisions by each phase of flight and on the bottom right we can see the amount of collisions by time of day.
+For coloring we use the birds_struck variable.
+
+```{r echo=FALSE, fig.asp=0.5}
+birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck),]) +
+  aes(x = phase_of_flt, fill=birds_struck) +
+  scale_fill_brewer(palette="Blues") +
+  geom_bar() +
+  coord_flip() +
+  theme(legend.position="none")
+
+# Most collisions during the day #genauer betrachten
+birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
+  aes(x=time_of_day, fill=birds_struck) + 
+  scale_fill_brewer(palette="Blues") +
+  geom_bar() +
+  coord_flip() +
+  theme(legend.position="none")
+
+### Birds struck out of birds seen
+birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
+  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
+  scale_fill_brewer(palette="Blues") +
+  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank(), legend.position="top")
+
+birds_struck_seen + birds_strike_phase / birds_strike_time
+## descent and en-route are about equal 
+```
+
+birds_struck out of birds_seen: The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
+So most often we do not see a birds before a strike happens.
+But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones.
+Most often 1 bird is struck.
+Most interesting here is that if the pilot sees birds for the group birds seen = 2-10 , he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+The event of hitting less birds than seen will later be analyzed more in detail.
+
+The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens the Landing, Climb and Take-Off phase.
+
+From the bottom right plot we can see that strikes happen most often during the Day with over 10500 occurences followed by Night with about 4500 occurences.
+Interesting to see is that strikes with larger groups of birds (11-100)
+
+We will now have a look on the influence of the sky variable for the different amounts of birds struck.
+
+```{r echo=FALSE, fig.asp=0.5}
+#relationship of amount of birds struck with clouds?
+birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
+  aes(x=birds_struck, fill=sky)+
+  scale_fill_manual(values = color_pal_sky)+
+  geom_bar(position = "fill")
+
+birds_strike_ac_mass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$birds_struck) & (birds$birds_struck=="0" | as.integer(birds$birds_struck)>3),])+
+aes(x=ac_mass, fill=birds_struck)+
+  scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
+  geom_bar(position = "fill")
+
+#birds_strike_ac_mass <- ggplot(birds[!is.na(birds$birds_struck) & !is.na(birds$ac_mass),]) +
+#  aes(x=ac_mass, fill=birds_struck)+
+#  scale_fill_brewer(palette="Blues") +
+  #scale_fill_manual(values = color_pal_sky)+
+#  geom_bar(position = "fill")
+
+birds_strike_sky +birds_strike_ac_mass
+```
+
+From the plot we can see that with increasing number of birds_struck the proportion of overcast increases and the proportion of no clouds decreases.
+The proportion of some clouds is more or less the same across all groups.
+From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy.
+
+TODO: add Chi-squared independence test for sky and birds struck
+http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r
+
+We will now have a look on the height variable together with phase of flight.
+
+### 3.1.1 Closer look on the Phases of Flight
+
+The plot on the left shows a boxplot of height for flight phases, which are above ground and on the right side we will have a closer on the height for the flight phases "En Route" and "Descent".
+
+```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
+above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
+height_by_phase <- ggplot(above_ground) +
+  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
+  stat_boxplot(geom="errorbar", width=0.4) +
+  scale_fill_brewer(palette="Pastel2") +
+  geom_boxplot()
+# Heights during Descent are on average higher than 
+# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?
+
+#plot height density for Descent and EnRoute
+density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
+  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
+  scale_fill_manual(values=c("#fdcdac", "#cbd5e8")) +
+  geom_density(adjust=1.5,alpha=.8)
+
+height_by_phase + density
+
+```
+
+The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
+The other two boxplots for "En Route" and "Descent" are more far away from ground.
+Interesting here is that the 25% quantile of the Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
+
+We are performing a t.test to compare the two groups and check for equality.
+
+t.test
+```{r echo=FALSE, fig.asp=1}
+#Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
+t.test(height~phase_of_flt, data = birds[as.integer(birds$phase_of_flt)>4 & as.integer(birds$phase_of_flt)<7 & !is.na(birds$phase_of_flt),])
+```
+
+The T-Test also show that the group means are different from each other.
+But surprisingly the mean in Group "Descent" is higher than mean of "En Route".
+One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach.
+So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route".
+The main difference between these flight phases is that the aircraft is not flying parralel to the ground when being in phase "Descent".
+
+### 3.1.2 density birdsstuck by height for flight phases above ground (Climb, En Route, Descent, Approach)
+```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
+#could include only birds from vys categorization
+#plot height density for Descent and EnRoute
+density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
+  aes(x = height, group=birds_struck,fill=birds_struck)+
+  geom_density(adjust=1.5,alpha=.6) +
+  scale_fill_brewer(palette="Blues") +
+  theme(legend.position = "top") + 
+  xlim(0, 3500)
+density
+```
+
+Peek for "Over 100" birds struck at a height of about 1800 foot is due to the small amount of observations from that category. But nevertheless 8 out of 17 strikes happened here.
+
+## 3.2 Analysis of birds seen
+
+there are only 2 levels in the data available for birds seen
+
+```{r echo=FALSE, fig.asp=0.5}
+#relationship of amount of birds seen with clouds?
+birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
+  aes(x=time_of_day, fill=birds_seen)+
+  geom_bar(position = "fill") +
+  scale_fill_brewer(palette="BuGn") +
+  theme(legend.position="none")
+
+birds_seenNo_sky <- ggplot(birds[!is.na(birds$sky),])+
+  aes(x=sky, fill=birds_seen)+
+  scale_fill_brewer(palette="BuGn") +
+  geom_bar(position="fill")
+
+#relationship of seen or not seen with daytime?
+#table(is.na(birds$birds_seen), birds$time_of_day)
+
+birds_seen_plt + birds_seenNo_sky
+```
+
+**Time of Day:** Based on the assumption that Missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is `r sum(birds$birds_seen!="None")/nrow(birds)`.
+The contingencytable looks like the following:
+
+```{r echo=FALSE, fig.asp=0.5}
+addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean) 
+```
+
+The proportion of overall birds seen at night is much lower compared to the other times (Dawn, Day, Dusk).
+And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).
+
+**Sky:** the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
+contradiction: would have expected to see less birds when amount of clouds increases. but especially the amount of birds in groups from 11-100 increases when the amount of clouds increases.
+
+```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
+par(mfrow=c(1,2),las = 2, cex.axis = 0.75, mar = c(7, 4.1, 4.1, 2.1))
+spineplot(birds$height, birds$birds_seen, ylab = "birds_seen", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
+spineplot(birds$speed, birds$birds_seen, ylab = "birds_seen", xlab = "speed",col = rev(brewer.pal(n=3,name="BuGn")))
+```
+
+The higher the speed & height, the less likely to see a bird before collision
+for height > 2000 foot probability to see a birds before collision decreases 
+for speed > 140 knots probability to see a bird before collision decreases
+
+## 3.3 effect when birdsstruck < birds seen
+
+```{r echo=FALSE, fig.asp=0.5}
+#Create new column
+#calculate only when seen is not na
+seen_vs_strike <- birds[birds$birds_seen!="None" & !is.na(birds$birds_struck),]
+seen_vs_strike$birds_struck <- factor(seen_vs_strike$birds_struck)
+seen_vs_strike$birds_seen <- factor(seen_vs_strike$birds_seen)
+seen_vs_strike$less_strike_than_seen <- as.factor(as.integer(seen_vs_strike$birds_struck) < as.integer(seen_vs_strike$birds_seen)+2)
+
+#TODO: Why is barplot different from ggplot???
+#barplot(prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2), xlab = #"less_strike_than_seen")
+#prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2)
+
+# phase of flight by birdsstruck < birds seen
+seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt),]) +
+  aes(x=phase_of_flt, fill=less_strike_than_seen) +
+  scale_fill_manual(values = color_pal_TF)+
+  geom_bar(position = "fill") +
+  coord_flip() +
+  theme(legend.position="top")
+
+#ac_mass by leess-strike-than-seen (MOSAIC)
+seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
+  geom_mosaic(aes(x=product(ac_mass), fill=less_strike_than_seen)) +
+  scale_fill_manual(values = color_pal_TF)+
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())+
+  ylab("")
+
+#num_engs by leess-strike-than-seen (MOSAIC)
+seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
+  geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
+  scale_fill_manual(values = color_pal_TF)+
+  theme(legend.position="none", axis.text.y=element_blank(),axis.ticks.y = element_blank())+
+  ylab("")
+
+seen_vs_strike_phase + seen_vs_strike_ac_mass / seen_vs_strike_num_engs
+```
+
+the num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
+lighter machines and machines with less sngines tend to hit less wildlife out of the seen wildlife. This might be to better maneuvarability of the smaller aircrafts.
+
+## 3.4 Analysis of effect variable
+
+```{r echo=FALSE, fig.asp=0.5}
+#comparison right? because we actually want to compare here the positive impact
+birds_struck_effect <- ggplot(birds[!is.na(birds$birds_struck) & !is.na(birds$effect),]) +
+  aes(x = birds_struck, fill = effect) +
+  scale_fill_brewer(palette="OrRd") +
+  geom_bar(position="fill")
+
+#mosaic in ggplot2
+#https://xang1234.github.io/mosaicdiagrams/
+
+#mosaic plot better than barplot visualizes also absolute differences
+birds_effect_acmass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$effect),]) +
+  geom_mosaic(aes(x = product(ac_mass), fill=effect)) +
+  scale_fill_brewer(palette="OrRd") +
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+#birdds_seenYes by effect
+birds_effect_seenYes <- ggplot(birds[!is.na(birds$effect),]) +
+  geom_mosaic(aes(x = product(birds_seenYes), fill=effect))+
+  scale_fill_brewer(palette="OrRd") +
+  #theme(legend.title = element_blank())
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+
+birds_struck_effect / (birds_effect_acmass + birds_effect_seenYes)
+
+```
+
+check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured
+
+proportion of effect is a little higher when birds were seen => contradiction, we would expect to have less amount of effect when pilot is able to see birds before, cause he could
+
+\newpage
+
+# 4. Is there a Relationship between State and Wildlife?
 
@@
 
-## Species by State
+## 4.1 Species by State
 
@@
 We can see that the most frequent species that appear in the data are birds which could not be identified further.
-Apart from those we see that most bird strikes involve birds with big populations, particularly geese and gulls in the United States.
+Apart from those we can verify the fact that most bird strikes involve birds with big populations, particularly geese and gulls in the United States [@wiki].
 The third most frequent animal, for example, are gulls.
@@
 
-## Collisions per State
+## 4.2 Collisions per State
 
@@
   guides(fill=guide_colorbar(title="No. of strucks")) + 
-  #theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
-  #      axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
   labs(x = "Longitude", y = "Latitude") +
@@
 
-## Bird collisions vs. non-bird collisions
+## 4.3 Bird collisions vs. non-bird collisions
 
@@
 bird_wordcloud<-ggplot(is.bird.species[order(is.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
-  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="bird.png")) +
+  geom_text_wordcloud_area(area_corr_power = 1) +
   labs(title="Frequent bird species") +
@@
       vjust = -30,
-      hjust = 0.5, # 0.5 for center
+      hjust = 0.5,
       size = 10,
@@
 non_bird_wordcloud<-ggplot(non.bird.species[order(non.bird.species$Freq, decreasing = TRUE),][1:30,], aes(label = Var1, size=Freq)) +
-  geom_text_wordcloud_area(area_corr_power = 1) + #, mask = png::readPNG(source="deer.png")) +
+  geom_text_wordcloud_area(area_corr_power = 1) + 
   labs(title="Frequent non-bird species") +
@@
   labs(x = "Longitude", y = "Latitude") +
-  #ggtitle('Bird strucks per State') + 
   theme.title +
@@
   labs(x = "Longitude", y = "") +
-  #ggtitle('Non-bird strucks per State') + 
   theme.title +
@@
 (map.bird.prop + map.non.bird.prop) / (map.bird + map.non.bird)
-
-########### Ratio of nonbird/bird collisions doesn't match
-state.coll.nonbird<-select(state.coll.bird.nonbird, state, nonbird)
-state.coll.nonbird$is.bird<-FALSE
-colnames(state.coll.nonbird)[2]<-"Freq"
-state.coll.bird<-select(state.coll.bird.nonbird, state, bird)
-state.coll.bird$is.bird<-TRUE
-colnames(state.coll.bird)[2]<-"Freq"
-state.coll.bird.nonbird<-rbind(state.coll.nonbird, state.coll.bird)
-
-nbp<-ggplot(state.coll.bird.nonbird, aes(map_id = state)) +
-  geom_map(aes(fill = Freq), map = us.state) +
-  scale_fill_distiller(palette = "BrBG") +
-  guides(fill=guide_colorbar(title="No. of strucks")) + 
-  expand_limits(x = us.state$long, y = us.state$lat) +
-  facet_wrap( ~ is.bird)
-###########
 ```
@@
 
-## Vulture collisions per State
+## 4.4 Vulture collisions per State
 
@@
 
-# Timeline - Does time matter?
+# 5. Timeline - Does time matter?
 
-## The date variable
+## 5.1 The date variable
 
@@
 
-## Monthly and weekly grouping
+## 5.2 Monthly and weekly grouping
 
@@
   labs(x = "Month", y = "Number of collisions") +
-  #theme(axis.title.y=element_blank()) +
   guides(fill=FALSE) +
@@
   geom_bar()
-# -> more collisions during autumn/fall
 
@@
 # Or weekly:
-week_plot<-ggplot(birds.filt) + #[as.integer(birds.filt$birds_struck) > 3,]) +
+week_plot<-ggplot(birds.filt) + 
   aes(x = day, group=birds_struck, fill=birds_struck) +
@@
   geom_bar(position="fill")
-# -> bird collisions not dependent on weekdays
 
@@
   geom_bar()
-# way more collisions in second half of the year for birds -> mehr Zugvögel?
-# But they would have to come back in spring which we can't see in the diagram?
 
@@
 xlabels <- c("","02","","04","","06","","08","","10","","12")
-#xlabels[seq(1, length(xlabels), 2)] <- ""
 
@@
   scale_x_discrete(labels=xlabels)
-# -> less overcast during summer months
 
@@
   scale_x_discrete(labels=xlabels)
-# -> possible explanation: It gets dark later in the summer (in the US)
-# -> less strucks during the night in the summer
 ```
@@
 (month_plot / woy_plot) / (week_plot + half.year.plot)
-
-#grid.arrange(month_plot, woy_plot, month_plot_filt, top = "Hawk collisions dover the Year", nrow=3) + theme.title
 ```
@@
 
-## Grouping per Year
+## 5.3 Grouping per Year
 
@@
 
-coll.year.bar<-ggplot(birds.filtered.struck.num, aes(x = year, group = year)) + geom_bar()
-
-coll.year / coll.year.month #/ coll.year.bar
+coll.year / coll.year.month
 ```
@@
 
-## Hawk collisions over Time and by State
+## 5.4 Hawk collisions over Time and by State
 
@@
   geom_polygon(color = "white") + 
-  #scale_fill_continuous(low="grey", high="red") +
   scale_fill_distiller(palette = "Reds", trans="reverse") +
@@
        title = "Hawk strikes in Summer") +
-  #ggtitle('Hawk collisions during Summer') + 
   theme.title +
@@
   geom_polygon(color = "white") + 
-  #scale_fill_continuous(low="grey", high="pink") +
   scale_fill_distiller(palette = "Reds", trans="reverse") +
@@
   theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
-  #ggtitle('Hawk collisions during Winter') + 
   theme.title +
@@
 
-#library(gridExtra)
-#grid.arrange(plot_summer, plot_winter, top = paste("Hawk collisions depending on the season", "In relation to total hawk collisions", sep="\n"), nrow=1) + theme.title
 plot_summer + plot_winter
@@
 
+\newpage
+
+# Summary
+
+Hier fehlt noch die summary.
+
+# Outlook
+
+* bird/non-bird: another dataset to have definite categorization
+* Datensatz mit Anzahl Flügen allgemein für proportionale Angaben
+* Start/Destination des Flugs für Richtung usw.
+* Vergleich mit anderen Ländern
+* Closer look at missing values
+
+\newpage
+
 # References

### 2021-07-19 draft_final.Rmd
Added plot titles and description as well as conlcusions from plots.

@@
 library(ggwordcloud)
+library(ggpubr)
+library(grid)
+library(gridExtra)
 
@@
 
-birds$ac_mass <- as.factor(birds$ac_mass)
+factor_ac_mass <- c("<=2250 kg", "2251-5700 kg", "5701-27000 kg", "27001-272000 kg", ">272000 kg")
+birds$ac_mass <- as.factor(birds$ac_mass)#, levels=factor_ac_mass)
+birds$ac_mass_names <- revalue(as.factor(birds$ac_mass),     c("1"="<2250",        "2"="2251-5700",        "3"="5701-27000",        "4"="27001-272000",       "5"=">272000"))
 birds$num_engs <- as.factor(birds$num_engs)
@@
 
-```{r echo=FALSE, message=FALSE, fig.asp=0.5}
+```{r echo=FALSE, warning=FALSE,message = FALSE,fig.align = "center", fig.asp=0.5}
 
@@
             num_engs = median(as.integer(num_engs)),
-            ac_mass = median(as.integer(ac_mass)))
+            ac_mass = median(as.integer(ac_mass)),
+            ac_mass_names = median(as.integer(ac_mass_names)))
 
@@
 machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
+machines$ac_mass_names <- factor(machines$ac_mass_names)
 
@@
   geom_text_wordcloud_area(
-    mask = png::readPNG(source="flugzeug6-3.png"),
+    mask = png::readPNG(source="C:/Users/Alex/Documents/flugzeug6-3.png"),
     rm_outside = TRUE) +
@@
   #ggtitle("top 50 aircraft models causing strikes")
-  theme(plot.title = element_text(vjust=-20)) +
-  labs(title="Top 50 aircrafts causing strikes")
+  #theme(plot.title = element_text(vjust=-20)) +
+  labs(
+    title="Top 50 aircrafts involved in collisions",
+    subtitle = "Model types by number of collisions"
+       )
 
 aircraft_speed_height <- ggplot(top50[!is.na(top50$ac_mass),]) +
-
 #aircraft_speed_height <- ggplot(machines[!is.na(machines$ac_mass),]) +
@@
   geom_smooth(method = lm, aes(x=max_speed, y = max_height, group = "black", fill = "black", ), se = FALSE) +
-  theme(legend.position="top") +
+  theme(legend.position="right") +
   scale_color_brewer(palette="BuPu") +
-  guides(fill = FALSE)
+  guides(fill = FALSE)+
+  labs(
+    subtitle = "max speed vs. max height"
+  )
 
@@
 This can be also seen from the linear model, that has been estimated by geom_smooth.
-The correlation of max_speed and height for the Top50 models is: `r cor(top50$max_speed, top50$max_height)`, which is positive and quite high.
-In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height and that the lighter aircrafts under 5700 kg (ac_mass=2) or even 27000 kg (ac_mass=3) are less prominent.
-The group with ac_mass=4 (27001-272000 kg) is the most prominent group under the Top50 aircrafts.
-Two aircraft models with ac_mass = 5 are also present in the Top50 which are the only aircrafts with ac_mass=5.
+The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high.
+In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height.
+We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent.
+The group with ac_mass 27001-272000 kg) is the most prominent group under the Top50 aircrafts.
+Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.
 
@@
   scale_color_brewer(palette="BuPu") +
-  theme_bw()
+  theme_bw() +
+  labs(
+    title = "All aircraft models",
+    subtitle = "max speed vs. max height by number of engines"
+  )
 
@@
 # 3. Analysis of influences for birds_struck, birds_seen and effect
-In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft.
+In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect).
 We will first start with the analysis of strikes.
@@
   coord_flip() +
-  theme(legend.position="none")
+  theme(legend.position="none", 
+        axis.title.y=element_blank(),
+        axis.title.x=element_blank(),
+        plot.subtitle = element_text(hjust =1)) +
+  labs(xlab = "phase of flight",
+       subtitle = "Number of collisions by phase of flight")
 
@@
   coord_flip() +
-  theme(legend.position="none")
+  theme(legend.position="none",
+        axis.title.y=element_blank(),
+        axis.title.x=element_blank(),
+        plot.subtitle = element_text(hjust =1)) +
+  labs(xlab = "",
+       subtitle = "Number of collisions by time of day     ")
 
@@
   scale_fill_brewer(palette="Blues") +
-  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank(), legend.position="top")
+  theme(axis.text.y=element_blank(), 
+        axis.title.y=element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.text.x = element_text(angle = 90),
+        legend.position="left")+
+  labs(title = "Birds struck overview",
+       subtitle = "bird struck out of birds seen")
 
@@
 
-birds_struck out of birds_seen: The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
+The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
 So most often we do not see a birds before a strike happens.
@@
 Most often 1 bird is struck.
-Most interesting here is that if the pilot sees birds for the group birds seen = 2-10 , he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+Interesting here is that if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
 The event of hitting less birds than seen will later be analyzed more in detail.
 
-The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens the Landing, Climb and Take-Off phase.
+The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens during Landing, Climb and Take-Off phase.
 
 From the bottom right plot we can see that strikes happen most often during the Day with over 10500 occurences followed by Night with about 4500 occurences.
-Interesting to see is that strikes with larger groups of birds (11-100)
+Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).
 
@@
   scale_fill_manual(values = color_pal_sky)+
-  geom_bar(position = "fill")
+  geom_bar(position = "fill")+
+  labs(title = "Birds struck comparisons",
+       subtitle = "Cloudiness by birds struck")
 
@@
 
-From the plot we can see that with increasing number of birds_struck the proportion of overcast increases and the proportion of no clouds decreases.
+From the plot on the left we can see that with **increasing number of birds_struck the proportion of overcast increases** and the proportion of no clouds decreases.
 The proportion of some clouds is more or less the same across all groups.
@@
 
-### 3.1.1 Closer look on the Phases of Flight
-
-The plot on the left shows a boxplot of height for flight phases, which are above ground and on the right side we will have a closer on the height for the flight phases "En Route" and "Descent".
+### 3.1.1 Closer look on height of collisions
+
+The top plot shows a boxplot of height for flight phases, which are above ground and on the bottom we will have a closer look on the height for the flight phases "En Route" and "Descent", because they show the most variability in height.
 
@@
   scale_fill_brewer(palette="Pastel2") +
-  geom_boxplot()
+  geom_boxplot() +
+  coord_flip() +
+  theme(axis.title.y = element_blank())+
+  labs(title = "Height distribution for flight phases above ground")+
+  ylim(0,20000)
 # Heights during Descent are on average higher than 
@@
   aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
-  scale_fill_manual(values=c("#fdcdac", "#cbd5e8")) +
-  geom_density(adjust=1.5,alpha=.8)
-
-height_by_phase + density
+  scale_fill_brewer(palette='Pastel2') +
+  geom_density(adjust=1.5,alpha=.8) +
+  xlim(0,20000)
+
+height_by_phase / density
 
@@
 The other two boxplots for "En Route" and "Descent" are more far away from ground.
-Interesting here is that the 25% quantile of the Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
+Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
 
@@
 
-t.test
 ```{r echo=FALSE, fig.asp=1}
@@
 The T-Test also show that the group means are different from each other.
-But surprisingly the mean in Group "Descent" is higher than mean of "En Route".
-One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach.
+But surprisingly the **mean in group "Descent" is higher** than mean of "En Route".
+One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach or during flight in general.
 So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route".
-The main difference between these flight phases is that the aircraft is not flying parralel to the ground when being in phase "Descent".
+The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent".
+So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**.
+And especially a negative angle increases the probability of hitting birds, based on the findings here and from 3.1.1 by comparing the number of collisions by phase of flight.
 
 ### 3.1.2 density birdsstuck by height for flight phases above ground (Climb, En Route, Descent, Approach)
+
+We will now have another look on the density of height, but this time split by the different numbers of birds involved in the collision. The following plot shows exactly this.
+
 ```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
@@
   scale_fill_brewer(palette="Blues") +
-  theme(legend.position = "top") + 
-  xlim(0, 3500)
+  theme(legend.position = "right") + 
+  xlim(0, 3500) +
+  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")
 density
 ```
-
-Peek for "Over 100" birds struck at a height of about 1800 foot is due to the small amount of observations from that category. But nevertheless 8 out of 17 strikes happened here.
+The density plot shows a **high peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
+This is, because most of the collisions happen near ground level at flight phases Take-Off, Climb, Approach and Landing Roll (see 3.1 Number of collisions by pflight phase).
+For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 150 foot to 500 foot.
+Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**.
+For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here.
+Unfortunately we have mostly missing values in the variable species for these observations, so we can not get any information about the birds species.
 
@@
 
-there are only 2 levels in the data available for birds seen
+there are only two levels in the data available for birds seen and we added one level for the missing values.
+The two plots show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot).
 
@@
   scale_fill_brewer(palette="BuGn") +
-  theme(legend.position="none")
+  theme(legend.position="none", axis.title.y = element_blank()) +
+  labs(title="Variables influencing birds seen",
+       subtitle = "Proportion of birds seen by time of day")
 
@@
   scale_fill_brewer(palette="BuGn") +
-  geom_bar(position="fill")
+  geom_bar(position="fill") +
+   theme(axis.title.y = element_blank()) +
+  labs(subtitle = "Proportion of birds seen by sky")
 
@@
 
-**Time of Day:** Based on the assumption that Missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is `r sum(birds$birds_seen!="None")/nrow(birds)`.
+Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is 0.217+0.069=0.286.
 The contingencytable looks like the following:
@@
 ```{r echo=FALSE, fig.asp=0.5}
-addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean) 
-```
-
-The proportion of overall birds seen at night is much lower compared to the other times (Dawn, Day, Dusk).
+round(addmargins(prop.table(table(birds$birds_seen, birds$time_of_day),2),2,FUN = mean),3) 
+```
+
+The proportion of overall **birds seen at night is much lower** compared to the other times (Dawn, Day, Dusk).
 And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).
 
-**Sky:** the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
-contradiction: would have expected to see less birds when amount of clouds increases. but especially the amount of birds in groups from 11-100 increases when the amount of clouds increases.
+**Sky:** the proportion of **birds seen increases with cloudiness**.
+This is a **contradiction** to what we would have expected.
+We would have thought that with increasing cloudiness the amount of birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds.
+So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky.
+Another reason could be that birds in groups fly more often when its cloudy.
+
+The following two spineplots show the two continious variables height and speed by the number of birds seen.
 
 ```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
-par(mfrow=c(1,2),las = 2, cex.axis = 0.75, mar = c(7, 4.1, 4.1, 2.1))
-spineplot(birds$height, birds$birds_seen, ylab = "birds_seen", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
-spineplot(birds$speed, birds$birds_seen, ylab = "birds_seen", xlab = "speed",col = rev(brewer.pal(n=3,name="BuGn")))
-```
-
-The higher the speed & height, the less likely to see a bird before collision
-for height > 2000 foot probability to see a birds before collision decreases 
-for speed > 140 knots probability to see a bird before collision decreases
-
-## 3.3 effect when birdsstruck < birds seen
+par(mfrow=c(1,3))#,las = 2, cex.axis = 0.75, mar = c(7, 4.1, 4.1, 2.1))
+spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")), main = "Birds seen by height")
+spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "speed",
+          col = rev(brewer.pal(n=3,name="BuGn")), main = "Birds seen by speed")
+
+# Extract the legend. Returns a gtable
+leg <- get_legend(birds_seenNo_sky)
+# Convert to a ggplot and print
+vp.BottomRight <- viewport(y=0.5, x=.8)
+print(as_ggplot(leg), vp=vp.BottomRight)
+```
+
+The higher the speed & height, the less likely to see a bird before collision: 
+* for height > 2000 foot, the probability to see a birds before collision decreases 
+* for speed > 140 knots, the probability to see a bird before collision decreases
+
+## 3.3 Influences when birdsstruck < birds seen
+
+In the following we will have a closer look on a self-generated variable, which compares the amount of birds seen with the amount of birds_struck.
+It will be true if the number of birds struck is smaller than birds seen and false if not.
+In the beginning we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck.
+By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.
+
+The plot on the left shows the proportion of less strikes than seen by phase of flight.
+On the top right we can see a comparison of the proportion of less strikes by the aircraft mass and on the bottom right we compare it by different number of engines.
 
@@
   coord_flip() +
-  theme(legend.position="top")
+  theme(legend.position="right", axis.title.y = element_blank())+
+  labs(title="Variables influencing less strikes than seen",
+       subtitle = "Proportion of less strike than seen \nby phase of flight",
+       fill='less strike \nthan seen'
+       )
 
@@
 ```
-
-the num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
-lighter machines and machines with less sngines tend to hit less wildlife out of the seen wildlife. This might be to better maneuvarability of the smaller aircrafts.
+Comparing less strike than seen across phases of flight we can see, that "En Route" it is about 12% more likely to strike less birds than seen compared to "Descent".
+So it is not even less likely to strike birds in general, but also to strike less birds than seen.
+The num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
+lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife.
+This might be due to better maneuvarability of the lighter aircrafts.
+Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines.
 
 ## 3.4 Analysis of effect variable
+
+The top plot shows the different proportions of effect for each category of birds_struck.
+The Bottom left plot shows the proportions of each effect by the aircraft mass and on the bottom right we compare effect with birds_seen or not seen.
 
@@
   scale_fill_brewer(palette="OrRd") +
-  geom_bar(position="fill")
+  geom_bar(position="fill") +
+  labs(
+    title = "Variables influencing effect",
+    subtitle = "Effect by number of birds struck"
+  )
 
@@
   scale_fill_brewer(palette="OrRd") +
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  theme(legend.position="none", 
+        axis.text.y=element_blank(), 
+        axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()) +
+  labs(subtitle = "Effect by aircraft mass")
 
@@
   #theme(legend.title = element_blank())
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  theme(legend.position="none", 
+        axis.text.y=element_blank(), 
+        axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()) +
+  labs(xlab= "birds seen?", subtitle = "Effect by birds seen or not")
 
@@
 
-check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured
-
-proportion of effect is a little higher when birds were seen => contradiction, we would expect to have less amount of effect when pilot is able to see birds before, cause he could
+birds_struck: The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
+Especially the proportion of "Precautionary Landing" increases.
+For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off".
+
+ac_mass: the higher the aircraft mass the less likely the collision has an effect on the aircraft.
+
+birds_seenYes: The "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and Precautionary Landing and Aborted Take-off are a little higher.
+We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike.
+But from the description of the dataset we can not really find out what is meant by the effect variable.
+Does "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision.
+When birds were seen we would expect to have less amount of effect on the aircraft, cause he could try to dodge the birds or abort take off/ land precautionary.
 

### 2021-07-19 draft_final.Rmd
PDF with aircraft models wordcloud (without shape)

@@
 output:
-  html_document:
+  pdf_document:
     toc: true
     toc_depth: 3
-  pdf_document:
+  html_document:
     toc: true
@@
   geom_text_wordcloud_area(
-    mask = png::readPNG(source="C:/Users/Alex/Documents/flugzeug6-3.png"),
+    #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
     rm_outside = TRUE) +

### 2021-07-19 draft_final.Rmd
minor plot changes.

--- 2021-07-19 draft_final.Rmd
@@
 }
+
+#set working dir to where the current file is located
+dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
+#setwd(dir)
 
@@
   geom_text_wordcloud_area(
-    mask = png::readPNG(source="C:/Users/Alex/Documents/flugzeug6-3.png"),
+    mask = png::readPNG(source="flugzeug6-3.png"),
     rm_outside = TRUE) +
@@
   geom_bar(position = "fill")+
+  theme(axis.text.x = element_text(angle = 90)) +
   labs(title = "Birds struck comparisons",
@@
   scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
-  geom_bar(position = "fill")
+  geom_bar(position = "fill") + 
+  labs(subtitle = "Birds struck subset by aircraft mass")
 
@@
   scale_fill_brewer(palette="Pastel2") +
-  geom_boxplot() +
+  geom_boxplot(outlier.size = .9) +
   coord_flip() +
@@
   labs(title="Variables influencing birds seen",
-       subtitle = "Proportion of birds seen by time of day")
+       subtitle = "Proportion of birds seen by time")
 

### 2021-07-19 draft_final.Rmd

@@
     mask = png::readPNG(source="flugzeug6-3.png"),
+    #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
     rm_outside = TRUE) +

### 2021-07-19 draft_final.Rmd
Plot titles and labels, restructuring.

--- 2021-07-19 draft_final.Rmd
@@
   geom_bar(stat="identity", fill="snow3") +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+        axis.title.x = element_blank())
 ```
@@
   #theme(plot.title = element_text(vjust=-20)) +
+  theme.title+
   labs(
@@
   guides(fill = FALSE)+
+  theme.title+
   labs(
@@
   scale_color_brewer(palette="BuPu") +
-  theme_bw() +
+  theme(plot.title = element_text(
+      hjust = 0, 
+      size = 12,
+      color = "#016c59",
+      face = "bold"
+    ),
+    plot.subtitle = element_text(
+      hjust = 0, 
+      size = 10,
+      color = "gray50",
+      face = "italic"
+    ),
+    strip.background = element_rect(colour = "black", fill = "white"),
+    panel.background = element_rect(colour = "grey50")) +
   labs(
@@
 Looking at the contingency table also confirms that:
+
 ```{r echo=FALSE, fig.asp=0.5}
-table(machines$num_engs, machines$ac_mass)
+# create some fake data (2x2, since we're building a confusion matrix) 
+dat <- matrix(table(machines$num_engs, machines$ac_mass), nrow = 4, dimnames=list(c(1:4), c(1:5)))
+# now set the names *of the dimensions* (not the row/colnames)
+names(dimnames(dat)) <- c("num_engs", "ac_mass")
+dat
 ```
@@
         plot.subtitle = element_text(hjust =1)) +
+  theme.title+
   labs(xlab = "phase of flight",
@@
         plot.subtitle = element_text(hjust =1)) +
+  theme.title+
   labs(xlab = "",
@@
         legend.position="left")+
+  theme.title+
   labs(title = "Birds struck overview",
@@
 The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
-So most often we do not see a birds before a strike happens.
+So **most often we do not see a birds before a strike** happens and most often 1 bird is struck.
 But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones.
-Most often 1 bird is struck.
-Interesting here is that if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+If the pilot sees birds, **quite often less birds were hit than seen**.
+Interesting here is, that if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
 The event of hitting less birds than seen will later be analyzed more in detail.
@@
 #relationship of amount of birds struck with clouds?
+struck_sky_sums <- rep(addmargins(table(birds$sky, birds$birds_struck),1)[4,],each=3)
+
 birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
@@
   geom_bar(position = "fill")+
+  geom_text(aes(label=ifelse(round(..count../struck_sky_sums,2) > 0.1,round(..count../struck_sky_sums,2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
   theme(axis.text.x = element_text(angle = 90)) +
+  theme.title+
   labs(title = "Birds struck comparisons",
@@
 
-birds_strike_ac_mass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$birds_struck) & (birds$birds_struck=="0" | as.integer(birds$birds_struck)>3),])+
+ac_struck_sums <- rep(addmargins(table(birds$birds_struck, birds$ac_mass)[c(1,4,5),],1)[4,],each=3)
+
+birds_struck_minorities <- birds[!is.na(birds$ac_mass) & !is.na(birds$birds_struck) & (birds$birds_struck=="0" | as.integer(birds$birds_struck)>3),]
+birds_struck_minorities$birds_struck <- droplevels(birds_struck_minorities$birds_struck)
+
+birds_strike_ac_mass <- ggplot(birds_struck_minorities)+
 aes(x=ac_mass, fill=birds_struck)+
@@
   geom_bar(position = "fill") +
-  labs(subtitle = "Birds struck subset by aircraft mass")
+  geom_text(aes(label=ifelse(round(..count../ac_struck_sums[1:14],2) > 0.1,round(..count../ac_struck_sums[1:14],2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
+  theme.title+
+  labs(subtitle = "Birds struck (subset) by aircraft mass")
 
@@
 The top plot shows a boxplot of height for flight phases, which are above ground and on the bottom we will have a closer look on the height for the flight phases "En Route" and "Descent", because they show the most variability in height.
+We defined Climb, En Route, Descent, Approach as phases which are above ground level.
 
@@
   theme(axis.title.y = element_blank())+
+  theme.title+
   labs(title = "Height distribution for flight phases above ground")+
@@
 
-### 3.1.2 density birdsstuck by height for flight phases above ground (Climb, En Route, Descent, Approach)
+### 3.1.2 density birdsstuck by height for flight phases above ground
 
@@
   theme(legend.position = "right") + 
+  theme.title+
   xlim(0, 3500) +
@@
 ```
+
 The density plot shows a **high peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
@@
 there are only two levels in the data available for birds seen and we added one level for the missing values.
-The two plots show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot).
-
-```{r echo=FALSE, fig.asp=0.5}
+The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot).
+At the bottom you can see two spineplots that show the two continuous variables height and speed by the number of birds seen.
+
+```{r echo=FALSE, fig.asp=0.4}
 #relationship of amount of birds seen with clouds?
+
+time_sums <- rep(addmargins(table(birds$birds_seen, birds$time_of_day),1)[4,],each=3)
 birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
@@
   geom_bar(position = "fill") +
+  geom_text(aes(label=ifelse(round(..count../time_sums,2) > 0.1,round(..count../time_sums,2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
   scale_fill_brewer(palette="BuGn") +
   theme(legend.position="none", axis.title.y = element_blank()) +
+  theme.title+
   labs(title="Variables influencing birds seen",
        subtitle = "Proportion of birds seen by time")
+
+sky_sums <- rep(addmargins(table(birds$birds_seen, birds$sky),1)[4,],each=3)
 
@@
   geom_bar(position="fill") +
+  geom_text(aes(label=ifelse(round(..count../sky_sums,2) > 0.1,round(..count../sky_sums,2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
   theme(axis.title.y = element_blank()) +
+  theme.title+
   labs(subtitle = "Proportion of birds seen by sky")
@@
 ```
+```{r echo=FALSE, fig.asp=0.4, fig.width = 7, fig.show="hold", fig.align = 'left'}
+par(mfrow=c(1,2))
+spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
+title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50")
+
+spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "speed",
+          col = rev(brewer.pal(n=3,name="BuGn")))
+title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50")
+```
 
@@
 
-**Sky:** the proportion of **birds seen increases with cloudiness**.
+Sky: The proportion of **birds seen increases with cloudiness**.
 This is a **contradiction** to what we would have expected.
@@
 Another reason could be that birds in groups fly more often when its cloudy.
-
-The following two spineplots show the two continious variables height and speed by the number of birds seen.
-
-```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
-par(mfrow=c(1,3))#,las = 2, cex.axis = 0.75, mar = c(7, 4.1, 4.1, 2.1))
-spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")), main = "Birds seen by height")
-spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "speed",
-          col = rev(brewer.pal(n=3,name="BuGn")), main = "Birds seen by speed")
-
-# Extract the legend. Returns a gtable
-leg <- get_legend(birds_seenNo_sky)
-# Convert to a ggplot and print
-vp.BottomRight <- viewport(y=0.5, x=.8)
-print(as_ggplot(leg), vp=vp.BottomRight)
-```
 
@@
 
+less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,],each=2)
+
 # phase of flight by birdsstruck < birds seen
@@
   aes(x=phase_of_flt, fill=less_strike_than_seen) +
-  scale_fill_manual(values = color_pal_TF)+
+  scale_fill_manual(values = color_pal_TF) +
   geom_bar(position = "fill") +
+ # geom_text(aes(label=..count..),
+  geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) > 0.1,round(..count../less_strike_sums[2:16],2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
   coord_flip() +
   theme(legend.position="right", axis.title.y = element_blank())+
+  theme.title+
   labs(title="Variables influencing less strikes than seen",
@@
 ```
+
 Comparing less strike than seen across phases of flight we can see, that "En Route" it is about 12% more likely to strike less birds than seen compared to "Descent".
@@
 ```{r echo=FALSE, fig.asp=0.5}
+
+effect_strike_sums <- rep(addmargins(table(birds$effect, birds$birds_struck),1)[6,],each=5)
+
 #comparison right? because we actually want to compare here the positive impact
@@
   geom_bar(position="fill") +
+  geom_text(aes(label=ifelse(round(..count../effect_strike_sums[3:25],2) > 0.4,round(..count../effect_strike_sums[3:25],2) , "")),
+            stat='count',
+            size = 3,
+            position=position_fill(vjust = 0.5))+
+  theme.title+
   labs(
@@
         axis.title.y = element_blank()) +
+  theme.title+
   labs(subtitle = "Effect by aircraft mass")
@@
         axis.title.y = element_blank()) +
+  theme.title+
   labs(xlab= "birds seen?", subtitle = "Effect by birds seen or not")

### 2021-07-19 draft_final.Rmd
Added Outlook, Summary, Preparation part & Sources

@@
 
-## 1.2 Data preparation
-
+## 1.2 Preparation
+
+#### Data Preparation: 
 Cleaning birds_seen variable: The 9412. observation had one wrong entry in the birds seen variable which has been replaced with NA, because we did not want to introduce a new level for one entry.
@@
   + phase_of_flt: ordering according from parking, over start of flight to end of flight
+
+#### Preparation for Visualization: 
+Since we wanted to visualize our findings and results in a simple and understandable manner, we were very considerate with the colors and type of plots we used.
+We looked at different color palettes [@datanovia; @cheatsheet; @colorbrewer] and made sure that colors were distinguishable when comparing different classes, or sequential when in-/decrease was important.
+We also tried to pick colors that matched with what we wanted to display (e.g.
+yellow for daylight/no clouds etc.).
+For the states we found that using a map for visualization worked very well, for timelines a lineplot with points was the easiest to interpret, we used scatter, bar-, density-, spine- and mosaicplots for comparing amounts and proportions/ratios and boxplots to compare different levels of a factor variable with each other.
+And for certain questions wordclouds [@wordcloud] worked the best.
+We then made sure that axis labels, plot titles and legends were concise and clear.
 
@@
   geom_text_wordcloud_area(
-    mask = png::readPNG(source="flugzeug6-3.png"),
+    #mask = png::readPNG(source="flugzeug6-3.png"),
     #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
@@
 The figure shows the number of animals of a specific type which has been displayed in a sorted barplot.
-We can see that the most frequent species that appear in the data are birds which could not be identified further.
+We can see that the most frequent species that appear in the data are birds which could not be identified further (e.g.
+over 4000 collisions for small unknown birds).
 Apart from those we can verify the fact that most bird strikes involve birds with big populations, particularly geese and gulls in the United States [@wiki].
-The third most frequent animal, for example, are gulls.
+The third most frequent animal, for example, are gulls with over 2500 collisions.
 
@@
 
-When we compare that with the most frequent species in West Virginia (WV), the most frequent species that aircrafts collided with there is surprisingly not a bird but white-tailed deers.
+When we compare that with the most frequent species in West Virginia (WV), the most frequent species that aircrafts have collided there with is surprisingly not a bird but the white-tailed deer (Over 30 collisions).
 And foxes are also among the list.
@@
   geom_bar(stat="identity", fill="darkseagreen") +
-  labs(subtitle="Top 10 states with most number of gulls collisions",
+  labs(subtitle="Top 5 states with most number of gulls coll.",
        x = "Number of Collisions",
@@
 
-When we look at the top states we see that those are all mainly US states that lie on the coastline, for example California, New York and Florida.
-This therefore makes sense that collisions with gulls happen there a lot since those birds are often found near where the sea is.
+When we look at the top states we see that those are all mainly US states that lie on the coastline, for example California, New York and Florida with between 200 to 300 gulls collisions each.
+This therefore makes sense that collisions with gulls happen there a lot since those birds are often found near where the sea is [@gulls].
 
@@
 We also further examined the **number of collisions per state** which is displayed below.  
-For this, we wanted to display our findings in a map for better visualization.
+For this, we wanted to display our findings in a map for better visualization [@maps; @polymap].
 In the following we have displayed the number of collisions, so the variable *birds_struck*, depending on the state.
@@
 
-The states with the most collisions seem to be California, Florida, New York and Texas.
+The states with the most collisions seem to be California, Florida, New York and Texas (each approx.
+around 4000 strucks).
 This is reasonable since those states are the most known ones and there is a lot of air traffic going on: People travel a lot to and from those states so it makes sense that there are the most collisions.
@@
   theme(legend.position="right") +
-  scale_y_discrete(breaks=every_nth(5))
+  scale_y_discrete(breaks=every_nth(3))
 
@@
 This could be due to the fact that those states are frequently flown to and from, so bigger aircrafts like big passenger planes are used there more than in other states.
-Another reason could be that there's a big aerospace industry in Washington and the Boeing Everett factory.
-It's said to be the best state for building boeings.
+Another reason could be that there's a big aerospace industry in Washington and the Boeing Everett factory [@washington].
+It's said to be the best state for building boeings [@boeings].
+This would also fit with our findings from earlier where the most frequent aircraft models in the heaviest *ac_mass* category were Boeings.
 
@@
 In the following we see wordclouds of the bird and non-bird species, the size of the words in regard to their respective frequency in the data.
-(Top 30 for each, the number of the words have been limited due to size and readability.)
+(The number of the species/words have been limited due to size and readability, only the most frequent ones are displayed.)
 
@@
       size = 10,
-      color = "darkcyan",
+      color = "#016c59",
       face = "bold"
@@
    theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
+  theme(plot.margin=unit(c(5.5,20,5.5,5.5),"points")) +
   coord_fixed(1.3)
@@
         axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
+  theme(plot.margin=unit(c(5.5,5.5,5.5,20),"points")) +
   coord_fixed(1.3)
@@
   theme(legend.position="top") +
+  theme(plot.margin=unit(c(5.5,20,5.5,5.5),"points")) +
   coord_fixed(1.3)
@@
   theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
+  theme(plot.margin=unit(c(5.5,5.5,5.5,20),"points")) +
   coord_fixed(1.3)
@@
 
-Vultures are more frequent in the South and warm regions which fits with our map above. Their habitat is not in the midwest and hence, there aren't any vulture collisions happening there.
+Vultures are more frequent in the South and warm regions [@vultures] which fits with our map above. Their habitat is not in the midwest and hence, there aren't any vulture collisions happening there.
 
@@
 So this is another indicator supporting the claim.
-Apart from this, our observation fit with general information that can be found with other sources: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.
+Apart from this, our observation fit with general information that can be found with other sources [@wiki]: Wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.
 
@@
 And overall there seem to be less hawk strikes during the winter.
-This again can be explained with what was mentioned earlier: Hawks are migratory birds as well.
+This again can be explained with what was mentioned earlier: Hawks are migratory birds as well [@hawks].
 During the winter months they travel to the southern regions, hence why there are almost no hawk strikes in the North during winter.
@@
 
-Hier fehlt noch die summary.
+We can see that we can find out a lot about bird strikes and aircraft collisions based on this dataset.  
+We saw that states that people travel to and from the most also have the most collisions.
+And while birds make up most part of the encountered species, there is a small percentage of strikes that involve non-bird wildlife.
+Depending on the state the species differ.
+We were also able to show the habitat of certain wildlife in the U.S.
+and how this influences the number of collisions and strikes with aircrafts by visualizing this with location data.
+By taking the time component into account we could also detect that wildlife strikes increase with each year and within a year there are more collisions during the fall/bird migration months.
+By combining this further with the location data we could also visualize the migratory behavior of the birds.
 
@@
 
-* bird/non-bird: another dataset to have definite categorization
-* Datensatz mit Anzahl Flügen allgemein für proportionale Angaben
-* Start/Destination des Flugs für Richtung usw.
-* Vergleich mit anderen Ländern
-* Closer look at missing values
+For further data exploration one could consider the following points:
+
+* Use another dataset with definite categorization for distinguishing birds and non-bird wildlife,
+* Use a dataset with the number of flights of airlines (in the U.S.) in general to have proper values for the proportions/ratios of flights with and without collisions.
+* Look for more information about the flights like start and destination for having direction, length of the flight etc. which could be included in further investigation,
+* Compare these results with those of other countries and see if there are any significant differences or similarities or
+* for more clean data: Have a closer look at the missing values and see if anything could be further replaced or imputed.
 

### 2021-07-19 draft_final.Rmd
Minor corrections

@@
 To begin, we first have a look at the dataset we're working with.
-This is the *birds* dataset form the *openintro* package [@birds_doc]: It is a collection of all **collisions between aircraft and wildlife** that were reported to the US Federal Aviation Administration **between 1990 and 1997**, with details on the circumstances of the collision.
+This is the *birds* dataset from the *openintro* package [@birds_doc]: It is a collection of all **collisions between aircraft and wildlife** that were reported to the US Federal Aviation Administration **between 1990 and 1997**, with details on the circumstances of the collision.
 It consists of **19302 observations and 17 variables** which are given in the table below.
@@
 #### Data Preparation: 
-Cleaning birds_seen variable: The 9412. observation had one wrong entry in the birds seen variable which has been replaced with NA, because we did not want to introduce a new level for one entry.
-
-Factor Levels: The factor levels of the following variables have been ordered for visualization purposes
+Cleaning *birds_seen* variable: The 9412th observation had one wrong entry in the *birds_seen* variable which has been replaced with *NA*, because we did not want to introduce a new level for one entry.
+
+Factor Levels: The factor levels of the following variables have been ordered for visualization purposes.
 
@@
   + ac_mass
-  + num_engs
-
+  + num_engs \
 * Individual ordering: 
-  + effect: reordering to put "None" effect first and seperate it from the "effect", to achieve a separation between "None" effect and effects
-  + phase_of_flt: ordering according from parking, over start of flight to end of flight
+  + effect: Reordered to put "None" effect first and separate it from the remaining "actual effect", to achieve a separation between "None" effect and effects.
+  + phase_of_flt: Ordered according from parking, over start of flight to end of flight
 
@@
 
-birds_struck: We are using the "remarks" variable to check for spare parts from which we can determine that no wild life has been strike.
+birds_struck: We are using the "remarks" variable to check for spare parts/hints in text from which we can determine that no wild life has been struck.
 By doing this we were able to decrease the missing values by 19 (from 39 to 20).
@@
 birds_seen: We assume that missing values in the "birds_seen" variable indicate that the pilot did not see wildlife before the strike occurred.
-Thus, we set all NA values in the "birds_seen" variable as "None".
+Thus, we set all *NA* values in the "birds_seen" variable as "None".
 This leads to `r sum(is.na(birds$birds_seen))` values which are set as "None".
@@
 Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is 0.217+0.069=0.286.
-The contingencytable looks like the following:
+The contingency table looks like the following:
 
@@
                       calendar_week=c("39", "48"))
-kable(map_date, format="markdown", caption="Feature Engineering of the date variable")
+kable(map_date, format="markdown", caption="Examples for Feature Engineering of the date variable")
 ```
@@
 We can see that we can find out a lot about bird strikes and aircraft collisions based on this dataset.
+
 We saw that states that people travel to and from the most also have the most collisions.

### 2021-07-20 draft_final.Rmd
Added plot description and interpretation as well as summary.

@@
 
-```{r echo=FALSE, warning=FALSE,message = FALSE,fig.align = "center", fig.asp=0.5}
+```{r echo=FALSE, warning=FALSE, message = FALSE,fig.align = "center", fig.asp=0.5}
 
@@
 top50_aircrafts_wordcloud <- ggplot(top50) +
-  aes(label = atype, size=n, color=ac_mass, angle=angle) +
+  aes(label = atype, size=n, color=ac_mass)+#, angle=angle) +
   geom_text_wordcloud_area(
@@
   scale_size_area(max_size = 3) +
-  scale_color_brewer(palette="BuPu") +#scale_color_continuous("Blues") +
+  scale_color_brewer(palette="BuPu") +
   #ggtitle("top 50 aircraft models causing strikes")
@@
 
-The left plot shows that especially the B-737 class (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727.
+The left plot shows that especially the **B-737 class** (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727.
 "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas.
 The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes.
+The B-737 is a narrow-body airliner produced by Boeing at its Renton Factory in Washington and was developed for short and thin routes.
 Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this.
@@
 
-From the right plot we can see, that aircraft models with higher max speed are also able to fly higher.
+From the right plot we can see, that aircraft **models with higher max speed are also able to fly higher**.
 This can be also seen from the linear model, that has been estimated by geom_smooth.
 The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high.
-In addition we can see via the coloring by ac_mass that lighter aircrafts tend to have less max_speed and max_height.
-We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent.
-The group with ac_mass 27001-272000 kg) is the most prominent group under the Top50 aircrafts.
+In addition we can see via the coloring by ac_mass that **lighter aircrafts tend to have less max_speed and max_height**.
+We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the Top50.
+The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the Top50 aircrafts.
 Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.
 
-The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all aircraft models (n=`r nrow(machines)`.
+The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all `r nrow(machines)` aircraft models.
 Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.
@@
 
-From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines.
-We can also see that lighter machines tend to have less engines first group includes nearly only machines with ac_mass=1, second group is mixed with ac_mass 2-4 and the last two groups include mostly aircrafts with ac_mass=4.
+From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines. We can also see that lighter machines tend to have less engines: 
+* first group includes nearly only machines with aircraft mass = 1 (less than 2250 kg), 
+* second group is mixed with aircraft mass 2-4 (2251-272000 kg), 
+* the last two groups include mostly aircrafts with mass = 5 (27001-272000 kg).
+
 Looking at the contingency table also confirms that:
 
-```{r echo=FALSE, fig.asp=0.5}
+```{r echo=FALSE, fig.asp=0.5, fig.align="center"}
 # create some fake data (2x2, since we're building a confusion matrix) 
@@
 ```
-Interesting is that for aircrafts with num_engs=1, there is no model with a speed > 200 and height > 10000.
-So we could conclude from that this is the maximum height and speed possible with one engine.
+Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot.
+So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.
 
@@
 In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect).
+Most of the time we will use barplots with position fill, mosaicplot and spineplots as we want to compare proportions and check for dependencies.
 We will first start with the analysis of strikes.
@@
 birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
-  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
+  geom_mosaic(aes(x = product(birds_struck,birds_seen), fill=birds_struck)) +
   scale_fill_brewer(palette="Blues") +
-  theme(axis.text.y=element_blank(), 
+  #scale_fill_manual(values=brewer.pal(5,"Blues"))+
+  theme(
+    axis.text.y=element_blank(), 
         axis.title.y=element_blank(),
@@
 The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
-So **most often we do not see a birds before a strike** happens and most often 1 bird is struck.
+So **most often we do not see birds before a strike** happens and most often 1 bird is struck.
 But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones.
 If the pilot sees birds, **quite often less birds were hit than seen**.
-Interesting here is, that if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
-The event of hitting less birds than seen will later be analyzed more in detail.
-
-The top right plot shows that most often strikes are occurring during the approach phase and then about equally likely a strike happens during Landing, Climb and Take-Off phase.
-
-From the bottom right plot we can see that strikes happen most often during the Day with over 10500 occurences followed by Night with about 4500 occurences.
+For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+The event of hitting less birds than seen will later be analyzed more in detail in chapter 3.3.
+
+The top right plot shows that collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll.
+The amount of collision in Approach phase is highest and about 2 times higher than the other start and end phases.
+During the middle of the flight in phases "En Route" and "Descent" less collisions occur.
+
+From the bottom right plot we can see that strikes happen most often during the Day with over 10500 collisions followed by Night with about 4500.
 Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).
@@
             stat='count',
-            size = 3,
+            size = 2.3,
             position=position_fill(vjust = 0.5))+
@@
 aes(x=ac_mass, fill=birds_struck)+
-  scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
+  #scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
+  scale_fill_manual(values=brewer.pal(5,"Blues")[c(1,4,5)]) +
   geom_bar(position = "fill") +
@@
 The proportion of some clouds is more or less the same across all groups.
-From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy.
-
-TODO: add Chi-squared independence test for sky and birds struck
-http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r
+From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy or the pilot as well as the birds are not able to see each other because of the clouds and so nobody can try to prevent the collision by dodging the other one.
+
+On the right side we can see a subset of the birds_struck groups, which focuses on the minor categories in our dataset.
+From that we can see that the aircraft mass has an influence on the amount of birds struck.
+For the lightest aircraft class up to 2250 kg the proportion of 0 birds struck is very high, which means that the pilot has seen a bird and was likely to hit one, but he did not.
+For this category we would expect that lighter machines are able to initiate evasive maneuvers as they are more agile.
+The relationship of striking less birds than seen will be analyzed more in detail in chapter 3.3.
+It also stands out, that the proportion of striking over 100 birds is significantly higher for the heaviest aircrafts above 272000 kg.
+This will most likely be due to the fact that heavier machines are also larger in size and thus they will hit more birds when flying through a flock of birds compared to lighter and thus smaller machines.
+
+### 3.1.1 Height of collisions above ground
 
 We will now have a look on the height variable together with phase of flight.
-
-### 3.1.1 Closer look on height of collisions
-
-The top plot shows a boxplot of height for flight phases, which are above ground and on the bottom we will have a closer look on the height for the flight phases "En Route" and "Descent", because they show the most variability in height.
+The top plot shows a boxplot of height for flight phases, which are above ground.
 We defined Climb, En Route, Descent, Approach as phases which are above ground level.
+On the bottom plot we will then have a closer look on the height for the flight phases "En Route" and "Descent" by visualizing the density.
 
@@
   aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
-  scale_fill_brewer(palette='Pastel2') +
+  scale_fill_manual(values=brewer.pal(4,"Pastel2")[2:3]) +
   geom_density(adjust=1.5,alpha=.8) +
@@
 The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
-The other two boxplots for "En Route" and "Descent" are more far away from ground.
+The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data.
+Thus, we will focus these two.
 Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
-
-We are performing a t.test to compare the two groups and check for equality.
+We will investigate more the differences of the two distributions by using a density plot.
+
+Both flight phases have the highest density for lower heights up to about 5000 foot.
+The curve for En Route is much steeper, but a little more narrow compared to "Descent".
+Both density cross at about 3750 which is the highest point of "Descent" phase.
+From that height upwards the probability to strike a birds is higher when being in Descent phase.
+This will be most likely because of the angle of the aircraft at which it is flying.
+In phase En Route it will be most likely parallel to the ground, so the height of the airplane does not change during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for Approach phase.
+
+We are performing a t.test to compare the two group height ("Descent" and "En Route"), so that we can confirm that both distributions are unequal.
 
@@
 
-The T-Test also show that the group means are different from each other.
-But surprisingly the **mean in group "Descent" is higher** than mean of "En Route".
-One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase where the aircraft is loosing height to prepare for approach or during flight in general.
+The T-Test also shows that the group means are different from each other and that the **mean in group "Descent" is higher** than mean of "En Route".
+One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase, where the aircraft is loosing height to prepare for approach or during flight in general.
 So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route".
@@
 So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**.
-And especially a negative angle increases the probability of hitting birds, based on the findings here and from 3.1.1 by comparing the number of collisions by phase of flight.
-
-### 3.1.2 density birdsstuck by height for flight phases above ground
+And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 by comparing the number of collisions by phase of flight.
+
+### 3.1.2 Density of height by birds_struck for flight phases above ground
 
@@
 density
-```
-
-The density plot shows a **high peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
-This is, because most of the collisions happen near ground level at flight phases Take-Off, Climb, Approach and Landing Roll (see 3.1 Number of collisions by pflight phase).
-For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 150 foot to 500 foot.
+
+#above_ground[above_ground$height>1500 & above_ground$height < 2000 & above_ground$birds_struck=="Over 100",]
+```
+
+The density plot shows that the probability for striking a bird is highes near ground level.
+This plot shows a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
+So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases Take-Off, Climb, Approach and Landing Roll have the highest number of strikes (see 3.1 Number of collisions by pflight phase).
+For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 foot to 500 foot.
 Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**.
 For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here.
-Unfortunately we have mostly missing values in the variable species for these observations, so we can not get any information about the birds species.
+So for this category we have the about the same probability of hitting birds near ground level and at about 1800 foot.
+One of the involved species here was a **flock of about 500 gulls**.
+Unfortunately the other observations had missing values in the variable "species", so we can not get more information about the species involved.
 
@@
 
-there are only two levels in the data available for birds seen and we added one level for the missing values.
+There originally were only two levels in the data available for birds seen and we added one level for the missing values.
 The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot).
@@
 #relationship of amount of birds seen with clouds?
-
 time_sums <- rep(addmargins(table(birds$birds_seen, birds$time_of_day),1)[4,],each=3)
@@
 
-Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is 0.217+0.069=0.286.
-The contingency table looks like the following:
-
-```{r echo=FALSE, fig.asp=0.5}
-round(addmargins(prop.table(table(birds$birds_seen, birds$time_of_day),2),2,FUN = mean),3) 
-```
-
-The proportion of overall **birds seen at night is much lower** compared to the other times (Dawn, Day, Dusk).
-And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).
-
-Sky: The proportion of **birds seen increases with cloudiness**.
+Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the average proportion of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285.
+So most often pilots do not see birds before a collision.
+The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day, Dusk).
+
+Sky: The proportion of **wildlife seen increases with cloudiness**.
 This is a **contradiction** to what we would have expected.
-We would have thought that with increasing cloudiness the amount of birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds.
+We would have thought that with increasing cloudiness the amount of wildlife/birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds.
 So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky.
-Another reason could be that birds in groups fly more often when its cloudy.
-
-The higher the speed & height, the less likely to see a bird before collision: 
-* for height > 2000 foot, the probability to see a birds before collision decreases 
-* for speed > 140 knots, the probability to see a bird before collision decreases
-
-## 3.3 Influences when birdsstruck < birds seen
+Another reason could be that birds fly more often when its cloudy.
+
+From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the speed & height, its becoming even less likely to see birds before a collision.
+For height greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.
+
+## 3.3 Influences when number of birds struck is less than birds seen
 
@@
 It will be true if the number of birds struck is smaller than birds seen and false if not.
-In the beginning we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck.
+In the beginning (chapter 3.1.1) we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck.
 By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.
@@
 
-less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,],each=2)
+less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,3:8],each=2)
 
 # phase of flight by birdsstruck < birds seen
-seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt),]) +
+seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt) & as.integer(seen_vs_strike$phase_of_flt)>2,]) +
   aes(x=phase_of_flt, fill=less_strike_than_seen) +
@@
  # geom_text(aes(label=..count..),
-  geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) > 0.1,round(..count../less_strike_sums[2:16],2) , "")),
+ # geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) >
+  geom_text(aes(label=ifelse(round(..count../less_strike_sums,2) >
+  #0.1,round(..count../less_strike_sums[2:16],2) , "")),
+                               0.1,round(..count../less_strike_sums,2) , "")),
             stat='count',
@@
 
-Comparing less strike than seen across phases of flight we can see, that "En Route" it is about 12% more likely to strike less birds than seen compared to "Descent".
-So it is not even less likely to strike birds in general, but also to strike less birds than seen.
-The num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
-lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife.
-This might be due to better maneuvarability of the lighter aircrafts.
-Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines.
+For the left barplot we observe a similar relationship compared to the plot of number of collisions by flight phases from chapter 3.1.1, over the start, middle and landing of the flight.
+The proportion of less strikes is increasing from the start of the flight (Take-Off Run) and reaches the highest proportion of 0.69 in the phase "En Route" and then decreases again, where it reaches a minimum of 0.48 in the last phase "Landing Roll".
+So we see again, that the angle of the flight must have an influence on striking less birds than seen.
+The number of engines and aircraft mass also have an influence of how much birds are hit out of the birds seen.
+Lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife.
+This might be again due to better maneuverability of the lighter and thus smaller aircrafts.
+Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines which matches also with the findings of the unique aircraft models from chapter 2.
 
@@
 
+The description of the effect variable from the dataset is insifficient.
+We can not really find out what is meant by the effect variable.
+Does for example "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision.
+We interpreted this variable as it really is the effect on the aircraft because a collision has occurred.
 The top plot shows the different proportions of effect for each category of birds_struck.
@@
 
-birds_struck: The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
+The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
 Especially the proportion of "Precautionary Landing" increases.
-For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off".
-
-ac_mass: the higher the aircraft mass the less likely the collision has an effect on the aircraft.
-
-birds_seenYes: The "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and Precautionary Landing and Aborted Take-off are a little higher.
-We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike.
-But from the description of the dataset we can not really find out what is meant by the effect variable.
-Does "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision.
-When birds were seen we would expect to have less amount of effect on the aircraft, cause he could try to dodge the birds or abort take off/ land precautionary.
+For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off", which makes sense, because the pilot had to intervene somehow to prevent the collision.
+For the aircraft mass, we can see that the larger the aircraft mass the less likely the collision has an effect on the aircraft.
+This might be due to the fact, that heavier machines are more robust.
+But for aircraft with an mass=5 (greater 272000 kg) we can see a an increase in the proportion of effect, which might be due to the fact that these machines have a higher probability to strike over 100 birds, which will then also have an effect on the most robust aircrafts.
+
+Having a look on the last plot, we can see that the "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and "Engine Shutdown" and "Other" are a little higher.
+We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike, because the pilot could then try to prevent the collision.
 
@@
 
+B-737 class causes the most strikes.
+It includes aircrafts from Boeing which were developed for short and thin routes.
+Aircraft models with higher max speed are also able to fly higher.
+Lighter machines tend to have less engines.
+With one engine max.
+speed and max.
+height is limited to 200 knots and 10000 foot
+
+Most of the time we do not see wildlife/birds before a collision and if the pilot sees something, he most often hits less amount of birds/wildlife than seen.
+Collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll.
+The proportion of overcast increases with the number of birds struck, this might be because of birds flying in flocks more often when it is cloudy or due to the lack of vision which prevents evasive maneuvers.
+The mass of the aircraft has an influence on the amount of birds struck, lighter machines are most likely more agile and smaller in size, thus they can better initiate evasive maneuvers, especially compared to machines with a mass greater 275000 kg, which have an about three times higher proportion of striking over 100 birds.
+Bird strikes have a higher probability to occur near ground level and the probability of hitting birds is dependent on the phase of the flight.
+The probability to hit a bird is higher when the aircraft is having an negative angle and looses height.
+
+Most often we do not see birds before a collision and at night its even less likely to see birds.
+Surprisingly the proportion of wildlife seen increases with cloudiness, this phenomenon needs to be further analyzed.
+Greater speed and height decreases the probability to see a bird.
+
+If the aircraft is flying parallel to the ground the proportion of pilots striking less birds out of the seen is higher.
+The greater the number of engines and aircraft mass the less likely less birds are hit out of the birds seen.
+
+The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
+The larger the aircraft mass the less likely the collision has an effect on the aircraft.
+There is less proportion of effect when birds were seen, which is a contradiction to our intrepretation of the effect variable.
+
 We saw that states that people travel to and from the most also have the most collisions.

### 2021-07-20 draft_final.Rmd
Update draft_final.Rmd

@@
 ```{r echo=FALSE, fig.asp=0.5}
-birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck),]) +
+birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck) & as.integer(birds$phase_of_flt)>2,]) +
   aes(x = phase_of_flt, fill=birds_struck) +
@@
   geom_mosaic(aes(x = product(birds_struck,birds_seen), fill=birds_struck)) +
-  scale_fill_brewer(palette="Blues") +
-  #scale_fill_manual(values=brewer.pal(5,"Blues"))+
+  #scale_fill_brewer(palette="Blues") +
+  scale_fill_manual(values=c("#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C"))+
   theme(
@@
   #scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
-  scale_fill_manual(values=brewer.pal(5,"Blues")[c(1,4,5)]) +
+  scale_fill_manual(values=brewer.pal(6,"Blues")[c(1,4,5)]) +
   geom_bar(position = "fill") +
@@
 
+\newpage
 ### 3.1.1 Height of collisions above ground
@@
 
+\newpage
 ## 3.2 Analysis of birds seen
@@
 
+\newpage
 ## 3.3 Influences when number of birds struck is less than birds seen
@@
 
+\newpage
 ## 3.4 Analysis of effect variable
@@
 
-```{r echo=FALSE, fig.asp=0.5}
+```{r echo=FALSE, fig.asp=0.3}
 
@@
 
-birds_struck_effect / (birds_effect_acmass + birds_effect_seenYes)
-
+birds_struck_effect
+
+```
+```{r echo=FALSE, fig.asp=0.35}
+(birds_effect_acmass + birds_effect_seenYes)
 ```

### 2021-07-20 draft_final.Rmd
Minor corrections

@@
 
-factor_ac_mass <- c("<=2250 kg", "2251-5700 kg", "5701-27000 kg", "27001-272000 kg", ">272000 kg")
-birds$ac_mass <- as.factor(birds$ac_mass)#, levels=factor_ac_mass)
-birds$ac_mass_names <- revalue(as.factor(birds$ac_mass),     c("1"="<2250",        "2"="2251-5700",        "3"="5701-27000",        "4"="27001-272000",       "5"=">272000"))
+#factor_ac_mass <- c("<=2250 kg", "2251-5700 kg", "5701-27000 kg", "27001-272000 kg", ">272000 kg")
+#birds$ac_mass <- as.factor(birds$ac_mass)#, levels=factor_ac_mass)
+#birds$ac_mass_names <- revalue(as.factor(birds$ac_mass),     c("1"="<2250",        "2"="2251-5700",        "3"="5701-27000",        "4"="27001-272000",       "5"=">272000"))
+
+ac_mass_names<- c("1"="< 2.25 t", "2"="2.25-5.7 t", "3"="5.7-27 t", "4"="27-272 t", "5"="> 272 t")
+birds$ac_mass_renamed<-revalue(as.factor(birds$ac_mass), ac_mass_names)
+
 birds$num_engs <- as.factor(birds$num_engs)
@@
 \newpage
+
+# Introduction
+
+In this paper we want to investigate a dataset of aircraft-wildlife collisions in the U.S. in the 90s. We start by introducing the data and some steps we took regarding data preparation.
+For our analysis we then first take a look at what kind of aircrafts are frequently involved in bird/wildlife strikes.
+After that we examine what kind of circumstances lead to such collisions and if there are any dependencies between certain variables.
+In the following chapter we then also take location into account to analyze if this has any influence on the kind of wildlife involved in the collisions.
+For the last research topic we focus on the time variable and put that in relation to the previous results.
+
+We then finish with a summary of our findings and further research questions that could possibly be further analyzed in the future.
 
@@
 
-```{r echo=FALSE, message=FALSE, fig.asp=0.5}
+```{r echo=FALSE, message=FALSE, fig.asp=0.4}
 #get rows where birds struck is missing
@@
             ac_mass = median(as.integer(ac_mass)),
-            ac_mass_names = median(as.integer(ac_mass_names)))
+            ac_mass_renamed = median(as.integer(ac_mass_renamed)))
 
@@
 machines$ac_mass <- factor(machines$ac_mass, levels=1:5)
-machines$ac_mass_names <- factor(machines$ac_mass_names)
-
-machines$ac_mass2 <- ifelse(as.integer(machines$ac_mass)<3,1,ifelse(as.integer(machines$ac_mass)==3,2,3))
-machines$ac_mass2 <- factor(machines$ac_mass2, levels=1:3)
-
-top10_aircrafts_bar <- machines %>%
-arrange(desc(n)) %>%
-head(10) %>%
-ggplot(aes(x=reorder(atype, n), y=n, fill = ac_mass))+
-  geom_bar(stat="identity") +
-  theme(legend.position="none") +
-  labs(title= "Top 10 frequent strikes by aircraft", x="aircraft model") +
-  coord_flip()
+machines$ac_mass_renamed <- revalue(factor(machines$ac_mass_renamed, levels=1:5), ac_mass_names)
 
@@
   geom_text_wordcloud_area(
-    #mask = png::readPNG(source="flugzeug6-3.png"),
-    #mask = png::readPNG(source="/Users/Vy/Documents/flugzeug6-3.png"),
+    mask = png::readPNG(source="flugzeug6-3.png"),
+    #mask = png::readPNG(source="/Users/Vy/Documents/R_Projects/Data_Visualization/flugzeug6-3.png"),
     rm_outside = TRUE) +
@@
 #aircraft_speed_height <- ggplot(machines[!is.na(machines$ac_mass),]) +
-  aes(x=max_speed, y=max_height, color = ac_mass) +
+  aes(x=max_speed, y=max_height, color = ac_mass_renamed) +
   geom_point() +
@@
   theme(legend.position="right") +
-  scale_color_brewer(palette="BuPu") +
+  scale_color_brewer(name="ac_mass in tons", palette="BuPu") +
   guides(fill = FALSE)+
@@
 ```{r echo=FALSE, fig.asp=0.5}
-machines_c <- machines[complete.cases(machines),]
 p <- ggplot(machines[!is.na(machines$num_engs) & !is.na(machines$ac_mass),]) +
-  aes(x=max_speed, y=max_height, color = ac_mass) +
+  aes(x=max_speed, y=max_height, color = ac_mass_renamed) +
   geom_point() +
-  scale_color_brewer(palette="BuPu") +
+  scale_color_brewer(name="ac_mass in tons", palette="BuPu") +
   theme(plot.title = element_text(
@@
 birds_strike_ac_mass <- ggplot(birds_struck_minorities)+
-aes(x=ac_mass, fill=birds_struck)+
+aes(x=ac_mass_renamed, fill=birds_struck)+
   #scale_fill_manual(values=c("#eff3ff", "#3182bd", "#08519c")) +
@@
   theme.title+
-  labs(subtitle = "Birds struck (subset) by aircraft mass")
+  labs(subtitle = "Birds struck (subset) by aircraft mass") + 
+  xlab("ac_mass in tons") + 
+  theme(axis.text.x = element_text(angle = 90))
 
@@
 
-```{r echo=FALSE, fig.asp=0.5}
+```{r echo=FALSE, fig.asp=0.3}
 #Create new column
@@
             position=position_fill(vjust = 0.5))+
-  coord_flip() +
+  #coord_flip() +
   theme(legend.position="right", axis.title.y = element_blank())+
@@
 
-#ac_mass by leess-strike-than-seen (MOSAIC)
+#ac_mass by less-strike-than-seen (MOSAIC)
 seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
-  geom_mosaic(aes(x=product(ac_mass), fill=less_strike_than_seen)) +
+  geom_mosaic(aes(x=product(ac_mass_renamed), fill=less_strike_than_seen)) +
   scale_fill_manual(values = color_pal_TF)+
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())+
-  ylab("")
-
-#num_engs by leess-strike-than-seen (MOSAIC)
+  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 90))+
+  ylab("") + xlab("ac_mass in tons")
+
+#num_engs by less-strike-than-seen (MOSAIC)
 seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
@@
 
-seen_vs_strike_phase + seen_vs_strike_ac_mass / seen_vs_strike_num_engs
+seen_vs_strike_phase
+```
+```{r echo=FALSE, fig.asp=0.35}
+seen_vs_strike_ac_mass + seen_vs_strike_num_engs
 ```
@@
 #mosaic plot better than barplot visualizes also absolute differences
-birds_effect_acmass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$effect),]) +
-  geom_mosaic(aes(x = product(ac_mass), fill=effect)) +
+birds_effect_acmass <- ggplot(birds[!is.na(birds$ac_mass_renamed) & !is.na(birds$effect),]) +
+  geom_mosaic(aes(x = product(ac_mass_renamed), fill=effect)) +
   scale_fill_brewer(palette="OrRd") +
@@
         axis.text.y=element_blank(), 
+        axis.text.x = element_text(angle =90),
         axis.ticks.y = element_blank(),
@@
   theme.title+
-  labs(subtitle = "Effect by aircraft mass")
+  labs(subtitle = "Effect by aircraft mass") +
+  xlab("ac_mass in tons")
 
@@
   theme.title+
-  labs(xlab= "birds seen?", subtitle = "Effect by birds seen or not")
+  labs(subtitle = "Effect by birds seen or not") +
+  xlab("birds_seen = YES")
 
@@
 # Number of collisions per state
-state_plot_df<-birds[birds$birds_struck != "0" & !is.na(birds$birds_struck) & !is.na(birds$state) & !is.na(birds$ac_mass),]
-state_plot_df$ac_mass_renamed<-revalue(as.factor(state_plot_df$ac_mass), 
-     c("1"="2250 kg or less", "2"="2251-5700 kg", "3"="5701-27000 kg", "4"="27001-272000 kg", "5"="above 272000 kg"))
+state_plot_df<-birds[birds$birds_struck != "0" & !is.na(birds$birds_struck) & !is.na(birds$state) & !is.na(birds$ac_mass_renamed),]
+#state_plot_df$ac_mass_renamed<-revalue(as.factor(state_plot_df$ac_mass), 
+     #c("1"="2250 kg or less", "2"="2251-5700 kg", "3"="5701-27000 kg", "4"="27001-272000 kg", "5"="above 272000 kg"))
 state_plot<-ggplot(data = state_plot_df) +
   aes(y = fct_rev(fct_infreq(state)), fill = ac_mass_renamed) + 
-  guides(fill=guide_legend(title="Mass of the aircraft")) + 
+  guides(fill=guide_legend(title="Mass of aircraft in tons")) + 
   scale_fill_brewer(palette="Purples") +
@@
   theme(legend.position="bottom") +
-   theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
+   theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
   theme(plot.margin=unit(c(5.5,20,5.5,5.5),"points")) +
@@
   theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(),
-        axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
+        axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
   theme(plot.margin=unit(c(5.5,5.5,5.5,20),"points")) +
@@
 
+\newpage
+
 ## 4.4 Vulture collisions per State
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, warning=FALSE, message=FALSE, }
 birds.filt<-birds[!is.na(birds$birds_struck),]
@@
 ```
-```{r echo=FALSE, fig.asp=0.5}
+```{r echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.5}
 (sky_month_plot + tod_plot)
@@
 
-```{r echo=FALSE, fig.asp=1.0}
+```{r echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.0}
 (month_plot / woy_plot) / (week_plot + half.year.plot)
@@
 
-```{r echo=FALSE, fig.asp=0.8}
+```{r echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8}
 birds$struck_num<-unclass(birds$birds_struck)
@@
 
+\newpage
+
 ## 5.4 Hawk collisions over Time and by State
@@
 
-```{r echo=FALSE, fig.asp=1.0}
+```{r echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.0}
 birds.with.statenames<-merge(x = birds.filtered.struck.num, y = df.state, by.x = "state", by.y = "state.abb")

### 2021-07-20 draft_final.Rmd
Final draft PDF

@@
   geom_text_wordcloud_area(
-    mask = png::readPNG(source="flugzeug6-3.png"),
+    #mask = png::readPNG(source="flugzeug6-3.png"),
     #mask = png::readPNG(source="/Users/Vy/Documents/R_Projects/Data_Visualization/flugzeug6-3.png"),

### 2021-07-21 draft_final.Rmd
Formatting of variables in text and some plot labelling.

@@
 
-We then finish with a summary of our findings and further research questions that could possibly be further analyzed in the future.
+We then finish with a summary of our findings and further research questions that could possibly be analyzed in the future.
 
@@
 
-birds_struck: We are using the "remarks" variable to check for spare parts/hints in text from which we can determine that no wild life has been struck.
+*birds_struck*: We are using the *remarks* variable to check for spare parts/hints in text from which we can determine that no wild life has been struck.
 By doing this we were able to decrease the missing values by 19 (from 39 to 20).
 
-birds_seen: We assume that missing values in the "birds_seen" variable indicate that the pilot did not see wildlife before the strike occurred.
-Thus, we set all *NA* values in the "birds_seen" variable as "None".
-This leads to `r sum(is.na(birds$birds_seen))` values which are set as "None".
+*birds_seen*: We assume that missing values in the *birds_seen* variable indicate that the pilot did not see wildlife before the strike occurred.
+Thus, we set all *NA* values in the *birds_seen* variable as *None*.
+This leads to `r sum(is.na(birds$birds_seen))` values which are set as *None*.
 
@@
 The dataset contains `r length(unique(birds$atype))` different aircraft models.
-We extracted the overall maximum speed and height for each aircraft to get an overview of the aircraft models and the relationship between speed and height.
-The following left plot shows the Top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the Top 50 aircrafts colored by their ac_mass.
+We extracted the overall maximum speed (*max_speed*) and height (*max_height*) for each aircraft to get an overview of the aircraft models and the relationship between speed and height.
+The following left plot shows the top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the top 50 aircrafts colored by their *ac_mass*.
 
@@
 
-The left plot shows that especially the **B-737 class** (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727.
+The left plot shows that especially the **B-737 class** (including B-737-200 and B-737-300) causes the most strikes together with MD-80 and B-727.
 "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas.
-The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes.
+The model causing the most strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes.
 The B-737 is a narrow-body airliner produced by Boeing at its Renton Factory in Washington and was developed for short and thin routes.
@@
 
-From the right plot we can see, that aircraft **models with higher max speed are also able to fly higher**.
+From the right plot we can see, that aircraft **models with higher *max_speed* are also able to fly higher**.
 This can be also seen from the linear model, that has been estimated by geom_smooth.
-The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high.
-In addition we can see via the coloring by ac_mass that **lighter aircrafts tend to have less max_speed and max_height**.
-We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the Top50.
-The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the Top50 aircrafts.
-Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.
-
-The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all `r nrow(machines)` aircraft models.
-Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.
+The correlation of *max_speed* and *max_height* for the top 50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high.
+In addition we can see via the coloring by *ac_mass* that **lighter aircrafts tend to have less max_speed and max_height**.
+We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the top 50.
+The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the top 50 aircrafts.
+Two aircraft models (B-747-1/200 and B-747-400) with *ac_mass* of more than 272000 kg are also Boeings and present in the top 50 which are the only aircrafts of that weight class overall.
+
+The next plot shows again the *max_speed* by *max_height* splitted by number of engines and colored by *ac_mass* but for all `r nrow(machines)` aircraft models.
+Since some models had very few strikes it can happen that they have been assigned an *max_height* of about 0.
 
@@
 
-From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines. We can also see that lighter machines tend to have less engines: 
-* first group includes nearly only machines with aircraft mass = 1 (less than 2250 kg), 
-* second group is mixed with aircraft mass 2-4 (2251-272000 kg), 
-* the last two groups include mostly aircrafts with mass = 5 (27001-272000 kg).
+From the plot we can see, that again higher *max_speed* causes higher *max_height* and most of the aircraft models have two engines.
+Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot.
+So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.
+We can also see that lighter machines tend to have less engines:
+
+* first group includes nearly only machines with aircraft mass less than 2250 kg, 
+* second group is mixed with aircraft mass between 2251-272000 kg, 
+* the last two groups include mostly aircrafts with a mass of 27001-272000 kg.
 
@@
 ```
-Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot.
-So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.
 
@@
 # 3. Analysis of influences for birds_struck, birds_seen and effect
-In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect).
+In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: *birds_struck*), the amount of birds that have been seen by the pilot before a strike happened (variable: *birds_seen*), and the different effects a strike had on the aircraft (variable: *effect*).
 Most of the time we will use barplots with position fill, mosaicplot and spineplots as we want to compare proportions and check for dependencies.
@@
 
-The following plot on the left shows the amount of birds struck out birds_seen.
-On the top right we can see the amount of collisions by each phase of flight and on the bottom right we can see the amount of collisions by time of day.
-For coloring we use the birds_struck variable.
+The following plot on the left shows the amount of birds struck out of birds seen.
+On the top right we visualized the number of collisions by each phase of flight and on the bottom right the plot shows the amount of collisions by time of day.
+For coloring we use the *birds_struck* variable.
 
@@
   coord_flip() +
+  theme.title+
   theme(legend.position="none",
+        axis.title = element_text(vjust=-10),
         axis.title.y=element_blank(),
         axis.title.x=element_blank(),
-        plot.subtitle = element_text(hjust =1)) +
-  theme.title+
+        plot.subtitle = element_text(hjust =4)) +
   labs(xlab = "phase of flight",
@@
   coord_flip() +
+  theme.title+
   theme(legend.position="none",
@@
         axis.title.x=element_blank(),
-        plot.subtitle = element_text(hjust =1)) +
-  theme.title+
+        plot.subtitle = element_text(hjust =3)) +
   labs(xlab = "",
@@
 The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100".
-So **most often we do not see birds before a strike** happens and most often 1 bird is struck.
-But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones.
+So **most often we do not see birds before a strike** happens and most often only 1 bird is struck.
+But if the pilot sees something, than according to the data he only sees birds in groups, where larger groups of birds are less likely than smaller ones.
 If the pilot sees birds, **quite often less birds were hit than seen**.
-For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen.
+For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group *birds_seen* between 11-100, about 85% of the time the pilot hits less birds than seen.
 The event of hitting less birds than seen will later be analyzed more in detail in chapter 3.3.
@@
 
-From the bottom right plot we can see that strikes happen most often during the Day with over 10500 collisions followed by Night with about 4500.
-Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).
-
-We will now have a look on the influence of the sky variable for the different amounts of birds struck.
+From the bottom right plot we can see that strikes happen most often during the day with over 10500 collisions followed by night with about 4500.
+Interesting to see is that strikes with larger groups of birds (11-100) happen mostly during the day (small values are not visible in the plot).
+
+We will now have a look on the influence of the sky variable for the different amounts of *birds_struck* and analyze the influence of the aircraft mass.
 
@@
 
-From the plot on the left we can see that with **increasing number of birds_struck the proportion of overcast increases** and the proportion of no clouds decreases.
+From the left plot we can see that with **increasing number of *birds_struck* the proportion of overcast increases** and the proportion of no clouds decreases.
 The proportion of some clouds is more or less the same across all groups.
@@
 
-On the right side we can see a subset of the birds_struck groups, which focuses on the minor categories in our dataset.
+On the right side we can see a subset of the *birds_struck* groups, which focuses on the less frequent categories in our dataset.
 From that we can see that the aircraft mass has an influence on the amount of birds struck.
 For the lightest aircraft class up to 2250 kg the proportion of 0 birds struck is very high, which means that the pilot has seen a bird and was likely to hit one, but he did not.
-For this category we would expect that lighter machines are able to initiate evasive maneuvers as they are more agile.
+For this category we would expect that lighter machines are better able to initiate unplanned evasive maneuvers as they are more agile.
 The relationship of striking less birds than seen will be analyzed more in detail in chapter 3.3.
@@
 The top plot shows a boxplot of height for flight phases, which are above ground.
-We defined Climb, En Route, Descent, Approach as phases which are above ground level.
+We defined Climb, En Route, Descent and Approach as phases which are above ground level.
 On the bottom plot we will then have a closer look on the height for the flight phases "En Route" and "Descent" by visualizing the density.
@@
   labs(title = "Height distribution for flight phases above ground")+
+  ylab("height in foot")+
   ylim(0,20000)
@@
   geom_density(adjust=1.5,alpha=.8) +
+  xlab("height in foot")+
   xlim(0,20000)
@@
 
-The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
+The height distribution for "Climb" and "Approach" are very similar and their medians are very close to the ground. 
 The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data.
-Thus, we will focus these two.
-Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile.
+Thus, we will focus on these two.
+Interesting here is that the 25% quantile of the "Descent" boxplot is above the one from "En Route", same for the median and 75% quantile.
 We will investigate more the differences of the two distributions by using a density plot.
@@
 Both flight phases have the highest density for lower heights up to about 5000 foot.
-The curve for En Route is much steeper, but a little more narrow compared to "Descent".
-Both density cross at about 3750 which is the highest point of "Descent" phase.
-From that height upwards the probability to strike a birds is higher when being in Descent phase.
+The curve for "En Route" is much steeper, but a little more narrow compared to "Descent".
+Both density cross at about 3750 foot, which is the highest point of "Descent" phase.
+From that height upwards the probability to strike a birds is higher when being in "Descent" phase.
 This will be most likely because of the angle of the aircraft at which it is flying.
-In phase En Route it will be most likely parallel to the ground, so the height of the airplane does not change during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for Approach phase.
-
-We are performing a t.test to compare the two group height ("Descent" and "En Route"), so that we can confirm that both distributions are unequal.
+In phase "En Route" it will be most likely parallel to the ground, so the height of the airplane does not change much during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for "Approach" phase.
+
+We are performing a t-test to compare the two group heights ("Descent" and "En Route"), so that we can confirm that both distributions are not equal.
 
@@
 
-The T-Test also shows that the group means are different from each other and that the **mean in group "Descent" is higher** than mean of "En Route".
+The t-test also shows that the **group means are different** from each other, because the p-value is less than 5% or even 1% level.
+The **mean in group "Descent" is higher** than mean of "En Route".
 One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase, where the aircraft is loosing height to prepare for approach or during flight in general.
-So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route".
+So since all aircrafts mostly have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height (greater 3750 foot) when being in phase "Descent" rather than in "En Route".
 The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent".
 So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**.
-And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 by comparing the number of collisions by phase of flight.
+And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 where we found out, that the "Approach" phase has the highest amount of strikes.
 
@@
 
-We will now have another look on the density of height, but this time split by the different numbers of birds involved in the collision. The following plot shows exactly this.
+We will now have another look on the density of *height*, but this time split by the different numbers of birds involved in the collision. The following plot shows exactly this.
 
@@
   xlim(0, 3500) +
-  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")
+  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")+
+  xlab("height in foot")
 density
@@
 
-The density plot shows that the probability for striking a bird is highes near ground level.
-This plot shows a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
-So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases Take-Off, Climb, Approach and Landing Roll have the highest number of strikes (see 3.1 Number of collisions by pflight phase).
-For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 foot to 500 foot.
-Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**.
+The density plot shows that the probability for striking a bird is highest near ground level.
+The density starts with a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0.
+So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases near ground level (Take-Off, Climb, Approach and Landing Roll) have the highest number of strikes (see 3.1 Number of collisions by flight phase).
+For *birds_struck* equal to 0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 to 500 foot.
+Remarkable is also the **peek for "Over 100" birds struck at a *height* of about 1800 foot**.
 For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here.
-So for this category we have the about the same probability of hitting birds near ground level and at about 1800 foot.
+So for this category we have the same probability of hitting birds near ground level and at about 1800 foot.
 One of the involved species here was a **flock of about 500 gulls**.
@@
 There originally were only two levels in the data available for birds seen and we added one level for the missing values.
-The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot).
-At the bottom you can see two spineplots that show the two continuous variables height and speed by the number of birds seen.
+The two plots at the top show the proportion of birds seen by the *time_of_day* (left plot) and the cloudiness of the *sky* (right plot).
+At the bottom you can see two spineplots that show the proportion of *birds_seen* along the two continuous variables *height* and *speed*.
 
@@
 ```
-```{r echo=FALSE, fig.asp=0.4, fig.width = 7, fig.show="hold", fig.align = 'left'}
+```{r echo=FALSE, fig.asp=0.4, fig.width = 7, fig.show="hold", fig.align = 'center'}
 par(mfrow=c(1,2))
-spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
-title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50")
-
-spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "speed",
+spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "", col = rev(brewer.pal(n=3,name="BuGn")))
+title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50", line = 0.5)
+mtext(text = "height in foot",
+      side = 1,#side 1 = bottom
+      line = 2)
+
+spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "",
           col = rev(brewer.pal(n=3,name="BuGn")))
-title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50")
-```
-
-Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the average proportion of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285.
+title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50",line = c(0.5))
+#title(xlab="speed", line=4, cex.lab=1.2, family="Calibri Light")
+mtext(text = "speed in knots",
+      side = 1,#side 1 = bottom
+      line = 2)
+```
+
+*Time of Day*: Based on the assumption that missing values in the data for *birds_seen* stands for the fact that the pilot did not see wildlife/birds before a strike happened, the average proportion (unweighted) of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285.
 So most often pilots do not see birds before a collision.
-The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day, Dusk).
-
-Sky: The proportion of **wildlife seen increases with cloudiness**.
+The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day and Dusk).
+
+*Sky*: The proportion of **wildlife seen increases with cloudiness**.
 This is a **contradiction** to what we would have expected.
@@
 
-From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the speed & height, its becoming even less likely to see birds before a collision.
-For height greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.
+From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the *speed* & *height*, its becoming even less likely to see birds before a collision.
+For heights greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.
 
@@
 
-In the following we will have a closer look on a self-generated variable, which compares the amount of birds seen with the amount of birds_struck.
+In the following we will have a closer look on a self-generated variable, which compares the amount of *birds_seen* with the amount of *birds_struck*.
 It will be true if the number of birds struck is smaller than birds seen and false if not.
-In the beginning (chapter 3.1.1) we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck.
+In the beginning (chapter 3.1.1) we have already seen, that the number of *birds_seen* does not always infer that the same amount of birds were struck.
 By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.
 
-The plot on the left shows the proportion of less strikes than seen by phase of flight.
-On the top right we can see a comparison of the proportion of less strikes by the aircraft mass and on the bottom right we compare it by different number of engines.
-
-```{r echo=FALSE, fig.asp=0.3}
+The top plot shows the proportion of less strikes than seen by phase of flight.
+On the bottom left we can see a comparison of the proportion of less strikes by the aircraft mass and on the bottom right we compare it by different number of engines.
+
+```{r echo=FALSE, warning=FALSE,fig.asp=0.3}
 #Create new column
@@
 
-#TODO: Why is barplot different from ggplot???
-#barplot(prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2), xlab = #"less_strike_than_seen")
-#prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2)
-
 less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,3:8],each=2)
-
 # phase of flight by birdsstruck < birds seen
 seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt) & as.integer(seen_vs_strike$phase_of_flt)>2,]) +
-  aes(x=phase_of_flt, fill=less_strike_than_seen) +
+  geom_bar(aes(x=phase_of_flt, fill=less_strike_than_seen), position = "fill") +
+      geom_line(aes(x=as.numeric(phase_of_flt)-2, y=..count../less_strike_sums[seq(1,12,2)], 
+                group=less_strike_than_seen,
+                color=less_strike_than_seen),
+            stat = "count") +
+  scale_color_manual(values = c("NA","black"))+
+  guides(color = FALSE)+
   scale_fill_manual(values = color_pal_TF) +
-  geom_bar(position = "fill") +
- # geom_text(aes(label=..count..),
- # geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) >
-  geom_text(aes(label=ifelse(round(..count../less_strike_sums,2) >
-  #0.1,round(..count../less_strike_sums[2:16],2) , "")),
+  geom_text(aes(x=phase_of_flt, fill=less_strike_than_seen, 
+                label=ifelse(round(..count../less_strike_sums,2) >
                                0.1,round(..count../less_strike_sums,2) , "")),
@@
             position=position_fill(vjust = 0.5))+
-  #coord_flip() +
   theme(legend.position="right", axis.title.y = element_blank())+
@@
   labs(title="Variables influencing less strikes than seen",
-       subtitle = "Proportion of less strike than seen \nby phase of flight",
+       subtitle = "Proportion of less strike than seen by phase of flight",
        fill='less strike \nthan seen'
@@
   scale_fill_manual(values = color_pal_TF)+
-  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 90))+
-  ylab("") + xlab("ac_mass in tons")
+  theme.title +
+  theme(legend.position="none", 
+        axis.text.y=element_blank(), 
+        axis.ticks.y = element_blank(), 
+        axis.text.x = element_text(angle = 90))+
+  labs(subtitle = "Prop. of less strike than seen by aircraft mass")+
+  ylab("") + 
+  xlab("ac_mass in tons")
 
@@
   scale_fill_manual(values = color_pal_TF)+
-  theme(legend.position="none", axis.text.y=element_blank(),axis.ticks.y = element_blank())+
+  theme.title +
+  theme(legend.position="none", 
+        axis.text.y=element_blank(),
+        axis.ticks.y = element_blank())+
+  labs(subtitle = "Prop. of less strike than seen by number of engines")+
   ylab("")
@@
 
-For the left barplot we observe a similar relationship compared to the plot of number of collisions by flight phases from chapter 3.1.1, over the start, middle and landing of the flight.
+For the top barplot we observe a relationship based on the phase of flight, like we did earlier in chapter 3.1.1 when comparing the number of collisions by flight phases over the start, middle and landing of the flight.
 The proportion of less strikes is increasing from the start of the flight (Take-Off Run) and reaches the highest proportion of 0.69 in the phase "En Route" and then decreases again, where it reaches a minimum of 0.48 in the last phase "Landing Roll".
-So we see again, that the angle of the flight must have an influence on striking less birds than seen.
+So we see again, that the angle of the flight must have an influence on striking birds, this time especially on striking less birds than seen.
 The number of engines and aircraft mass also have an influence of how much birds are hit out of the birds seen.
 Lighter machines and machines with less engines tend to hit less wildlife out of the seen wildlife.
-This might be again due to better maneuverability of the lighter and thus smaller aircrafts.
-Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines which matches also with the findings of the unique aircraft models from chapter 2.
+This might be again due to better agility of the lighter and thus smaller aircrafts.
+Most prominent are aircrafts with ac_mass = 4 (27001-272000 kg) and two engines which matches also with the proportions of unique aircraft models from chapter 2.
 
@@
 
-The description of the effect variable from the dataset is insifficient.
-We can not really find out what is meant by the effect variable.
+The description of the *effect* variable from the dataset is insufficient.
+We can not really find out what is meant by the *effect* variable.
 Does for example "Precautionary landing" and "Aborted take-off" mean that the pilot did this to prevent striking birds, or did he had to do this because of the collision.
-We interpreted this variable as it really is the effect on the aircraft because a collision has occurred.
-The top plot shows the different proportions of effect for each category of birds_struck.
-The Bottom left plot shows the proportions of each effect by the aircraft mass and on the bottom right we compare effect with birds_seen or not seen.
+We interpreted this variable as it really is the effect on the aircraft because a collision occurred.
+The top plot shows the different proportions of *effect* for each category of *birds_struck*.
+The Bottom left plot shows the proportions of each *effect* by the aircraft mass and on the bottom right we compare effect with *birds_seen* or not seen.
 
@@
 
-The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
+The greater the number of *birds_struck*, the more likely that the collision has an *effect* on the aircraft.
 Especially the proportion of "Precautionary Landing" increases.
-For birds_struck 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off", which makes sense, because the pilot had to intervene somehow to prevent the collision.
+For *birds_struck* equals 0 we have very small proportion of "None" effect and high proportion of "Other" and "Aborted Take-off", which would actually speak for the *effect* variable representing an action that was done in order to prevent the collision.
+So the interpretation of the *effect* variable needs further research or the publisher of the dataset should be contacted asked for a more precise description.
 For the aircraft mass, we can see that the larger the aircraft mass the less likely the collision has an effect on the aircraft.
 This might be due to the fact, that heavier machines are more robust.
-But for aircraft with an mass=5 (greater 272000 kg) we can see a an increase in the proportion of effect, which might be due to the fact that these machines have a higher probability to strike over 100 birds, which will then also have an effect on the most robust aircrafts.
-
-Having a look on the last plot, we can see that the "None" proportion of effect is a little smaller when the pilot has seen birds before the strike and "Engine Shutdown" and "Other" are a little higher.
-We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike, because the pilot could then try to prevent the collision.
+But for aircrafts with a mass greater 272000 kg we can see a an increase in the proportion of *effect*, which might be due to the fact that these machines have a higher probability to strike over 100 birds, which will then also have an effect on the heavier aircrafts.
+
+Having a look on the last plot, we can see that the "None" proportion of *effect* is a little smaller when the pilot has seen birds before the strike and "Engine Shutdown" and "Other" are a little higher.
+We would have actually expected to have less effects on the aircraft when the pilot has seen birds before the strike, because the pilot could then try to prevent the collision, but this is not the case here.
+So this again would speak for a different interpretation of the *effect* variable.
 
@@
 So all in all, we can see that there is a relationship between wildlife and the state.
-Depending on where the aircraft collided with wildlife the most frequent species differ from each other.
+Depending on where the aircraft collided with wildlife, the most frequent species differ from each other.
 We can also extract information about the habitat of certain wildlife with the given data (e.g.
@@
 This would also match with the observation we made jut before with the *sky* variable.
-There are correspondingly also more collisions during the day in the summer which could also be because there are more flights happening compared to thee winter time.
+There are correspondingly also more collisions during the day in the summer which could also be because there are more flights happening compared to the winter time.
 Therefore the probability of more collisions is higher then, too.
@@
 speed and max.
-height is limited to 200 knots and 10000 foot
+height is limited to 200 knots and 10000 foot.
 
 Most of the time we do not see wildlife/birds before a collision and if the pilot sees something, he most often hits less amount of birds/wildlife than seen.
-Collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll.
+Collisions mostly occur at daytime and at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll.
 The proportion of overcast increases with the number of birds struck, this might be because of birds flying in flocks more often when it is cloudy or due to the lack of vision which prevents evasive maneuvers.
@@
 
-If the aircraft is flying parallel to the ground the proportion of pilots striking less birds out of the seen is higher.
+If the aircraft is flying parallel to the ground the proportion of pilots striking less birds out of the seen is higher compared to all other flight phases.
 The greater the number of engines and aircraft mass the less likely less birds are hit out of the birds seen.
 
-The greater the number of birds struck, the more likely that the collision has an effect on the aircraft.
-The larger the aircraft mass the less likely the collision has an effect on the aircraft.
-There is less proportion of effect when birds were seen, which is a contradiction to our intrepretation of the effect variable.
+The greater the number of birds struck, the more likely that the collision has an effect on the aircraft. The larger the aircraft mass the less likely the collision has an effect on the aircraft.
+There is less proportion of effect when birds were seen, which is a contradiction to our interpretation of the effect variable.
 
@@
 * for more clean data: Have a closer look at the missing values and see if anything could be further replaced or imputed.
+* The meaning of the "effect" variable is unclear, investigate more or contact the publisher of the dataset.
 

### 2021-07-21 draft_final.Rmd
Modified NA barplot and removed unnecessary y-axis labels.

@@
 
+There were no missing values for the following variables: atype, birds_seen, date, operator, opid and species.
+
 *birds_struck*: We are using the *remarks* variable to check for spare parts/hints in text from which we can determine that no wild life has been struck.
 By doing this we were able to decrease the missing values by 19 (from 39 to 20).
-
 *birds_seen*: We assume that missing values in the *birds_seen* variable indicate that the pilot did not see wildlife before the strike occurred.
@@
 This leads to `r sum(is.na(birds$birds_seen))` values which are set as *None*.
+
+The following barplot shows the amount of missing values for each of the variables, that had missing values.
 
@@
 
-ggplot(missing_df) + 
+cols_missings <- c("speed", "effect", "sky", "height", "remarks", "time_of_day", "phase_of_flt", "num_engs", "ac_mass", "state", "birds_struck")
+
+ggplot(missing_df[missing_df$column %in% cols_missings,]) + 
   aes(x=reorder(column, -count), y=count) +
   geom_bar(stat="identity", fill="snow3") +
+  theme.title+
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
-        axis.title.x = element_blank())
+        axis.title.x = element_blank())+
+  labs(subtitle = "Amount of missing values by variable")
 ```
@@
             position=position_fill(vjust = 0.5))+
-  theme(axis.text.x = element_text(angle = 90)) +
+  theme(axis.text.x = element_text(angle = 90),
+        axis.title.y = element_blank()) +
   theme.title+
@@
   xlab("ac_mass in tons") + 
-  theme(axis.text.x = element_text(angle = 90))
+  theme(axis.text.x = element_text(angle = 90),
+        axis.title.y = element_blank())
 
@@
 One of the involved species here was a **flock of about 500 gulls**.
-Unfortunately the other observations had missing values in the variable "species", so we can not get more information about the species involved.
+Unfortunately the other observations had no information about the *species* that were involved in the collision.
 
@@
 spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "", col = rev(brewer.pal(n=3,name="BuGn")))
-title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50", line = 0.5)
+title(main = "Proportion of birds seen by height", cex.main=0.8, font.main= 3, adj=0, col.main = "gray50", line = 0.5)
 mtext(text = "height in foot",
@@
           col = rev(brewer.pal(n=3,name="BuGn")))
-title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50",line = c(0.5))
-#title(xlab="speed", line=4, cex.lab=1.2, family="Calibri Light")
+title(main = "Proportion of birds seen by speed", cex.main=0.8, font.main= 3, adj=0, col.main = "gray50",line = c(0.5))
 mtext(text = "speed in knots",
@@
   theme.title+
+  theme(axis.title.y = element_blank())+
   labs(
     title = "Variables influencing effect",
-    subtitle = "Effect by number of birds struck"
+    subtitle = "Proportion of effect by number of birds struck"
   )

### 2021-07-21 draft_final.Rmd
Updated PDF and minor corrections

@@
 
-In this paper we want to investigate a dataset of aircraft-wildlife collisions in the U.S. in the 90s. We start by introducing the data and some steps we took regarding data preparation.
+In this paper we want to investigate a dataset of aircraft-wildlife collisions in the U.S. from the 90s. We start by introducing the data and some steps we took regarding data preparation.
 For our analysis we then first take a look at what kind of aircrafts are frequently involved in bird/wildlife strikes.
@@
 #### Data Preparation: 
-Cleaning *birds_seen* variable: The 9412th observation had one wrong entry in the *birds_seen* variable which has been replaced with *NA*, because we did not want to introduce a new level for one entry.
-
-Factor Levels: The factor levels of the following variables have been ordered for visualization purposes.
+Cleaning *birds_seen* variable: The 9412th observation had one wrong entry in the *birds_seen* variable which has been replaced with a *NA* value, because we did not want to introduce a new level for one entry.
+
+Factor Levels: The factor levels of the following variables have been (re-)ordered for visualization purposes.
 
@@
 By doing this we were able to decrease the missing values by 19 (from 39 to 20).
+\
 *birds_seen*: We assume that missing values in the *birds_seen* variable indicate that the pilot did not see wildlife before the strike occurred.
@@
 
-The following barplot shows the amount of missing values for each of the variables, that had missing values.
-
-```{r echo=FALSE, message=FALSE, fig.asp=0.4}
+The following barplot shows the amount of missing values for each of the variables that had missing values.
+
+```{r echo=FALSE, message=FALSE, fig.asp=0.35}
 #get rows where birds struck is missing
@@
 Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot.
-So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.
+So we could conclude from that, that this is the maximum height and speed possible for aircrafts with one engine.
 We can also see that lighter machines tend to have less engines:
@@
 So **most often we do not see birds before a strike** happens and most often only 1 bird is struck.
-But if the pilot sees something, than according to the data he only sees birds in groups, where larger groups of birds are less likely than smaller ones.
+But if the pilot sees something, then according to the data he only sees birds in groups, where larger groups of birds are less likely than smaller ones.
 If the pilot sees birds, **quite often less birds were hit than seen**.
@@
 The top right plot shows that collisions mostly occur at the beginning and end of a flight in phases Take-Off, Climb, Approach and Landing Roll.
-The amount of collision in Approach phase is highest and about 2 times higher than the other start and end phases.
+The number of collisions in the Approach phase is the highest and about 2 times higher than in the other start and end phases.
 During the middle of the flight in phases "En Route" and "Descent" less collisions occur.
@@
 The height distribution for "Climb" and "Approach" are very similar and their medians are very close to the ground. 
-The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data.
+The other two boxplots for "En Route" and "Descent" are more far away from the ground and show more variance in the data.
 Thus, we will focus on these two.
@@
 This will be most likely because of the angle of the aircraft at which it is flying.
-In phase "En Route" it will be most likely parallel to the ground, so the height of the airplane does not change much during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for "Approach" phase.
+In phase "En Route" it will be most likely parallel to the ground, so the height of the airplane does not change much during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare, for example, for "Approach" phase.
 
@@
 So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**.
-And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 where we found out, that the "Approach" phase has the highest amount of strikes.
+And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 where we found out that the "Approach" phase has the highest number of strikes.
 
@@
 This is visualized in the maps below:
-The left column shows the bird collisions, thr right one the non-bird ones.
-The upper maps show the ratio, so the proportion of bird/non-bird collisions and total collisions by state.
-The two bottom maps show the absolute values, so the number of collisions by state, for bird and non-bird encounters.
+The left column shows the bird strikes, the right one the non-bird ones.
+The upper maps show the ratio, so the proportion of bird/non-bird collisions in relation to total collisions by state.
+The two bottom maps show the absolute values for bird and non-bird encounters.
 
@@
 
-We see that, all in all, the states that have the most bird strucks also have the most non-bird strucks.
-Though states in the midwest and that do not lie on the coastline have a higher ratio of non-bird collisions than those that do.
+We see that, all in all, the states that have the most bird strikes also have the most non-bird strikes.
+Though states in the midwest and ones that do not lie on the coastline have a higher ratio of non-bird collisions than those that do.
 And, what was already quite obvious in the beginning: There are a lot more aircraft-wildlife collisions with birds than other animals (Over 90%).

### 2021-11-17 final_paper.Rmd
Rename files and add Readme

--- 2021-07-21 draft_final.Rmd
@@
         axis.title.x=element_blank(),
-        plot.subtitle = element_text(hjust =4)) +
+        plot.subtitle = element_text(hjust =2)) +
   labs(xlab = "phase of flight",
@@
         axis.title.x=element_blank(),
-        plot.subtitle = element_text(hjust =3)) +
+        plot.subtitle = element_text(hjust =2)) +
   labs(xlab = "",
